---
title: "Hypothesis Testing in R"
subtitle: "A Comprehensive Lecture — From Theory to Practice"
author: "Ange Maxime TCHOUTANG"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    theme: flatly
    highlight: tango
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  comment = "##"
)
```

```{r packages}
library(tidyverse)
```

---

# Introduction to Inferential Statistics

**Inferential statistics** allows us to draw conclusions about a population based on data collected from a sample. Rather than measuring every individual in a population (which is usually impossible), we study a representative sample and use statistical methods to generalize our findings.

**Hypothesis testing** is the core tool of inferential statistics. It provides a rigorous, reproducible framework for making decisions from data — whether you are comparing drug treatments, studying disease risk factors, or analyzing gene expression levels.

> ⚠️ **Key principle:** Statistical tests never *prove* anything. They assess how *compatible* our data are with a given assumption. We either **reject** or **fail to reject** a hypothesis — we never "accept" it.

---

# The Logic of Hypothesis Testing

## The Two Hypotheses

Every hypothesis test involves two competing statements:

- **Null Hypothesis (H₀):** The "default" assumption — usually that there is **no effect, no difference, or no association**. It is what we try to disprove.

- **Alternative Hypothesis (H₁ or Hₐ):** What we wish to demonstrate — that there **is** an effect, a difference, or an association.

**Examples:**

| Research Question | H₀ | H₁ |
|---|---|---|
| Does a new drug reduce blood pressure? | Mean BP before = Mean BP after | Mean BP after < Mean BP before |
| Is gender associated with survival? | Gender and survival are independent | Gender and survival are associated |
| Do three drug doses have the same effect? | All group means are equal | At least one mean differs |

## The P-Value

The **p-value** is the probability of obtaining results at least as extreme as those observed, *assuming H₀ is true*.

- A **small p-value** (typically < 0.05) means the observed data are unlikely under H₀ — we reject H₀.
- A **large p-value** (≥ 0.05) means the data are compatible with H₀ — we fail to reject H₀.

> ⚠️ **Common misconceptions about the p-value:**
>
> ❌ "p = 0.03 means there is a 3% probability that H₀ is true." → **FALSE**
>
> ❌ "p > 0.05 proves H₀ is true." → **FALSE** (absence of evidence ≠ evidence of absence)
>
> ❌ "p < 0.05 means the effect is clinically important." → **FALSE** (significance ≠ importance)
>
> ✅ "p = 0.03 means: if H₀ were true, there would be a 3% chance of observing results this extreme or more extreme."

## Type I and Type II Errors

| Decision \ Reality | H₀ is TRUE | H₀ is FALSE |
|---|---|---|
| **Fail to reject H₀** | ✅ Correct (1 − α) | ❌ Type II error (β) |
| **Reject H₀** | ❌ Type I error (α) | ✅ Correct — **Power** (1 − β) |

- **Type I error (α):** Rejecting H₀ when it is actually true (false positive). Conventionally set at α = 0.05.
- **Type II error (β):** Failing to reject H₀ when it is actually false (false negative).
- **Statistical power (1 − β):** The probability of correctly detecting a true effect. A power of ≥ 80% is the standard target.

## The 7-Step Process of Hypothesis Testing

1. **State the hypotheses** (H₀ and H₁)
2. **Set the significance level** (usually α = 0.05)
3. **Choose the appropriate test** based on data type and design
4. **Verify assumptions** (normality, equal variance, independence)
5. **Compute the test statistic and p-value**
6. **Make a decision** (reject H₀ if p ≤ α)
7. **Draw a conclusion** in the context of your research question

---

# Choosing the Right Test

The choice of test depends on the **type of your variables** and the **number of groups** you are comparing. The table below summarizes the most common situations:

| Question | Normal data? | Test | Non-parametric alternative |
|---|---|---|---|
| Compare sample mean to a reference value | Yes | One-sample t-test | Wilcoxon signed-rank (1 sample) |
| Compare 2 independent groups | Yes | Student / Welch t-test | Mann-Whitney U |
| Compare 2 paired/matched measurements | Yes | Paired t-test | Wilcoxon signed-rank (paired) |
| Compare ≥ 3 independent groups | Yes | One-way ANOVA | Kruskal-Wallis |
| Compare ≥ 3 repeated measures | Yes | Repeated-measures ANOVA | Friedman |
| Association between 2 categorical variables | — | Chi-squared | Fisher's exact (small n) |
| Association between 2 continuous variables | Yes | Pearson correlation | Spearman / Kendall |

> 💡 **Parametric tests** assume a specific distribution (usually normal). **Non-parametric tests** make no such assumption and are safer when that assumption is violated.

---

# Checking Assumptions Before Testing

Before applying any parametric test, you must verify its assumptions.

## Normality — Shapiro-Wilk Test

The Shapiro-Wilk test checks whether a sample comes from a normal distribution.

- **H₀:** The data follow a normal distribution
- **H₁:** The data do not follow a normal distribution

```{r shapiro}
set.seed(42)
# Systolic blood pressure in 20 patients
PAS <- c(120, 130, 125, 118, 135, 122, 128, 132, 119, 126,
         124, 131, 127, 121, 134, 123, 129, 133, 117, 130)

sw <- shapiro.test(PAS)
print(sw)

# Automatic interpretation
if (sw$p.value > 0.05) {
  cat("Normality not rejected (p =", round(sw$p.value, 4), ")\n")
  cat("=> Parametric tests are applicable.\n")
} else {
  cat("Normality rejected (p =", round(sw$p.value, 4), ")\n")
  cat("=> Consider a non-parametric test.\n")
}
```

### Visualizing normality: histogram and QQ-plot

```{r normality-plots, fig.width=9, fig.height=4}
par(mfrow = c(1, 2))

hist(PAS, probability = TRUE, col = "lightblue",
     main = "Histogram + Normal Curve",
     xlab = "Systolic BP (mmHg)", border = "white")
curve(dnorm(x, mean = mean(PAS), sd = sd(PAS)),
      add = TRUE, col = "red", lwd = 2)

qqnorm(PAS, main = "QQ-Plot", pch = 19, col = "steelblue")
qqline(PAS, col = "red", lwd = 2)

par(mfrow = c(1, 1))
```

> 💡 **For large samples (n ≥ 2000)**, Shapiro-Wilk loses relevance. Use the **Kolmogorov-Smirnov** or **Anderson-Darling** test instead (from the `nortest` package).

## Equality of Variances

When comparing two or more groups, parametric tests often assume equal variances (homoscedasticity).

```{r variance-tests}
set.seed(10)
groupA <- rnorm(30, mean = 50, sd = 8)
groupB <- rnorm(30, mean = 55, sd = 9)
groupC <- rnorm(30, mean = 48, sd = 7)

dat_var <- data.frame(
  value = c(groupA, groupB, groupC),
  group = rep(c("A", "B", "C"), each = 30)
)

# Bartlett test (assumes normality within each group)
bartlett.test(value ~ group, data = dat_var)

# Levene test (more robust to departures from normality)
car::leveneTest(value ~ group, data = dat_var)
```

- **p > 0.05** → variances are homogeneous → classical ANOVA applicable
- **p ≤ 0.05** → heterogeneous variances → use Welch's t-test or Welch's ANOVA (`oneway.test()`)

---

# The Student's t-Test

The t-test compares means. It uses the **t-statistic**:

$$t = \frac{\bar{x} - \mu_0}{s / \sqrt{n}}$$

There are three variants depending on your design.

## One-Sample t-Test

Compares the mean of a sample to a known reference value.

**Example:** Are the blood lead levels of children in an industrial area higher than the WHO threshold of 5 µg/dL?

```{r one-sample-ttest}
set.seed(7)
blood_lead <- c(7.2, 6.8, 8.1, 5.9, 9.3, 7.5, 6.4, 8.7, 7.1, 6.9,
                8.3, 7.8, 6.2, 9.1, 7.4, 8.0, 6.7, 7.6, 8.5, 7.3,
                6.1, 8.9, 7.0, 6.6, 8.2)

# One-sided test: is the mean greater than 5?
t.test(blood_lead, mu = 5, alternative = "greater")
```

The very small p-value leads us to **reject H₀** — blood lead levels significantly exceed the WHO threshold.

## Independent (Unpaired) t-Test

Compares the means of **two separate, unrelated groups**.

**Example:** Do male and female Star Wars characters differ in height?

```{r independent-ttest}
my_sw <- starwars %>%
  select(sex, height) %>%
  filter(sex %in% c("male", "female")) %>%
  drop_na(height)

# Visualize first
my_sw %>%
  ggplot(aes(x = height, fill = sex)) +
  geom_density(alpha = 0.5) +
  theme_minimal() +
  labs(title = "Height Distribution by Sex — Star Wars Characters",
       x = "Height (cm)", fill = "Sex")
```

```{r independent-ttest-test}
t.test(height ~ sex, data = my_sw)
```

**Reading the output:**

- **t-statistic:** How many standard errors separate the two means
- **df:** Degrees of freedom (Welch adjustment for unequal variances)
- **p-value:** Decision criterion
- **95% CI:** Plausible range for the true difference in means
- **Sample estimates:** The observed means for each group

**Example from biology (penguins):**

```{r penguins-ttest}
library(palmerpenguins)

penguins %>%
  filter(species %in% c("Adelie", "Chinstrap")) %>%
  drop_na(bill_length_mm) %>%
  t.test(bill_length_mm ~ species, data = .)
```

## Paired t-Test

Used when each observation in one group is **matched or linked** to an observation in the other — e.g., the same patient measured before and after a treatment.

**Example:** Do two sleep drugs produce different amounts of extra sleep in the same patients?

```{r paired-ttest-viz}
# The built-in `sleep` dataset: 10 patients, 2 drugs each
sleep %>%
  mutate(group = paste("Drug", group)) %>%
  group_by(group) %>%
  mutate(mean_extra = mean(extra)) %>%
  ungroup() %>%
  ggplot(aes(x = extra, fill = group)) +
  geom_density(alpha = 0.5) +
  geom_vline(aes(xintercept = mean_extra, color = group),
             linewidth = 1, linetype = "dashed") +
  theme_minimal() +
  labs(title = "Extra Sleep by Drug — Same 10 Patients",
       x = "Extra sleep (hours)", fill = "Drug", color = "Drug")
```

```{r paired-ttest-test}
sleep_wide <- sleep %>%
  pivot_wider(names_from = group, values_from = extra, names_prefix = "drug_")

t.test(sleep_wide$drug_1, sleep_wide$drug_2, paired = TRUE)
```

> ⚠️ **Independent vs. Paired — choose correctly!**
>
> - Different subjects in each group → **Independent t-test**
> - Same subjects measured twice, or matched pairs → **Paired t-test**
>
> Using the wrong test can lead to incorrect conclusions or reduced statistical power.

## Key Arguments of `t.test()`

| Argument | Meaning | Default |
|---|---|---|
| `x`, `y` | Numeric vectors (or formula `y ~ group`) | — |
| `alternative` | `"two.sided"`, `"less"`, or `"greater"` | `"two.sided"` |
| `mu` | Reference value under H₀ | `0` |
| `paired` | Is the test paired? | `FALSE` |
| `var.equal` | Assume equal variances? (`FALSE` = Welch) | `FALSE` |
| `conf.level` | Confidence level for the CI | `0.95` |

## Extracting Results Programmatically

```{r extract-ttest}
result <- t.test(height ~ sex, data = my_sw)

result$p.value      # p-value
result$statistic    # t-statistic
result$conf.int     # confidence interval
result$estimate     # group means
```

---

# The Non-Parametric Alternative: Wilcoxon Test

When normality is violated, use `wilcox.test()` — it makes no distributional assumption and compares **medians** (or rank-based differences).

```{r wilcoxon}
# One-sample: is the median different from a reference?
wilcox.test(blood_lead, mu = 5, alternative = "greater")

# Two-sample (independent):
AH <- c(7, 9, 8, 9, 10, 7, 8, 9, 8, 9)   # group 1
AF <- c(5, 6, 7, 5, 6, 7, 6, 5, 7, 6)   # group 2
wilcox.test(AH, AF)

# Paired version:
wilcox.test(sleep_wide$drug_1, sleep_wide$drug_2, paired = TRUE)
```

---

# The Chi-Squared Test

The chi-squared test (`chisq.test()`) assesses whether two **categorical variables** are independent. It compares **observed frequencies** to what we would **expect** if the variables were unrelated.

- **H₀:** The two variables are independent (no association)
- **H₁:** The two variables are not independent (there is an association)

## Chi-Squared Test of Independence

**Example:** Was survival on the Titanic associated with passenger class?

```{r chisq-titanic}
titanic_df <- Titanic %>%
  as_tibble() %>%
  uncount(n) %>%
  filter(Class != "Crew")

# Contingency table
tab <- table(titanic_df$Class, titanic_df$Survived)
tab
```

```{r chisq-titanic-test}
chisq.test(tab)
```

**Interpreting the output:**

- **X-squared:** The test statistic. Larger = more deviation from independence.
- **df:** (rows − 1) × (columns − 1)
- **p-value:** If < 0.05, reject H₀

The extremely small p-value confirms a **highly significant association** between class and survival.

## Chi-Squared Goodness-of-Fit

When you have a **single categorical variable** and want to test whether its proportions match a theoretical distribution:

```{r chisq-gof}
library(forcats)

my_marital <- gss_cat %>%
  select(marital) %>%
  filter(marital %in% c("Married", "Never married", "Divorced")) %>%
  mutate(marital = fct_drop(marital))

chisq.test(table(my_marital))
```

A tiny p-value: the three marital statuses are **not equally distributed** in the population.

## Fisher's Exact Test

For **small samples** where expected cell counts are < 5, use Fisher's exact test:

```{r fisher}
# Small contingency table
cancer_types <- c("Lung", "Breast", "Prostate", "Lung", "Breast")
response     <- c("Yes", "No", "Yes", "No", "Yes")

fisher.test(table(cancer_types, response))
```

> 💡 **Rule of thumb:** Use `chisq.test()` when all expected frequencies ≥ 5. Use `fisher.test()` otherwise.

---

# ANOVA — Analysis of Variance

ANOVA tests whether the **means of three or more groups** are significantly different. Despite the name, it works by comparing **variances**: the variance *between* groups vs. the variance *within* groups.

- **H₀:** All group means are equal (μ₁ = μ₂ = ... = μₖ)
- **H₁:** At least one group mean is different

The test statistic is the **F-ratio:**

$$F = \frac{\text{Variance between groups}}{\text{Variance within groups}}$$

A large F-ratio means the groups differ more than expected by chance → reject H₀.

## One-Way ANOVA

One categorical independent variable, one continuous dependent variable.

**Example:** Does life expectancy differ across continents in 2007?

```{r anova-gapminder-viz}
library(gapminder)

gapminder_2007 <- gapminder %>% filter(year == 2007)

gapminder_2007 %>%
  ggplot(aes(x = fct_reorder(continent, lifeExp),
             y = lifeExp, fill = continent)) +
  geom_boxplot(alpha = 0.6, width = 0.5) +
  geom_jitter(aes(colour = continent), width = 0.15, alpha = 0.5, size = 2) +
  stat_summary(fun = mean, geom = "point",
               shape = 23, size = 3, fill = "white") +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "Life Expectancy by Continent (2007)",
       subtitle = "Diamond = group mean",
       x = NULL, y = "Life Expectancy (years)")
```

```{r anova-gapminder-test}
mod <- aov(lifeExp ~ continent, data = gapminder_2007)
summary(mod)
```

**Reading the ANOVA table:**

| Column | Meaning |
|---|---|
| **Df** | Degrees of freedom: between groups = k−1; within = n−k |
| **Sum Sq** | Total variability explained by the factor (between) or residual (within) |
| **Mean Sq** | Sum of squares / degrees of freedom |
| **F value** | Ratio of between-group to within-group variance |
| **Pr(>F)** | p-value |

With p < 2e-16, we **reject H₀** — life expectancy differs significantly across continents.

## Post-Hoc Test: Tukey HSD

ANOVA tells you *that* at least one group differs — but not *which* pairs. **Tukey's Honest Significant Differences** (HSD) test performs all pairwise comparisons with a correction for multiple testing.

```{r tukey}
TukeyHSD(mod)
```

```{r tukey-plot, fig.height=6}
plot(TukeyHSD(mod), las = 1)
```

Each row shows the estimated difference and 95% CI for a pair of continents. Intervals that **do not cross zero** indicate a significant difference.

## Two-Way ANOVA

When you have **two categorical independent variables**.

```{r two-way-anova}
set.seed(42)
group <- rep(c("A1", "A2", "A3"), times = 4)
value <- runif(12, 10, 20)
year  <- as.factor(c(rep("2010", 6), rep("2020", 6)))
df2   <- data.frame(value, group, year)

# Interaction model (group * year)
mod2 <- aov(value ~ group * year, data = df2)
summary(mod2)
```

- `group`: main effect of group
- `year`: main effect of year
- `group:year`: interaction — does the effect of group differ depending on year?

## ANOVA with the Pipe

```{r anova-pipe}
msleep %>%
  select(vore, sleep_rem) %>%
  drop_na() %>%
  aov(sleep_rem ~ vore, data = .) %>%
  summary()
```

## Non-Parametric Alternative: Kruskal-Wallis Test

When normality is violated in ≥ 3 groups:

```{r kruskal}
kruskal.test(sleep_rem ~ vore,
             data = msleep %>% drop_na(vore, sleep_rem))
```

---

# Correlation Tests

Correlation measures the **strength and direction** of a linear (or monotonic) relationship between two continuous variables.

> ⚠️ **Correlation implies association, not causation.**

## Pearson Correlation

For two **normally distributed** continuous variables. Measures **linear** association.

- **r = 1:** Perfect positive linear relationship
- **r = 0:** No linear relationship
- **r = −1:** Perfect negative linear relationship

```{r pearson}
cor.test(mtcars$wt, mtcars$mpg, method = "pearson")
```

## Spearman's Rank Correlation

For **ordinal** data or when normality is violated. Measures **monotonic** association.

```{r spearman}
cor.test(mtcars$wt, mtcars$mpg, method = "spearman")
```

## Kendall's Tau

Another rank-based correlation, often preferred for small samples or when there are many ties.

```{r kendall}
cor.test(mtcars$wt, mtcars$mpg, method = "kendall")
```

## Interpreting Correlation Coefficients

| |r| range | Interpretation |
|---|---|
| 0.00 – 0.19 | Negligible |
| 0.20 – 0.39 | Weak |
| 0.40 – 0.59 | Moderate |
| 0.60 – 0.79 | Strong |
| 0.80 – 1.00 | Very strong |

A negative sign means the variables move in opposite directions.

## Extracting Results Programmatically

```{r cor-extract}
res <- cor.test(mtcars$wt, mtcars$mpg, method = "pearson")

res$estimate    # correlation coefficient (r)
res$p.value     # p-value
res$conf.int    # 95% confidence interval for r
```

## Correlation Matrix

To explore associations among multiple variables simultaneously:

```{r cor-matrix}
# Correlation coefficients
cor(iris[, 1:4], method = "pearson")
```

For a visual correlation matrix with significance:

```{r cor-visual, fig.height=5}
iris[, 1:4] %>%
  cor() %>%
  as.data.frame() %>%
  rownames_to_column("Var1") %>%
  pivot_longer(-Var1, names_to = "Var2", values_to = "r") %>%
  ggplot(aes(x = Var1, y = Var2, fill = r)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(r, 2)), size = 4) +
  scale_fill_gradient2(low = "steelblue", mid = "white",
                       high = "tomato", midpoint = 0,
                       limits = c(-1, 1)) +
  theme_minimal() +
  labs(title = "Correlation Matrix — Iris Dataset",
       x = NULL, y = NULL, fill = "r")
```

---

# Normality and Variance Tests — Summary

## Test of Variance Equality — Fisher's F Test

Tests whether two samples have the same variance:

```{r var-test}
var.test(mtcars$mpg, mtcars$wt)
```

- **H₀:** The two variances are equal
- **H₁:** The variances are different

## Kolmogorov-Smirnov Test

Tests whether two samples come from the same distribution:

```{r ks-test}
ks.test(rnorm(100, 5, 1), rnorm(100, 5, 1.5))
```

---

# Complete Statistical Test Reference

## Summary Table

```{r summary-table, echo=FALSE}
tibble::tribble(
  ~Test,                   ~Function,            ~H0,                                    ~When_to_use,
  "One-sample t-test",     "t.test(x, mu=)",     "Mean = reference value",               "1 group, compare to reference",
  "Independent t-test",    "t.test(x, y)",       "Mean(x) = Mean(y)",                    "2 independent groups",
  "Paired t-test",         "t.test(x,y,paired=TRUE)", "Mean difference = 0",            "2 matched/repeated measures",
  "Wilcoxon (1-sample)",   "wilcox.test(x,mu=)", "Median = reference",                   "1 group, non-normal",
  "Mann-Whitney U",        "wilcox.test(x,y)",   "Median(x) = Median(y)",                "2 independent groups, non-normal",
  "Wilcoxon signed-rank",  "wilcox.test(x,y,paired=TRUE)", "Median difference = 0",     "2 paired groups, non-normal",
  "One-way ANOVA",         "aov(y ~ group)",      "All group means equal",               "≥3 groups, parametric",
  "Kruskal-Wallis",        "kruskal.test()",      "All group medians equal",             "≥3 groups, non-parametric",
  "Chi-squared",           "chisq.test()",        "Variables are independent",           "2 categorical variables",
  "Fisher's exact",        "fisher.test()",       "Variables are independent",           "2 categorical, small n",
  "Shapiro-Wilk",          "shapiro.test()",      "Data are normally distributed",       "Test normality assumption",
  "Bartlett",              "bartlett.test()",     "All group variances equal",           "Test homoscedasticity",
  "Levene",                "car::leveneTest()",   "All group variances equal",           "Robust alternative to Bartlett",
  "Fisher F-test",         "var.test()",          "Var(x) = Var(y)",                     "Compare 2 variances",
  "KS test",               "ks.test()",           "Same distribution",                   "Compare 2 distributions",
  "Pearson correlation",   "cor.test(method='pearson')", "No linear association",       "2 normal continuous variables",
  "Spearman correlation",  "cor.test(method='spearman')", "No monotonic association",  "Ordinal or non-normal data",
  "Tukey HSD",             "TukeyHSD(aov_obj)",   "No pairwise difference",             "Post-hoc after ANOVA"
) %>%
  knitr::kable(col.names = c("Test", "R function", "H₀", "When to use"))
```

---

# A Complete Worked Example

## Setting Up a Simulated School Dataset

```{r school-data}
set.seed(7)
s      <- c("Female", "Male")
sexe   <- sample(s, 423, replace = TRUE)
a      <- c(15, 15, 16, 16, 16, 17, 17, 17, 17, 18, 18, 19, 20)
age    <- sample(a, 423, replace = TRUE)
cl     <- c("A", "B", "C")
classe <- sample(cl, 423, replace = TRUE)

anglais  <- round(rnorm(423, 7, 0.5) + ifelse(sexe == "Female", 2, 0), 0)
info     <- round(rnorm(423, 5, 3)   + ifelse(age >= 19, 2, 0), 0)
art      <- round(rnorm(423, 4, 3)   + ifelse(classe == "C", 2, -1), 0)
histoire <- round(rnorm(423, 6, 2)   + ifelse(classe == "B" & sexe == "Male", 2, 0), 0)
moyenne  <- (anglais + info + art + histoire) / 4

ecole <- data.frame(age, sexe, classe, anglais, info, art, histoire, moyenne)
rm(age, classe, sexe, anglais, info, art, histoire)

# Clip scores to [0, 10]
for (v in c("anglais", "info", "art", "histoire")) {
  ecole[[v]] <- pmin(pmax(ecole[[v]], 0), 10)
}
```

## Chi-Squared: Is class associated with student level?

```{r chi2-ecole}
ecole$niveau <- sample(c(1, 2, 3), 423, replace = TRUE)
ecole$niveau <- ordered(ecole$niveau, labels = c("Average", "Good", "Excellent"))

# Contingency table with proportions
round(100 * prop.table(table(ecole$classe, ecole$niveau), 1), 0)

# Chi-squared test
chisq.test(table(ecole$classe, ecole$niveau))
```

**Interpretation:** If p < 0.05, we reject H₀ and conclude that class and student level are **not independent** — there is a significant association.

## t-Test: Do females and males have the same English scores?

```{r ttest-ecole}
t.test(anglais ~ sexe, data = ecole)
```

## ANOVA: Do mean scores differ across classes?

```{r anova-ecole}
mod_ecole <- aov(moyenne ~ classe, data = ecole)
summary(mod_ecole)
TukeyHSD(mod_ecole)
```

## Correlation: Is the history score related to the overall average?

```{r cor-ecole}
cor.test(ecole$histoire, ecole$moyenne, method = "pearson")
```

## Shapiro-Wilk: Is the `info` score normally distributed?

```{r shapiro-ecole}
shapiro.test(ecole$info)
```

If p < 0.05 → data are **not** normally distributed → use Wilcoxon instead of t-test.

---

# Confidence Intervals

A **confidence interval** provides a range of plausible values for the true population parameter. A 95% CI means: if we repeated this study 100 times, 95 of those intervals would contain the true value.

```{r confidence-interval}
# Cholesterol levels of 7 patients
cholesterol <- c(210, 220, 230, 250, 260, 240, 245)

# 95% confidence interval for the mean
t.test(cholesterol)$conf.int

# 99% confidence interval
t.test(cholesterol, conf.level = 0.99)$conf.int
```

> 💡 A wider CI reflects **more uncertainty** (smaller sample, higher confidence level). A narrower CI reflects **more precision**.

---

# Practice Exercises

**Exercise 1.** Using the `ChickWeight` dataset, test whether the average weight at day 21 differs between Diet 1 and Diet 4. First visualize the distributions, check normality, then apply the appropriate test.

**Exercise 2.** Using the `Titanic` dataset, test whether survival is associated with sex. Build the contingency table, visualize it, and run both a chi-squared test and Fisher's exact test.

**Exercise 3.** Using the `msleep` dataset, test whether total sleep differs across diet types (`vore`). Apply a one-way ANOVA, check its assumptions with Shapiro-Wilk and Levene, and run Tukey HSD if the result is significant.

**Exercise 4.** Using `mtcars`, test whether there is a significant correlation between engine displacement (`disp`) and fuel efficiency (`mpg`). Run all three correlation methods (Pearson, Spearman, Kendall) and compare the results.

**Exercise 5.** Using the `sleep` dataset, compare the extra sleep produced by Drug 1 vs. Drug 2 using both an **independent** and a **paired** t-test. Explain why the results differ and which test is more appropriate.

**Exercise 6.** Create a dataset of blood pressure measurements for 30 patients before and after a treatment. Test whether the treatment had a significant effect using the paired t-test. Then check normality and, if violated, repeat with the Wilcoxon signed-rank test.

---

```{r cleanup, include=FALSE}
rm(list = ls())
```

---

*End of Lecture — Hypothesis Testing in R*
