---
title: "Lecture 10: Manipulating Text Data in R with {stringr}"
author: "Ange Maxime TCHOUTANG"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    number_sections: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)
```

---

# Learning Objectives

By the end of this lecture, you will be able to:

- Explain why string manipulation is essential in data science workflows
- Use the core `{stringr}` functions: `str_detect()`, `str_c()`, `str_sub()`, `str_split()`, `str_locate()`, `str_to_upper()`, `str_to_lower()`, and `str_replace()`
- Combine string operations with `{dplyr}` verbs (`mutate()`, `filter()`, `select()`) in a tidyverse pipeline
- Extract first names from full names using `str_locate()` + `str_sub()`
- Present clean results using `{gt}` and `{gtExtras}` publication-ready tables

---

# Why String Manipulation Matters

Real-world data is messy. Names, IDs, labels, addresses, and free-text fields rarely arrive in a clean, analysis-ready format. Text manipulation is therefore one of the most commonly needed data wrangling skills.

> *"80% of data science work is data cleaning — and a large part of cleaning involves text."*

## Common text data problems you will encounter

| Problem | Example | Solution |
|---|---|---|
| Inconsistent capitalisation | `"human"` vs `"Human"` | `str_to_upper()` / `str_to_lower()` |
| Need to detect a pattern | Find all cars with "M" in name | `str_detect()` |
| Need to split a field | Split full name into first + last | `str_split()` / `str_locate()` + `str_sub()` |
| Need to shorten labels | Use first 3 characters of model | `str_sub()` |
| Need to combine fields | Create a description column | `str_c()` |
| Need to replace text | Remove unwanted characters | `str_replace()` / `str_replace_all()` |

---

# Required Packages

```{r load-packages}
library(tidyverse)   # includes stringr, dplyr, tibble
library(gt)          # Publication-quality tables
library(gtExtras)    # Extra themes for gt tables
```

---

# The `{stringr}` Function Family

All `{stringr}` functions follow a consistent naming convention: they begin with `str_` followed by a descriptive verb. This makes them easy to discover and remember.

```{r stringr-overview, echo=FALSE}
tibble::tibble(
  Function         = c("str_detect()", "str_c()", "str_sub()",
                       "str_split()", "str_locate()", "str_to_upper()",
                       "str_to_lower()", "str_replace()", "str_length()",
                       "str_trim()"),
  Purpose          = c("Test if a pattern exists → returns TRUE/FALSE",
                       "Concatenate (join) strings",
                       "Extract a substring by character position",
                       "Split a string into parts",
                       "Find the position of a pattern",
                       "Convert to UPPER CASE",
                       "Convert to lower case",
                       "Replace first match of a pattern",
                       "Count the number of characters",
                       "Remove leading/trailing whitespace"),
  `Returns`        = c("logical", "character", "character",
                       "list", "matrix (start, end)", "character",
                       "character", "character", "integer", "character")
) |>
  gt() |>
  tab_header(title = "Core {stringr} Functions") |>
  cols_align("left") |>
  gt_theme_538()
```

---

# Tutorial 1 — Detecting Patterns with `str_detect()`

## Goal

Filter the `mtcars` dataset to keep only car models whose name contains the letter **"M"**.

## Step-by-step explanation

```{r str-detect-steps}
mtcars |>
  rownames_to_column(var = "model") |>   # Move row names into a column
  mutate(
    has_M = str_detect(model, "M")       # TRUE if "M" appears anywhere in model
  ) |>
  filter(has_M == TRUE) |>               # Keep only rows where has_M is TRUE
  select(model, mpg, cyl, disp) |>       # Select relevant columns
  head(10)
```

> **How `str_detect()` works:**  
> `str_detect(string, pattern)` returns `TRUE` if `pattern` is found inside `string`, and `FALSE` otherwise. The pattern is case-sensitive by default — `"M"` will find "Maserati" but not "mercedes".

## Publication-quality table with {gt}

```{r str-detect-gt}
mtcars |>
  rownames_to_column(var = "model") |>
  mutate(has_M = str_detect(model, "M")) |>
  filter(has_M == TRUE) |>
  select(model, mpg, cyl, disp) |>
  gt() |>
  tab_header(
    title    = "Cars whose name contains the letter M",
    subtitle = "Filtered using str_detect() from {stringr}"
  ) |>
  cols_align(align = "left") |>
  cols_label(model = "Model", mpg = "MPG", cyl = "Cylinders", disp = "Displacement") |>
  gt_theme_538()
```

## Case-insensitive detection

To catch both `"M"` and `"m"`, use a regex with `(?i)`:

```{r str-detect-insensitive}
mtcars |>
  rownames_to_column(var = "model") |>
  mutate(has_m = str_detect(model, "(?i)m")) |>   # (?i) = case-insensitive
  filter(has_m) |>
  select(model, mpg, cyl) |>
  head(8)
```

---

# Tutorial 2 — Splitting Strings with `str_split()`

## Goal

Split Star Wars character names at the **space** to separate first name from the rest.

```{r str-split-demo}
starwars |>
  mutate(
    name_parts = str_split(name, " ")   # Split at every space; returns a list
  ) |>
  select(name, name_parts) |>
  head(8)
```

> `str_split()` returns a **list column** — each element is a character vector of the parts. This is useful when names have a variable number of parts (first name only, first + last, first + middle + last, etc.).

## Extracting just the first element of the split

```{r str-split-first}
starwars |>
  mutate(
    first_name = sapply(str_split(name, " "), `[`, 1)  # Take first element
  ) |>
  select(name, first_name) |>
  head(10)
```

---

# Tutorial 3 — Core String Functions on the `iris` and `mtcars` Datasets

## `str_c()` — Concatenating Strings

`str_c()` pastes text values together, with full control over the separator and any additional text you want to insert.

```{r str-c-demo}
iris |>
  select(Species, Sepal.Length) |>
  mutate(
    Description = str_c(
      Species,       # First piece of text
      ": ",          # Separator (literal text)
      Sepal.Length,  # Second piece of text
      " cm"          # Units appended at the end
    )
  ) |>
  head(10)
```

```{r str-c-gt}
iris |>
  select(Species, Sepal.Length) |>
  mutate(
    Description = str_c(Species, ": ", Sepal.Length, " cm")
  ) |>
  head(15) |>
  gt() |>
  tab_header(
    title    = "Iris data with Description column",
    subtitle = "Created using str_c()"
  ) |>
  cols_align("left") |>
  gt_theme_538()
```

## `str_sub()` — Extracting a Substring by Position

`str_sub(string, start, end)` extracts the characters from position `start` to position `end` (inclusive).

```{r str-sub-demo}
mtcars |>
  rownames_to_column(var = "model") |>
  mutate(
    short_name = str_sub(model, 1, 3)   # Characters 1 to 3
  ) |>
  select(model, short_name, mpg, cyl, disp) |>
  head(10)
```

```{r str-sub-gt}
mtcars |>
  rownames_to_column(var = "model") |>
  mutate(short_name = str_sub(model, 1, 3)) |>
  select(model, short_name, mpg, cyl, disp) |>
  gt() |>
  tab_header(
    title    = "Models and their 3-character abbreviations",
    subtitle = "Created using str_sub(start = 1, end = 3)"
  ) |>
  cols_align("left") |>
  gt_theme_538()
```

> **Tip:** You can use negative indices to extract from the **end** of a string. `str_sub(model, -3, -1)` extracts the last 3 characters.

## `str_to_upper()` — Changing Case

```{r str-to-upper}
starwars |>
  select(name, species) |>
  mutate(
    species_upper = str_to_upper(species),    # All capitals
    species_lower = str_to_lower(species),    # All lowercase
    species_title = str_to_title(species)     # Title Case
  ) |>
  head(10)
```

```{r str-to-upper-gt}
starwars |>
  select(name, species) |>
  mutate(species = str_to_upper(species)) |>
  head(15) |>
  gt() |>
  tab_header(
    title    = "Star Wars characters — Species in UPPER CASE",
    subtitle = "Transformed using str_to_upper()"
  ) |>
  cols_align("left") |>
  gt_theme_538()
```

---

# Tutorial 4 — Locating Patterns with `str_locate()`

## Goal

Extract only the **first name** of each Star Wars Human character by finding the position of the space and extracting everything before it.

## How `str_locate()` works

`str_locate(string, pattern)` returns a two-column **matrix** with `start` and `end` positions of the first match.

```{r str-locate-demo}
# Step 1: See the positions of the first space in each name
starwars |>
  pull(name) |>                        # Extract the name column as a vector
  str_locate(" ") |>                   # Find position of first space
  head(10)
```

> The output shows `start` and `end` columns. For a single-character pattern like `" "` (space), `start` and `end` are always the same position.

## Extracting first names using `str_locate()` + `str_sub()`

```{r str-locate-first-name}
starwars |>
  filter(species == "Human") |>
  mutate(
    first_name = str_sub(
      name,
      start = 1,
      end   = str_locate(name, " ")[, 1] - 1  # One position before the space
    )
  ) |>
  select(name, first_name) |>
  head(15)
```

**How the indexing works:**

- `str_locate(name, " ")` returns a matrix for each row
- `[, 1]` extracts the `start` column (position of the space)
- `- 1` moves one position back — the last character of the first name

```{r str-locate-gt}
starwars |>
  filter(species == "Human") |>
  mutate(
    first_name = str_sub(name, 1, str_locate(name, " ")[, 1] - 1)
  ) |>
  select(name, first_name) |>
  gt() |>
  tab_header(
    title    = "First names of Star Wars Human characters",
    subtitle = "Extracted with str_locate() + str_sub()"
  ) |>
  cols_align("left") |>
  gt_theme_538()
```

---

# Tutorial 5 — Combining Strings with `str_c()` (Scalars and Vectors)

## Scalar combination

```{r str-c-scalar}
first_name <- "John"
last_name  <- "Earl"

full_name <- str_c(first_name, last_name, sep = " ")
cat("Full name:", full_name, "\n")
```

## Vector combination

```{r str-c-vector}
first_names <- c("John", "Keith", "Alice")
last_names  <- c("Earl", "Martin", "Walker")

full_names <- str_c(first_names, last_names, sep = " ")
print(full_names)
```

## Creating a new column with `str_c()` in a pipeline

```{r str-c-pipeline}
iris |>
  select(Species, Sepal.Length) |>
  mutate(
    Description = str_c(Species, " : ", Sepal.Length, " cm")
  ) |>
  gt() |>
  tab_header(
    title    = "Iris data with Description column",
    subtitle = "str_c() combining Species, separator, Sepal.Length, and unit"
  ) |>
  cols_align("center") |>
  gt_theme_538()
```

---

# Tutorial 6 — Additional Useful Functions

## `str_replace()` — Replacing Patterns

```{r str-replace}
starwars |>
  select(name) |>
  mutate(
    name_no_space = str_replace(name, " ", "_"),         # Replace first space
    name_all_dash = str_replace_all(name, " ", "-")      # Replace ALL spaces
  ) |>
  head(10)
```

## `str_length()` — Counting Characters

```{r str-length}
starwars |>
  select(name) |>
  mutate(name_length = str_length(name)) |>
  arrange(desc(name_length)) |>                          # Longest names first
  head(10)
```

## `str_trim()` — Removing Whitespace

```{r str-trim}
messy_names <- c("  Luke Skywalker  ", "Darth Vader ", " Leia Organa")
clean_names <- str_trim(messy_names)                      # Remove leading/trailing spaces

tibble(messy = messy_names, clean = clean_names)
```

## `str_pad()` — Padding Strings to a Fixed Width

Useful for creating standardised ID codes:

```{r str-pad}
ids <- c(1, 5, 42, 100, 999)
padded_ids <- str_pad(ids, width = 4, side = "left", pad = "0")

tibble(original = ids, padded = padded_ids)
```

---

# Summary

| Function | Syntax | Key use case |
|---|---|---|
| `str_detect()` | `str_detect(string, pattern)` | Filter rows by text pattern |
| `str_c()` | `str_c(..., sep = " ")` | Build description or label columns |
| `str_sub()` | `str_sub(string, start, end)` | Extract characters by position |
| `str_split()` | `str_split(string, pattern)` | Split a string into parts |
| `str_locate()` | `str_locate(string, pattern)` | Find position of a pattern |
| `str_to_upper()` | `str_to_upper(string)` | Standardise capitalisation |
| `str_replace()` | `str_replace(string, pattern, replacement)` | Clean or recode values |
| `str_length()` | `str_length(string)` | Count characters |
| `str_trim()` | `str_trim(string)` | Remove accidental whitespace |
| `str_pad()` | `str_pad(string, width, side, pad)` | Create fixed-width ID codes |

---

# Practice Exercises

1. Using `mtcars`, create a column `brand` that extracts only the **first word** of each model name (e.g., `"Mazda"` from `"Mazda RX4"`). Use `str_split()` or `str_locate()` + `str_sub()`.

2. In the `starwars` dataset, use `str_detect()` to find all characters whose name contains the string `"Darth"`. How many are there?

3. Using `iris`, create a new column `label` that looks like: `"setosa_5.1cm"` (species + underscore + Sepal.Length + "cm"). Use `str_c()` and `str_to_lower()`.

4. From the `starwars` dataset, extract the **last name** of all Human characters (the part of the name after the last space). Use `str_locate()` and `str_sub()`.

5. *Challenge:* Use `str_replace_all()` to remove all vowels (a, e, i, o, u — case-insensitive) from car model names in `mtcars`. Which model name becomes empty?

---

# References

- Wickham, H. (2022). *{stringr}: Simple, Consistent Wrappers for Common String Operations*. https://stringr.tidyverse.org
- Wickham, H. & Grolemund, G. (2017). *R for Data Science*, Chapter 14: Strings. https://r4ds.had.co.nz/strings.html
- Regular expressions reference: https://regexr.com
