---
title: "The Pipe Function in R"
subtitle: "A Comprehensive Lecture — from `%>%` to `|>` and the full magrittr family"
author: "Ange Maxime TCHOUTANG"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    theme: flatly
    highlight: tango
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  comment = "##",
  message = FALSE,
  warning = FALSE
)
```

---

# What is the Pipe?

**One sentence summary:** *Take what is on the left, and pass it to the right.*

The pipe operator is one of the most transformative ideas in modern R programming. It lets you chain a sequence of operations together in a single, readable expression — left to right, step by step — instead of nesting function calls inside each other or creating a string of throwaway intermediate variables.

There are two pipe operators you will encounter in R:

| Operator | Origin | Available since |
|---|---|---|
| `%>%` | `magrittr` package | 2014 |
| `\|>` | Base R (native) | R 4.1.0 (2021) |

Both do roughly the same thing. We start with `%>%` because it is more powerful and still the most widely used in the tidyverse ecosystem.

---

# A Brief History

The `%>%` operator was created by **Stefan Milton Bache** and popularized by **Hadley Wickham** through the `magrittr` package in 2014. It was inspired by:

- The **Unix pipe** `|` (which sends the output of one command to the next)
- **F#** language pipelines
- The desire to make R code read more like prose — a sequence of instructions rather than a tangle of nested brackets

The operator became so central to R workflows that in 2021, R 4.1.0 introduced a **native pipe `|>`** built directly into the language, requiring no package at all.

---

# The Problem the Pipe Solves

Imagine you want to perform four operations on a vector in sequence: take the logarithm, take the absolute value, take the square root, then compute the mean.

## Approach 1 — Nested Function Calls

```{r nested-calls}
# Hard to read — must parse from inside out
mean(sqrt(abs(log(1:10))))
```

This works, but you have to read it **inside-out**: `log` first, then `abs`, then `sqrt`, then `mean`. This is the opposite of how we think about a sequence of steps.

## Approach 2 — Intermediate Variables

```{r intermediate-vars}
a1 <- 1:10
a2 <- log(a1)
a3 <- abs(a2)
a4 <- sqrt(a3)
a5 <- mean(a4)
a5
```

This is readable, but now we have five variables (`a1` through `a5`) that we will never use again. They clutter the environment and require us to invent names for things we don't care about.

## Approach 3 — The Pipe ✅

```{r pipe-solution}
library(magrittr)

1:10 %>%
  log() %>%
  abs() %>%
  sqrt() %>%
  mean()
```

The pipe reads exactly like a recipe: *start with 1 to 10, then take the log, then the absolute value, then the square root, then the mean*. No nesting, no temporary variables.

> 💡 **Core rule:** `x %>% f()` is exactly equivalent to `f(x)`. The left-hand side is automatically inserted as the **first argument** of the function on the right.

---

# How the Pipe Works — Step by Step

```{r pipe-step-by-step}
# These four expressions are all equivalent:

sum(1:10)           # classic nested call
1:10 %>% sum()      # pipe version

# Multi-step chain — each result flows into the next
1:10 %>%
  log()  %>%    # log(1:10)
  abs()  %>%    # abs(result above)
  sqrt() %>%    # sqrt(result above)
  mean()        # mean(result above)
```

```{r pipe-mental-model}
# Think of each line as: "take what I just computed, and pass it here"
# Step 1: 1:10         → log()   → NaN-free log values
# Step 2: those values → abs()   → absolute values
# Step 3: those values → sqrt()  → square roots
# Step 4: those values → mean()  → single number
```

---

# Keyboard Shortcut

You will use the pipe constantly, so memorize the shortcut:

| OS | Shortcut |
|---|---|
| **Windows / Linux** | `Ctrl + Shift + M` |
| **Mac** | `Cmd + Shift + M` |

> 💡 In RStudio, you can configure which pipe operator this shortcut inserts: **Tools → Global Options → Code → Use native pipe operator `|>`**. By default it inserts `%>%`.

---

# Assigning a Pipe Result

A pipeline is an expression like any other — you can assign its result to a variable:

```{r pipe-assign}
a <- 1:10

# Assign the result of the pipeline to b
b <- a %>%
  log()  %>%
  abs()  %>%
  sqrt() %>%
  mean(na.rm = TRUE)

a   # a is unchanged — the pipe never modifies the original object
b   # b holds the final computed value
```

You can also use the **right-assignment arrow** `->` at the end of a pipeline (though less common):

```{r right-assign}
a %>%
  log()  %>%
  abs()  %>%
  sqrt() %>%
  mean(na.rm = TRUE) -> b

b
```

> 💡 The original object `a` is **never modified** by the pipe. The pipe always produces a new value; it does not change its inputs in place.

---

# The Pipe with dplyr — Data Analysis

The pipe truly shines when combined with `dplyr` for data manipulation. Each step modifies the data in a clean, readable sequence:

```{r pipe-dplyr}
library(dplyr)

mtcars %>%
  filter(cyl == 6) %>%           # Step 1: Keep only 6-cylinder cars
  group_by(am) %>%               # Step 2: Group by transmission type
  summarize(avg_mpg = mean(mpg), # Step 3: Compute mean mpg per group
            n = n()) %>%
  arrange(desc(avg_mpg))         # Step 4: Sort from highest to lowest
```

Without the pipe, the equivalent would be either deeply nested or full of temporary variables — both harder to write and maintain.

## A Complete Data Pipeline

```{r complete-pipeline}
library(dplyr)
library(tidyr)

# Full pipeline: filter → compute → reshape → display
mtcars %>%
  filter(mpg > 15) %>%
  mutate(
    efficiency = ifelse(mpg > 25, "High", "Medium"),
    hp_per_cyl = hp / cyl
  ) %>%
  group_by(cyl, efficiency) %>%
  summarise(
    mean_mpg     = round(mean(mpg), 1),
    mean_hp_cyl  = round(mean(hp_per_cyl), 1),
    count        = n(),
    .groups      = "drop"
  ) %>%
  arrange(cyl, desc(mean_mpg))
```

---

# Adding Extra Arguments to Piped Functions

The pipe passes the left-hand side as the **first argument**. You can freely add additional arguments to the receiving function:

```{r extra-args}
c(1, 2, NA, 4, 5) %>%
  mean(na.rm = TRUE)    # na.rm is an extra argument, not affected by the pipe

c(3, 1, 4, 1, 5, 9) %>%
  sort(decreasing = TRUE)

c("banana", "apple", "cherry") %>%
  paste(collapse = ", ")
```

---

# The `.` Placeholder — Piping to a Non-First Argument

By default, the pipe places the left-hand side as the **first** argument. When you need it in a different position, use **`.`** as a placeholder:

```{r dot-placeholder}
# lm() expects data= as its second argument — use . to place data there
iris %>%
  lm(Sepal.Length ~ Sepal.Width, data = .)

# Using . in a calculation
iris %>%
  with(Sepal.Length / Sepal.Width) %>%  # compute ratio
  mean(., trim = 0.1)                   # . = the ratio vector above
```

```{r dot-indexing}
# You can even index into . with [[ ]] or []
mtcars %>%
  summary() %>%
  .[, "mpg"]    # extract the mpg column from the summary matrix
```

> ⚠️ The `.` placeholder is specific to `%>%` from `magrittr`. The native `|>` uses `_` instead (R 4.2+), but with more restrictions — more on this below.

---

# Using `%>%` with Curly Braces `{}`

Wrapping an expression in `{}` gives you a mini-function body where you can write any R code, and `.` always refers to the piped value:

```{r curly-braces}
# Apply an arithmetic expression that isn't a named function
5 %>% { . + 3 }

10 %>%
  { . * 2 } %>%
  { . - 5 } %>%
  { . / 3 }
```

This is also how you use arithmetic operators directly with the pipe:

```{r arithmetic-pipe}
# Using backtick-quoted operators (infix form)
5 %>% `+`(3)   # same as 5 + 3

10 %>%
  `+`(2)  %>%
  `-`(5)  %>%
  `/`(3)
```

---

# `%>%` with Conditional Logic

You can embed `if/else` logic inside a pipeline using `{}`:

```{r pipe-conditional}
mtcars %>%
  { if (nrow(.) > 20) filter(., mpg > 20) else . } %>%
  summarise(mean_mpg = mean(mpg),
            n        = n())
```

Here `.` inside the `{}` refers to the data frame passed by the pipe. The entire `{}` block is treated as a single expression whose result flows forward.

---

# `%>%` with Anonymous Functions

In `magrittr`, you can create a **reusable partial pipeline** by starting a chain with `.`:

```{r anon-function-pipe}
# Define a reusable pipeline function
multiply_by_pi <- . %>% { . * pi }

# Call it like a normal function
multiply_by_pi(5)
multiply_by_pi(10)
multiply_by_pi(c(1, 2, 3))
```

This is equivalent to writing `function(x) x * pi` — the `.` at the start signals "this is a function, not a computation".

---

# `%>%` with `stringr` and `ggplot2`

## String Manipulation

```{r pipe-stringr}
library(stringr)

" Hello, World! " %>%
  str_to_lower()          %>%    # lowercase
  str_replace_all("o", "0") %>%  # replace o with 0
  str_trim()                     # strip leading/trailing spaces
```

## Data Visualization with ggplot2

> 💡 Notice that inside a `ggplot2` chain you use `+` to add layers, **not** `%>%`. The pipe hands off the data to `ggplot()`, and `+` takes over for the plot layers.

```{r pipe-ggplot}
library(ggplot2)

mtcars %>%
  filter(cyl %in% c(4, 6, 8)) %>%         # pipe step: filter data
  ggplot(aes(x = wt, y = mpg,             # hand off to ggplot
             color = factor(cyl))) +
  geom_point(size = 3, alpha = 0.7) +     # + from here on
  geom_smooth(method = "lm", se = FALSE) +
  labs(title  = "Weight vs Fuel Economy",
       x      = "Weight (1000 lbs)",
       y      = "Miles per Gallon",
       color  = "Cylinders") +
  theme_bw()
```

---

# Nested Pipelines

Pipelines can be nested — one pipeline computing a value that feeds into another:

```{r nested-pipeline}
mtcars %>%
  group_by(cyl) %>%
  summarise(
    mean_mpg = mean(mpg),
    sd_mpg   = sd(mpg),
    count    = n(),
    .groups  = "drop"
  ) %>%
  mutate(
    cv  = round(sd_mpg / mean_mpg, 3),     # coefficient of variation
    pct = round(count / sum(count) * 100, 1) # share of total cars
  ) %>%
  arrange(desc(mean_mpg))
```

---

# The Native Pipe `|>` (R 4.1+)

Since R 4.1.0, you can use `|>` without loading any package. It is built directly into the language:

```{r native-pipe-enable, eval=FALSE}
# Enable in RStudio: Tools → Global Options → Code → Use native pipe operator
```

```{r native-pipe-basic}
# Basic usage — identical to %>% for simple chains
mtcars |>
  subset(cyl == 6) |>
  summary()
```

## `%>%` vs `|>` — Key Differences

| Feature | `%>%` (magrittr) | `\|>` (native R) |
|---|---|---|
| **Package required** | `magrittr` or `tidyverse` | None |
| **Available since** | 2014 | R 4.1.0 (2021) |
| **Placeholder** | `.` (flexible, anywhere) | `_` (R 4.2+, named args only) |
| **Anonymous functions** | `. %>% { expr }` | `(\(x) expr)()` |
| **Curly brace expressions** | `%>% { ... }` | Not directly supported |
| **Speed** | Slightly slower | Slightly faster |
| **Ecosystem support** | Huge (all tidyverse docs) | Growing |

```{r pipe-comparison}
# Doubling every element

# With %>%
1:10 %>% { . * 2 }

# With |> (requires anonymous function syntax)
1:10 |> (\(x) x * 2)()
```

```{r native-pipe-placeholder}
# Native pipe placeholder _ (R 4.2+ only, named arguments only)
mtcars |>
  lm(formula = mpg ~ wt, data = _) |>
  summary()
```

> 💡 **Which one to use?** Use `%>%` if you are working in the tidyverse or need the `.` placeholder and curly-brace tricks. Use `|>` if you want a dependency-free script for simpler chains.

---

# The Full magrittr Family

`magrittr` provides four pipe operators, each for a different purpose:

## `%>%` — The Standard Pipe

The main pipe — passes the left-hand side as the first argument to the right.

```{r magrittr-main}
library(magrittr)

1:5 %>% sum()   # sum(1:5)
```

## `%T>%` — The "Tee" Pipe (Side Effect + Continue)

`%T>%` passes the value to a function **for a side effect** (like printing or plotting), then **continues** with the original value unchanged:

```{r tee-pipe}
# Without %T>%: nrow() would receive the plot's invisible return value
# With %T>%: the original data frame flows through to nrow()
mtcars %>%
  select(hp, mpg) %T>%
  plot() %>%          # side effect: create plot
  nrow()              # continues with the data frame → 32
```

Think of `%T>%` as a "tap" on the pipe — you branch off for a side effect, then return to the main flow.

## `%$%` — The Exposition Pipe

`%$%` **exposes** the columns (or components) of the left-hand side so they can be referred to by name directly:

```{r exposition-pipe}
# Without %$%: need to use $ accessor
cor(mtcars$hp, mtcars$mpg)

# With %$%: column names are directly in scope
mtcars %$%
  cor(hp, mpg)

mtcars %$%
  plot(hp, mpg)
```

This is particularly useful with functions that don't have a `data =` argument.

## `%<>%` — The Assignment Pipe (In-Place Modification)

`%<>%` applies the pipeline **and assigns the result back** to the original variable:

```{r assignment-pipe}
# Classic version — two lines
x <- 1:3
x <- sum(x)
x

# With %<>% — one line
x <- 1:3
x %<>% sum()
x

# More complex in-place transformation
v <- c(3, 1, 4, 1, 5, 9, 2, 6)
v %<>% sort() %<>% rev()   # sort ascending then reverse to descending
v
```

> ⚠️ Use `%<>%` carefully — in-place modification can make code harder to debug, since the original value is lost.

## The Four Pipes — Summary

| Operator | Name | What it does |
|---|---|---|
| `%>%` | Standard pipe | Pass LHS as first arg to RHS |
| `%T>%` | Tee pipe | Side effect on RHS, then pass LHS forward |
| `%$%` | Exposition pipe | Expose names of LHS inside RHS |
| `%<>%` | Assignment pipe | Apply pipeline and save result back to LHS |

---

# Custom Infix Operators

The pipe `%>%` is itself an **infix operator** — it sits between its two arguments. Under the hood it is just a function:

```{r pipe-as-function}
# These are identical
1:10 %>% sum()
`%>%`(1:10, sum())
```

R lets you define your **own infix operators**. The rules are simple:

1. The name must start and end with `%`
2. Define it like a regular two-argument function

```{r custom-infix}
# A custom addition operator
`%plus%` <- function(a, b) a + b

3 %plus% 7
`%plus%`(3, 7)   # called as a regular function

# A "not in" operator (complement of %in%)
`%not_in%` <- function(x, table) !x %in% table

c(1, 2, 3, 4, 5) %not_in% c(2, 4)   # TRUE where element is NOT in the set

# A percent-change operator
`%pct_change%` <- function(new, old) round((new - old) / old * 100, 1)

120 %pct_change% 100   # +20%
85  %pct_change% 100   # -15%
```

---

# Pipe Style Guide

Well-formatted pipes make code easy to read and maintain:

```{r style-good}
# Good: one step per line, consistent indentation
mtcars %>%
  filter(cyl == 4) %>%
  select(mpg, wt, hp) %>%
  mutate(wt_kg = wt * 453.6) %>%
  summarise(mean_mpg = mean(mpg))
```

```{r style-bad, eval=FALSE}
# Avoid: everything on one line (hard to read and debug)
mtcars %>% filter(cyl == 4) %>% select(mpg, wt, hp) %>% mutate(wt_kg = wt * 453.6) %>% summarise(mean_mpg = mean(mpg))

# Avoid: chains longer than ~10 steps — consider breaking into named steps
```

**Recommended conventions:**

- Put `%>%` at the **end** of a line, never at the start of the next
- Indent each step by 2 spaces
- Limit chains to steps that form a logical unit; save intermediate results when needed
- Use blank lines to separate logical groups within a long pipeline
- Assign the result when you will use it again

---

# Debugging Pipelines

When a pipeline produces an unexpected result, isolate the problem by running it **step by step** — add or remove one `%>%` at a time:

```{r debug-pipeline}
# Full pipeline
mtcars %>%
  filter(cyl == 4) %>%
  mutate(wt_kg = wt * 453.6) %>%
  summarise(mean_wt_kg = mean(wt_kg))

# Debug: check what filter produces
mtcars %>%
  filter(cyl == 4)

# Debug: check what mutate produces
mtcars %>%
  filter(cyl == 4) %>%
  mutate(wt_kg = wt * 453.6)
```

You can also use `%T>%` to inspect intermediate values without breaking the chain:

```{r debug-tee}
mtcars %>%
  filter(cyl == 4) %T>%
  { cat("After filter:", nrow(.), "rows\n") } %>%
  mutate(wt_kg = wt * 453.6) %T>%
  { cat("After mutate: columns are", names(.), "\n") } %>%
  summarise(mean_wt_kg = mean(wt_kg))
```

---

# Complete Real-World Examples

## Example 1 — Full Data Analysis Pipeline

```{r example-analysis}
library(dplyr)
library(tidyr)

# Analyse sleep patterns by diet type in mammals
msleep %>%
  drop_na(vore, sleep_total, sleep_rem) %>%
  group_by(vore) %>%
  summarise(
    n           = n(),
    mean_sleep  = round(mean(sleep_total), 1),
    median_rem  = round(median(sleep_rem), 1),
    sd_sleep    = round(sd(sleep_total), 2),
    .groups     = "drop"
  ) %>%
  mutate(rem_ratio = round(median_rem / mean_sleep, 2)) %>%
  arrange(desc(mean_sleep))
```

## Example 2 — String Cleaning Pipeline

```{r example-strings}
library(stringr)

raw_names <- c("  Alice SMITH ", "bob jones", "CARLA  De La Cruz", "  dr. david LEE")

raw_names %>%
  str_trim()          %>%    # remove leading/trailing whitespace
  str_squish()        %>%    # collapse internal spaces
  str_to_title()             # capitalise each word
```

## Example 3 — Data Reshaping Pipeline

```{r example-reshape}
library(tidyr)

# Wide to long, then summarise
mtcars %>%
  select(cyl, mpg, hp, wt) %>%
  pivot_longer(cols      = -cyl,
               names_to  = "metric",
               values_to = "value") %>%
  group_by(cyl, metric) %>%
  summarise(mean_val = round(mean(value), 2),
            .groups  = "drop") %>%
  pivot_wider(names_from  = metric,
              values_from = mean_val)
```

## Example 4 — Visualization Pipeline

```{r example-viz}
library(ggplot2)

msleep %>%
  drop_na(vore) %>%
  mutate(vore = factor(vore,
                       levels = c("carni", "omni", "herbi", "insecti"))) %>%
  ggplot(aes(x = vore, y = sleep_total, fill = vore)) +
  geom_boxplot(alpha = 0.6) +
  geom_jitter(width = 0.15, alpha = 0.4, size = 2) +
  labs(title    = "Total Sleep by Diet Type",
       subtitle = "Each point is one species",
       x = "Diet type", y = "Sleep (hours)",
       fill = "Diet") +
  theme_bw() +
  theme(legend.position = "none")
```

---

# Quick Reference

```r
# Core pipe
x %>% f()                    # f(x)
x %>% f(extra_arg)           # f(x, extra_arg)
x %>% f() %>% g() %>% h()   # h(g(f(x)))

# Placeholder
x %>% f(y, data = .)         # f(y, data = x)
x %>% { . + 1 }              # x + 1

# Assignment
b <- x %>% f()               # compute and store
x %<>% f()                   # modify x in place

# Side effects
x %>% f() %T>% print() %>% g()  # print after f(), continue to g()

# Expose names
df %$% plot(col1, col2)      # use column names directly

# Native pipe (no package needed)
x |> f()                     # same as x %>% f() for simple cases
x |> (\(x) x + 1)()          # anonymous function equivalent of x %>% {. + 1}
```

---

# Practice Exercises

**Exercise 1.** Rewrite the following nested call as a pipe chain:  
`round(mean(sqrt(abs(c(-4, 9, -16, 25, -36)))), digits = 2)`

**Exercise 2.** Using the `starwars` dataset and a single pipeline, filter to humans only, drop rows with missing `height` or `mass`, compute BMI (`mass / (height/100)^2`), and return the top 5 characters by BMI with only their `name`, `height`, `mass`, and `bmi` columns.

**Exercise 3.** Use `%T>%` to build a pipeline on `airquality` that: (1) removes rows with NAs, (2) **prints the number of remaining rows** as a side effect, (3) then computes the mean `Ozone`, `Solar.R`, `Wind`, and `Temp` with `summarise()`.

**Exercise 4.** Use `%<>%` to take the vector `v <- c(5, 3, 8, 1, 9, 2, 7)` and apply the following in-place: sort it, then compute the cumulative sum.

**Exercise 5.** Create a custom infix operator `%between%` that returns `TRUE` if a number falls strictly between two values. For example, `5 %between% c(3, 10)` should return `TRUE` and `15 %between% c(3, 10)` should return `FALSE`.

**Exercise 6.** Build a single pipeline starting from `msleep` that: groups by `order`, computes mean `sleep_total` per order, reorders the `order` factor by mean sleep (using `forcats::fct_reorder()`), and plots a horizontal bar chart with `ggplot2` showing the top 10 sleepiest orders.

---

```{r cleanup, include=FALSE}
rm(list = ls())
```

---

*End of Lecture — The Pipe Function in R*
