---
title: "Data Visualization with ggplot2"
subtitle: "A Comprehensive Lecture — From Grammar to Publication-Ready Plots"
author: "Ange Maxime TCHOUTANG"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    theme: flatly
    highlight: tango
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  comment = "##",
  fig.align = "center",
  fig.width = 7,
  fig.height = 4.5
)
```

```{r load-packages}
library(tidyverse)   # includes ggplot2, dplyr, etc.
```

---

# Introduction to ggplot2

`ggplot2` is one of the most widely used packages in R, and for good reason. It produces beautiful, customizable, and reproducible graphics with a consistent and logical syntax. Once you understand how `ggplot2` works, you will rarely need to reach for Excel, GraphPad, or any other tool for your plots.

`ggplot2` is based on the **Grammar of Graphics** — a systematic theory of how statistical charts are constructed. Instead of memorizing different functions for each plot type, you learn a single coherent grammar that lets you build virtually any visualization by combining a small set of components.

## The Grammar of Graphics

Every `ggplot2` plot is made of the following components:

| Component | Description |
|-----------|-------------|
| **Data** | A `data.frame` containing the variables to visualize |
| **Aesthetics (`aes()`)** | Mappings between variables and visual properties (x, y, color, size, shape, ...) |
| **Geometries (`geom_*()`)** | The visual marks representing the data (points, lines, bars, boxes, ...) |
| **Scales (`scale_*()`)** | Controls how data values map to visual properties |
| **Facets (`facet_*()`)** | Splits the plot into subpanels based on a variable |
| **Themes (`theme*()`)** | Controls all non-data visual elements (fonts, backgrounds, grids, ...) |
| **Coordinate system** | Cartesian (default), polar, flipped, etc. |
| **Statistical transformations (`stat`)** | Summarizes data before plotting (counts, means, bins, ...) |

The basic template for any ggplot2 plot is:

```r
ggplot(data = <DATA>, aes(<MAPPINGS>)) +
  <GEOM_FUNCTION>() +
  <other layers>
```

> ⚠️ **Rule #1:** Your data **must** be in a `data.frame` (or `tibble`). There are no exceptions.

---

# Your First Plot

Let's use the built-in `ChickWeight` dataset to build our first plot step by step.

```{r first-look}
head(ChickWeight)
```

## Step 1 — Initialize the canvas

```{r canvas-only}
ggplot(data = ChickWeight)
```

This creates an empty canvas. Nothing is visible yet because we have not told ggplot2 *what* to draw or *where* to put things.

## Step 2 — Add aesthetic mappings

```{r canvas-with-aes}
ggplot(data = ChickWeight, aes(x = Time, y = weight))
```

Now ggplot2 knows which variables go on each axis, so it draws the axis scales. But still no data is drawn — we need a **geometry**.

## Step 3 — Add a geometry layer

```{r first-scatterplot}
ggplot(data = ChickWeight, aes(x = Time, y = weight)) +
  geom_point()
```

This is our first complete plot — a **scatterplot**. Notice the `+` operator: it adds layers to a ggplot object, just like addition.

## Step 4 — Enrich with more aesthetics

We can map additional variables to other visual properties inside `aes()`:

```{r richer-scatterplot}
ggplot(data = ChickWeight, aes(x = Time, y = weight, color = Diet)) +
  geom_point(size = 3) +
  labs(title = "Chick Weight Gain Over Time by Diet",
       x = "Days since birth",
       y = "Weight (g)")
```

> 💡 **Key distinction:** Aesthetics **inside `aes()`** are *mapped to a variable* (and generate a legend). Aesthetics **outside `aes()`** are *fixed values* that apply to all data points (e.g., `geom_point(color = "blue", size = 3)`).

---

# Core Geometry Types

## Scatterplot — `geom_point()`

Used to visualize the relationship between **two numerical variables**.

```{r scatterplot}
ggplot(data = ChickWeight, aes(x = Time, y = weight, color = Diet)) +
  geom_point(size = 2, alpha = 0.6) +
  labs(title = "Weight vs Time", x = "Time (days)", y = "Weight (g)")
```

### Adding a trend line — `geom_smooth()`

```{r smooth}
ggplot(data = ChickWeight, aes(x = Time, y = weight)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = TRUE, color = "darkred") +
  labs(title = "Weight vs Time with Linear Trend")
```

`method = "lm"` fits a linear regression line. `se = TRUE` adds the confidence interval band. Replace `"lm"` with `"loess"` for a smooth curved fit.

> ⚠️ **Where you place `color` inside `aes()` matters.** Placing it in the top-level `ggplot(aes(...))` call applies it to *all* layers. Placing it inside a specific `geom_()` layer restricts it to only that layer:

```{r aes-placement}
# color applies to points only, smooth line is a single global fit
ggplot(data = ChickWeight, aes(x = Time, y = weight)) +
  geom_point(aes(color = Diet), alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE)
```

## Bar Chart — `geom_bar()` and `geom_col()`

`geom_bar()` is for **counting** — it computes frequencies automatically from a categorical variable.  
`geom_col()` is used when you already have the **pre-computed values** to plot.

```{r barplot}
library(forcats)

msleep %>%
  drop_na(vore) %>%
  ggplot(aes(x = fct_infreq(vore))) +   # fct_infreq orders bars by frequency
  geom_bar(fill = "#97B3C6") +
  coord_flip() +
  theme_bw() +
  labs(title = "Number of Animals by Diet Type",
       x = "Diet type",
       y = NULL)
```

### Grouped bar charts

To compare two categorical variables simultaneously, map one to `x` and the other to `fill`:

```{r grouped-bar}
starwars %>%
  filter(hair_color %in% c("black", "brown")) %>%
  drop_na(sex) %>%
  ggplot(aes(hair_color, fill = sex)) +
  geom_bar(position = "dodge", alpha = 0.8) +
  theme_bw() +
  labs(title = "Hair Colour and Sex in Star Wars",
       x = "Hair Colour", y = "Count")
```

`position = "dodge"` places bars side by side. Use `position = "fill"` for proportional stacked bars, or `position = "stack"` (default) for stacked counts.

## Histogram — `geom_histogram()`

Used to visualize the **distribution of a single numerical variable**.

```{r histogram}
msleep %>%
  ggplot(aes(x = sleep_total)) +
  geom_histogram(binwidth = 2, fill = "#97B3C6", color = "white") +
  theme_bw() +
  labs(title = "Distribution of Total Sleep in Mammals",
       x = "Total sleep (hours)", y = "Count")
```

> ⚠️ Always set `binwidth` explicitly. The default of 30 bins is rarely optimal and ggplot2 will warn you about it.

## Density Plot — `geom_density()`

A smooth version of the histogram, useful for comparing distributions across groups.

```{r density}
msleep %>%
  drop_na(vore) %>%
  filter(vore %in% c("herbi", "omni", "carni")) %>%
  ggplot(aes(x = sleep_total, fill = vore)) +
  geom_density(alpha = 0.4) +
  theme_bw() +
  labs(title = "Sleep Distribution by Diet Type",
       x = "Total sleep (hours)")
```

The `alpha` parameter controls transparency — essential when curves overlap.

## Boxplot — `geom_boxplot()`

Displays the **distribution** of a numerical variable across categories: median, quartiles, and outliers.

```{r boxplot}
ggplot(data = ChickWeight, aes(x = Diet, y = weight)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, color = "orange", alpha = 0.4) +
  theme_bw() +
  labs(title = "Chick Weight Distribution by Diet",
       x = "Diet", y = "Weight (g)")
```

`geom_jitter()` adds the raw data points on top, offset slightly to avoid overlap. This reveals sample size and data spread that the boxplot alone hides.

### Violin plot — `geom_violin()`

A richer alternative to the boxplot, showing the full probability density:

```{r violin}
ggplot(data = iris, aes(x = Species, y = Petal.Width, fill = Species)) +
  geom_violin(alpha = 0.5) +
  geom_jitter(aes(color = Species), width = 0.1, alpha = 0.3) +
  theme_bw() +
  labs(title = "Petal Width by Iris Species")
```

## Line Graph — `geom_line()`

Ideal for **time series** or any ordered sequence.

```{r line-graph}
Orange %>%
  filter(Tree %in% c("1", "3", "5")) %>%
  ggplot(aes(x = age, y = circumference, color = Tree)) +
  geom_point(size = 3, alpha = 0.5) +
  geom_line(linewidth = 1) +
  theme_minimal() +
  labs(title = "Orange Tree Growth Over Time",
       x = "Age (days)", y = "Circumference (mm)")
```

## Heatmap — `geom_tile()`

Visualizes a **third variable** as color across a grid of two categorical (or discrete) variables.

```{r heatmap}
# Build a simple correlation heatmap from the iris dataset
cor_mat <- cor(iris[, 1:4])
cor_df  <- as.data.frame(as.table(cor_mat))
names(cor_df) <- c("Var1", "Var2", "Correlation")

ggplot(cor_df, aes(x = Var1, y = Var2, fill = Correlation)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "steelblue", mid = "white",
                       high = "darkred", midpoint = 0) +
  theme_minimal() +
  labs(title = "Correlation Heatmap — Iris Dataset")
```

## Area Plot — `geom_area()`

Useful to compare magnitudes of overlapping groups across a continuous axis.

```{r area-plot}
ggplot(mpg, aes(x = displ)) +
  geom_area(aes(y = hwy, fill = "Highway"), alpha = 0.6) +
  geom_area(aes(y = cty, fill = "City"),    alpha = 0.6) +
  labs(title = "Highway vs City Fuel Efficiency",
       x = "Engine Displacement (L)",
       y = "Miles per Gallon",
       fill = "") +
  theme_bw()
```

---

# Labels and Annotations

## `labs()` — Titles and axis labels

```{r labs}
ggplot(mpg, aes(x = displ, y = hwy, color = drv)) +
  geom_point(size = 2) +
  labs(title   = "Engine Size vs Highway Fuel Economy",
       subtitle = "Grouped by drive type",
       caption  = "Source: ggplot2::mpg dataset",
       x = "Engine displacement (L)",
       y = "Highway miles per gallon",
       color = "Drive type") +
  theme_bw()
```

## `annotate()` — Adding text and arrows

```{r annotate}
mtcars %>%
  ggplot(aes(x = wt, y = mpg, color = factor(gear))) +
  geom_point(size = 3, alpha = 0.7) +
  annotate("text",
           x = 4, y = 30,
           label = "Notice the difference\nbetween 3 and 4 gear cars",
           color = "black", size = 3.5) +
  geom_segment(x = 4.5, y = 27, xend = 5, yend = 17,
               arrow = arrow(length = unit(0.5, "cm")),
               color = "darkred", linewidth = 1) +
  labs(title = "Weight vs MPG by Number of Gears",
       x = "Weight (1000 lbs)", y = "MPG", color = "Gears") +
  theme_bw()
```

---

# Facets — Small Multiples

Facets split your data into subplots based on the values of one or two categorical variables. This is one of the most powerful features in ggplot2 for exploratory analysis.

## `facet_wrap()` — Wrap panels into rows and columns

```{r facet-wrap}
ggplot(data = ChickWeight, aes(x = Time, y = weight)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ Diet) +
  theme_bw() +
  labs(title = "Weight over Time — One Panel per Diet")
```

Control layout with `nrow` and `ncol`:

```{r facet-wrap-layout}
library(forcats)

gss_cat %>%
  filter(partyid %in% c("Strong democrat", "Strong republican", "Independent")) %>%
  ggplot(aes(age)) +
  geom_histogram(binwidth = 5, fill = "steelblue", alpha = 0.8) +
  facet_wrap(~ partyid, nrow = 1, strip.position = "bottom") +
  labs(title = "Age distribution by political affiliation",
       x = " ", y = " ") +
  theme_bw()
```

## `facet_grid()` — Two-way faceting

`facet_grid(rows ~ cols)` creates a full grid where rows correspond to one variable and columns to another.

```{r facet-grid}
Salaries <- car::Salaries

Salaries %>%
  mutate(discipline = factor(discipline,
                             levels = c("A", "B"),
                             labels = c("Agriculture", "Biology"))) %>%
  ggplot(aes(yrs.since.phd, salary)) +
  geom_point(aes(color = rank), alpha = 0.5) +
  geom_smooth() +
  facet_grid(sex ~ discipline) +
  labs(title   = "Salary vs Years Since PhD",
       x = "Years since PhD",
       y = "Salary (USD)",
       color = "Position") +
  theme_bw()
```

---

# Scales

Scales control the mapping from data values to visual properties. Every aesthetic has a corresponding scale that can be customized.

## Axis scales — `scale_x_*()` and `scale_y_*()`

```{r axis-scale}
ggplot(iris, aes(x = Sepal.Width, y = Sepal.Length)) +
  geom_point() +
  scale_x_continuous(name   = "Sepal Width (cm)",
                     breaks = seq(2, 5, by = 0.5),
                     limits = c(2, 5)) +
  scale_y_continuous(name   = "Sepal Length (cm)") +
  theme_bw()
```

## Renaming tick labels — `scale_x_discrete()`

```{r discrete-scale}
Salaries %>%
  ggplot(aes(x = rank, y = salary, fill = sex)) +
  geom_boxplot(alpha = 0.5) +
  scale_x_discrete(labels = c("AsstProf"  = "Assistant\nProfessor",
                               "AssocProf" = "Associate\nProfessor",
                               "Prof"      = "Full\nProfessor")) +
  labs(title = "Salary by Academic Rank and Gender",
       x = "", y = "Salary (USD)", fill = "Gender") +
  theme_bw()
```

## Color scales — `scale_color_*()` and `scale_fill_*()`

### Discrete (categorical) colors

```{r color-brewer}
mpg %>%
  ggplot(aes(displ, hwy, color = drv)) +
  geom_jitter(size = 3) +
  scale_color_brewer(palette = "Set2") +
  theme_minimal() +
  labs(title = "Fuel Economy by Drive Type (ColorBrewer palette)")
```

### Manual colors

```{r manual-color}
mpg %>%
  ggplot(aes(displ, hwy, color = drv)) +
  geom_jitter(size = 3) +
  scale_color_manual(values = c("4" = "#719AC9",
                                "f" = "#75B99C",
                                "r" = "#C98071")) +
  theme_minimal() +
  labs(title = "Fuel Economy — Manual Hex Colors")
```

> 💡 Find hex color codes at **htmlcolorcodes.com**. Use `display.brewer.all(colorblindFriendly = TRUE)` from the `RColorBrewer` package to see all colorblind-safe palettes.

### Continuous (gradient) colors

```{r gradient-color}
ggplot(iris, aes(x = Sepal.Width, y = Sepal.Length,
                 color = Petal.Length)) +
  geom_point(size = 3) +
  scale_color_gradient(low = "blue", high = "red") +
  theme_bw() +
  labs(title = "Sepal Dimensions Colored by Petal Length")
```

---

# Themes

Themes control all non-data visual elements: text, backgrounds, grid lines, axis lines, legend appearance, etc.

## Built-in themes

```{r themes-showcase, fig.width=9, fig.height=6}
p <- ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  labs(x = "Weight", y = "MPG")

library(patchwork)
(p + ggtitle("theme_bw()")      + theme_bw())      +
(p + ggtitle("theme_classic()") + theme_classic())  +
(p + ggtitle("theme_minimal()") + theme_minimal())  +
(p + ggtitle("theme_dark()")    + theme_dark())
```

## `theme_set()` — Apply a theme globally

Set your preferred theme once at the top of your script:

```{r theme-set, eval=FALSE}
theme_set(theme_bw())   # All subsequent plots use theme_bw()
```

## Customizing with `theme()`

The `theme()` function gives granular control over every visual element:

```{r custom-theme}
ggplot(mtcars, aes(wt, mpg, color = factor(cyl))) +
  geom_point(size = 3) +
  labs(title = "A Customized Plot",
       x = "Weight (1000 lbs)",
       y = "Miles per Gallon",
       color = "Cylinders") +
  theme_bw() +
  theme(
    plot.title    = element_text(size = 16, face = "bold",
                                  color = "steelblue", hjust = 0.5),
    axis.title    = element_text(size = 12, color = "steelblue"),
    legend.position = c(0.85, 0.80),
    panel.grid.minor = element_blank()
  )
```

Key `theme()` arguments:

| Element | What it controls |
|---------|-----------------|
| `plot.title` | The plot title text |
| `axis.title.x/y` | Axis label text |
| `axis.text.x/y` | Axis tick labels |
| `axis.line.x/y` | Axis lines |
| `panel.background` | The plot area background |
| `plot.background` | The entire figure background |
| `panel.grid.major/minor` | Grid lines |
| `legend.position` | Legend location (`"top"`, `"bottom"`, `"none"`, or `c(x, y)`) |
| `strip.text` | Facet panel labels |

---

# Error Bars — `geom_errorbar()`

Displaying uncertainty is critical in scientific visualization.

```{r error-bars}
ToothGrowth %>%
  filter(supp == "VC") %>%
  mutate(dose_f = as.factor(dose)) %>%
  group_by(dose_f) %>%
  summarise(mean_len = mean(len), sd_len = sd(len)) %>%
  ggplot(aes(dose_f, mean_len)) +
  geom_point(size = 4, color = "steelblue") +
  geom_errorbar(aes(ymin = mean_len - sd_len,
                    ymax = mean_len + sd_len),
                width = 0.08) +
  labs(title = "Average Tooth Growth by Dose (Supplement: VC)",
       x = "Dose (mg/day)", y = "Mean tooth length") +
  theme_bw()
```

You can combine error bars with bar charts too:

```{r error-bars-bar}
ToothGrowth %>%
  filter(supp == "VC") %>%
  mutate(dose_f = as.factor(dose)) %>%
  group_by(dose_f) %>%
  summarise(mean_len = mean(len), sd_len = sd(len)) %>%
  ggplot(aes(dose_f, mean_len)) +
  geom_bar(stat = "identity", fill = "skyblue", alpha = 0.7) +
  geom_errorbar(aes(ymin = mean_len - sd_len,
                    ymax = mean_len + sd_len),
                width = 0.08) +
  labs(title = "Average Tooth Growth — Bar Chart with Error Bars",
       x = "Dose (mg/day)", y = "Mean tooth length") +
  theme_bw()
```

---

# Lollipop Chart

A lollipop chart is a clean alternative to bar charts for comparing ranked values. It combines `geom_segment()` and `geom_point()`.

```{r lollipop}
msleep %>%
  group_by(order) %>%
  summarise(mean_sleep = mean(sleep_total)) %>%
  mutate(order = fct_reorder(order, mean_sleep)) %>%
  ggplot(aes(x = order, y = mean_sleep)) +
  geom_hline(yintercept = mean(msleep$sleep_total),
             color = "grey50", linewidth = 0.8) +
  geom_segment(aes(x = order, xend = order,
                   y = mean(msleep$sleep_total),
                   yend = mean_sleep),
               color = "grey70") +
  geom_point(aes(color = mean_sleep), size = 4) +
  scale_color_gradient(low = "steelblue", high = "orange") +
  coord_flip() +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        legend.position  = "none") +
  labs(title = "Average Sleep Time of Mammals by Order",
       x = "", y = "Hours of sleep")
```

---

# Combining Multiple Plots — `patchwork`

The `patchwork` package lets you compose multiple ggplot objects into one figure.

```{r patchwork}
library(patchwork)

p1 <- ggplot(iris, aes(Sepal.Length, Sepal.Width, color = Species)) +
  geom_point() + theme_bw() + labs(title = "Scatter")

p2 <- ggplot(iris, aes(x = Species, y = Sepal.Length, fill = Species)) +
  geom_boxplot() + theme_bw() + labs(title = "Boxplot")

p3 <- ggplot(iris, aes(x = Sepal.Length, fill = Species)) +
  geom_density(alpha = 0.4) + theme_bw() + labs(title = "Density")

# Layout: p1 on top, p2 and p3 side by side on the bottom
p1 / (p2 + p3) +
  plot_annotation(title = "Iris Dataset — Three Views",
                  theme = theme(plot.title = element_text(hjust = 0.5,
                                                          face = "bold")))
```

Operators: `+` places side by side, `/` places one on top of the other.

---

# Animated Plots — `gganimate`

`gganimate` extends ggplot2 to create animated GIFs or videos by adding a time dimension.

```{r gganimate-demo, eval=FALSE}
library(gganimate)
library(gapminder)

ggplot(gapminder,
       aes(x = gdpPercap, y = lifeExp, size = pop, color = continent)) +
  geom_point(alpha = 0.7) +
  scale_x_log10() +
  labs(title = "Year: {frame_time}",
       x = "GDP per Capita", y = "Life Expectancy") +
  theme_bw() +
  transition_time(year)   # <-- the gganimate layer

# Save as GIF:
# anim_save("gapminder.gif")
```

Key `gganimate` functions:

| Function | Purpose |
|----------|---------|
| `transition_time(var)` | Animate over a continuous time variable |
| `transition_states(var)` | Animate over discrete states |
| `transition_reveal(var)` | Progressively reveal data along a variable |
| `anim_save("file.gif")` | Save the animation as a GIF |

---

# Saving Plots — `ggsave()`

```{r ggsave-demo, eval=FALSE}
my_plot <- ggplot(iris, aes(Sepal.Length, Sepal.Width, color = Species)) +
  geom_point(size = 2) +
  theme_bw()

# Save to a file — most common parameters:
ggsave("my_plot.png",
       plot   = my_plot,   # omit to save the last plot
       width  = 10,
       height = 7,
       units  = "cm",
       dpi    = 300)       # 300 dpi for publication quality

# PDF format (vector, infinitely scalable):
ggsave("my_plot.pdf", width = 10, height = 7)
```

You can also use the classic approach with `png()` / `pdf()` + `dev.off()`:

```{r dev-off-demo, eval=FALSE}
pdf("my_plot2.pdf", width = 8, height = 6)
print(my_plot)
dev.off()
```

---

# A Bioinformatics Application

Here is how the ggplot2 grammar applies directly to gene expression data. The code below illustrates the patterns — assume `dat.long` is a long-format RNA-seq data frame with columns `gene`, `samples`, `FPKM`, `tissue`, and `metastasis`.

```{r bioinformatics, eval=FALSE}
# 1 — Bar chart: BRCA1 expression by sample, colored by tissue type
dat.long %>%
  filter(gene == "BRCA1") %>%
  ggplot(aes(x = samples, y = FPKM, fill = tissue)) +
  geom_col() +
  labs(title = "BRCA1 Expression by Sample",
       x = "Sample", y = "FPKM") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# 2 — Density: BRCA1 expression distribution in normal vs tumor tissue
dat.long %>%
  filter(gene == "BRCA1") %>%
  ggplot(aes(x = FPKM, fill = tissue)) +
  geom_density(alpha = 0.4) +
  labs(title = "BRCA1 Expression Density by Tissue Type") +
  theme_bw()

# 3 — Boxplot: expression by metastasis status
dat.long %>%
  filter(gene == "BRCA1") %>%
  ggplot(aes(x = metastasis, y = FPKM)) +
  geom_boxplot() +
  geom_jitter(width = 0.15, alpha = 0.3, color = "tomato") +
  labs(title = "BRCA1 Expression by Metastasis Status") +
  theme_bw()

# 4 — Scatterplot with regression line: BRCA1 vs BRCA2
dat.long %>%
  filter(gene %in% c("BRCA1", "BRCA2")) %>%
  spread(key = gene, value = FPKM) %>%
  ggplot(aes(x = BRCA1, y = BRCA2, color = tissue)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "BRCA1 vs BRCA2 Expression Correlation") +
  theme_bw()

# 5 — Heatmap: multiple genes across all samples
genes.of.interest <- c("BRCA1", "BRCA2", "TP53", "ALK", "MYCN")
dat.long %>%
  filter(gene %in% genes.of.interest) %>%
  ggplot(aes(x = samples, y = gene, fill = FPKM)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "red") +
  labs(title = "Gene Expression Heatmap") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

---

# Quick Reference — Common geoms

| `geom_*` | Best for |
|----------|---------|
| `geom_point()` | Two continuous variables (scatter) |
| `geom_jitter()` | Scatter with overlap (adds random noise) |
| `geom_line()` | Time series, ordered sequences |
| `geom_bar()` | Counts of a categorical variable |
| `geom_col()` | Pre-computed bar heights |
| `geom_histogram()` | Distribution of one numeric variable |
| `geom_density()` | Smoothed distribution |
| `geom_boxplot()` | Distribution summary + outliers |
| `geom_violin()` | Full distribution shape |
| `geom_tile()` | Heatmaps |
| `geom_area()` | Area under a curve |
| `geom_smooth()` | Trend/regression line |
| `geom_errorbar()` | Uncertainty/confidence intervals |
| `geom_segment()` | Custom line segments |
| `geom_hline()` / `geom_vline()` | Reference lines |
| `annotate()` | Custom text or shapes |

---

# Practice Exercises

**Exercise 1.** Using the built-in `Indometh` dataset, make a scatterplot of `time` vs `conc`, then a boxplot of `conc` by `Subject`, and finally a faceted version showing one panel per subject.

**Exercise 2.** Using `msleep`, create a density plot showing the distribution of `sleep_total` for carnivores and herbivores side by side (use `fill` and `alpha`).

**Exercise 3.** Using the `esoph` dataset, calculate the frequency of esophageal cancer per observation (`ncases / (ncases + ncontrols)`) and make three separate boxplots to compare this frequency by `agegp`, `alcgp`, and `tobgp`. Add jitter to each.

**Exercise 4.** Using `gapminder` (install if needed), filter to Africa and Europe and create a faceted scatterplot of `gdpPercap` vs `lifeExp`, with point size proportional to `pop` and color encoding `year`. Apply `theme_bw()`.

**Exercise 5.** Customize any plot you've made by modifying: the title color and size, the axis label font style, the legend position, and the panel background. Use the `theme()` function.

**Exercise 6.** Use `patchwork` to combine two different plots about the same dataset into a single figure with a shared title.

---

```{r cleanup, include=FALSE}
rm(list = ls())
```

---

*End of Lecture — Data Visualization with ggplot2*
