---
title: "Lecture 6: Exploratory Data Analysis — A Complete Walkthrough"
author: "Ange Maxime TCHOUTANG"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width  = 8,
  fig.height = 5
)
```

---

# Learning Objectives

By the end of this lecture, you will be able to:

- Define **Exploratory Data Analysis (EDA)** and describe its place in the data science workflow
- Load data from **remote URLs** and perform initial inspection with `str()`, `glimpse()`, `summary()`
- Manipulate data with `tidyverse` tools: `arrange()`, `select()`, `filter()`, `group_by()`, `tally()`
- Create and interpret key EDA plots: **scatter plots, histograms, density plots, violin plots, and residual plots**
- Compute and test **correlations** (Pearson, Spearman, Kendall)
- Implement the **Tietjen-Moore test** for outlier detection
- Work with **normal distribution functions**: `dnorm()`, `pnorm()`, `qnorm()`, `rnorm()`
- Assess normality with the **Shapiro-Wilk test**

---

# What is Exploratory Data Analysis?

**Exploratory Data Analysis (EDA)** is the critical first step of any data analysis project. The goal is to understand what your data contains *before* fitting models or running formal tests.

> *"EDA is an attitude, a state of flexibility, a willingness to look for those things that we believe are not there, as well as those we believe to be there."*  
> — John W. Tukey (1977), the father of EDA

## The EDA Mindset

EDA answers four fundamental questions about your dataset:

1. **What does my data look like?** (structure, types, size)
2. **Is it clean?** (missing values, duplicates, impossible values)
3. **What is the distribution of each variable?** (histograms, density plots)
4. **How do variables relate to each other?** (scatter plots, correlations)

## Key ggplot2 Components (Glossary)

Before visualising, understand the grammar of graphics:

| Component | Description |
|---|---|
| **Data** | The dataset to visualise |
| **Layer** | Geometric elements (points, lines, bars) + statistical transformations |
| **Scales** | Maps data values to visual properties (colour, size, shape) |
| **Coordinate system** | Describes how coordinates are projected (Cartesian, polar) |
| **Faceting** | Splits data into sub-panels for multiple groups |
| **Theme** | Controls non-data elements (font, background, grid lines) |

---

# Required Packages

```{r load-packages}
library(tidyverse)    # dplyr, ggplot2, tidyr, readr, lubridate, etc.
library(lubridate)    # Date handling
library(gridExtra)    # Arranging multiple base-R plots
library(scales)       # Axis formatting
```

---

# Part 1: Loading and Inspecting Data

## Loading a Remote CSV File

R can read CSV files directly from a URL without downloading them first:

```{r load-remote, eval=FALSE}
site <- "https://raw.githubusercontent.com/nytimes/covid-19-data/master/colleges/colleges.csv"
college_Data <- read.csv(site)

# Save a local copy as an RData file
save(college_Data, file = "college_Data.RData")
```

```{r load-local, echo=FALSE}
# For reproducible demonstration, we simulate the data structure
set.seed(42)
college_Data <- data.frame(
  college      = paste("College", sample(LETTERS, 200, replace = TRUE), sample(1:99, 200, replace = TRUE)),
  state        = sample(c("Louisiana","Texas","Arkansas","Mississippi","California","New York"), 200, replace = TRUE),
  county       = paste("County", sample(1:30, 200, replace = TRUE)),
  city         = paste("City", sample(1:20, 200, replace = TRUE)),
  cases        = round(rnorm(200, mean = 200, sd = 150)),
  cases_2021   = round(rnorm(200, mean = 300, sd = 200))
)
college_Data$cases[college_Data$cases < 0] <- 0
college_Data$cases_2021[college_Data$cases_2021 < 0] <- 0
```

## Structure Inspection

```{r inspect-str}
str(college_Data)
```

> `str()` reveals the **structure** of any R object: number of observations, number of variables, variable names, data types, and a preview of values. This is always the **first function** to call on a new dataset.

```{r inspect-glimpse}
glimpse(college_Data)
```

> `glimpse()` from `{dplyr}` is a tidyverse-friendly alternative to `str()`. It transposes the output so you can see more variables at once — particularly useful for wide datasets.

```{r inspect-summary}
summary(college_Data)
```

> `summary()` provides descriptive statistics for each column: min, Q1, median, mean, Q3, max for numeric columns, and counts for factors.

---

# Part 2: Data Manipulation

## Sorting with `arrange()`

Sort data alphabetically or by any variable:

```{r arrange}
alphabetical <- college_Data %>%
  arrange(college)

head(alphabetical, 5)
```

## Selecting Columns with `select()`

Retain only the columns you need:

```{r select}
college_cases <- select(college_Data, college, cases)
head(college_cases)
```

## Filtering Rows with `filter()`

Extract observations matching a condition:

```{r filter-one-state}
louisiana_cases <- filter(college_Data, state == "Louisiana")
nrow(louisiana_cases)
```

Filter for **multiple states** using the `|` (OR) operator:

```{r filter-south}
south_cases <- filter(college_Data,
                      state == "Louisiana" |
                      state == "Texas"     |
                      state == "Arkansas"  |
                      state == "Mississippi")

cat("Rows in southern states subset:", nrow(south_cases), "\n")
table(south_cases$state)
```

## Grouping and Tallying

Count observations per group:

```{r group-tally}
state_cases   <- group_by(college_Data, state)
days_reported <- tally(state_cases)

days_reported %>% arrange(desc(n))
```

> `group_by()` + `tally()` tells you how many records each state has contributed — useful for understanding data coverage before visualisation.

---

# Part 3: Scatter Plots

Scatter plots reveal **relationships between two continuous variables**.

## College Data: Cases 2020 vs 2021

```{r scatter-college}
ggplot(data = college_Data,
       aes(x = cases, y = cases_2021)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  labs(
    title = "COVID-19 Cases: 2020 vs 2021 — US Colleges",
    x     = "Cases (2020)",
    y     = "Cases (2021)"
  ) +
  theme_minimal()
```

## Iris: Sepal Width vs Sepal Length

```{r scatter-iris}
ggplot(data = iris,
       aes(x = Sepal.Width, y = Sepal.Length)) +
  geom_point(alpha = 0.7) +
  labs(
    title = "Iris: Sepal Width vs Sepal Length",
    x     = "Sepal Width (cm)",
    y     = "Sepal Length (cm)"
  ) +
  theme_minimal()
```

## Adding Colour by a Categorical Variable

Map a third variable to the `color` aesthetic:

```{r scatter-color-college}
ggplot(data = college_Data,
       aes(x = cases, y = cases_2021, color = state)) +
  geom_point(alpha = 0.6) +
  labs(
    title = "COVID-19 Cases Coloured by State",
    x     = "Cases (2020)",
    y     = "Cases (2021)",
    color = "State"
  ) +
  theme_minimal()
```

```{r scatter-color-iris}
ggplot(data = iris,
       aes(x = Sepal.Width, y = Sepal.Length, color = Species)) +
  geom_point(alpha = 0.8, size = 2) +
  labs(
    title = "Iris: Sepal dimensions coloured by species",
    x     = "Sepal Width (cm)",
    y     = "Sepal Length (cm)",
    color = "Species"
  ) +
  theme_minimal()
```

> **Observation:** The iris data shows that *setosa* (typically lower right) is clearly separated from the other two species. *versicolor* and *virginica* overlap, suggesting that sepal dimensions alone are not sufficient to fully separate them.

---

# Part 4: Histograms

Histograms reveal the **distribution** of a single continuous variable.

## Base R Histogram

```{r hist-base}
hist(louisiana_cases$cases,
     freq    = TRUE,
     breaks  = 10,
     xlab    = "Total Cases",
     ylab    = "Frequency",
     main    = "COVID-19 Cases — Louisiana Colleges",
     col     = "steelblue",
     border  = "white")
```

## ggplot2 Histogram with Fill by Group

```{r hist-gg-college}
ggplot(data = louisiana_cases, aes(x = cases)) +
  geom_histogram(binwidth = 100,
                 color = "black",
                 aes(fill = county),
                 alpha = 0.8) +
  labs(
    title = "COVID-19 Cases — Louisiana Colleges by County",
    x     = "Cases",
    y     = "Frequency"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r hist-gg-iris}
ggplot(data = iris, aes(x = Sepal.Width)) +
  geom_histogram(binwidth = 0.2,
                 color    = "black",
                 aes(fill = Species),
                 alpha    = 0.7) +
  labs(
    title = "Iris Sepal Width Distribution by Species",
    x     = "Sepal Width (cm)",
    y     = "Frequency"
  ) +
  theme_minimal()
```

---

# Part 5: Density Plots

Density plots are **smoothed histograms**. They work well when comparing distributions across multiple groups because overlapping curves are easier to read than overlapping bars.

```{r density-south}
ggplot(south_cases) +
  geom_density(aes(x = cases, fill = state), alpha = 0.35) +
  labs(
    title = "Distribution of COVID-19 Cases — Southern US States",
    x     = "Cases",
    y     = "Density",
    fill  = "State"
  ) +
  theme_minimal()
```

```{r density-iris}
ggplot(iris) +
  geom_density(aes(x = Sepal.Width, fill = Species), alpha = 0.4) +
  labs(
    title = "Iris Sepal Width Density by Species",
    x     = "Sepal Width (cm)",
    y     = "Density"
  ) +
  theme_minimal()
```

---

# Part 6: Violin Plots

A **violin plot** combines a boxplot with a density estimate, showing both summary statistics and the full distributional shape.

```{r violin-iris}
ggplot(data = iris,
       aes(x = Species, y = Sepal.Length, color = Species, fill = Species)) +
  geom_violin(alpha = 0.4) +
  geom_boxplot(width = 0.1, color = "black", alpha = 0.8) +
  labs(
    title = "Iris Sepal Length by Species — Violin + Boxplot",
    x     = "Species",
    y     = "Sepal Length (cm)"
  ) +
  theme_classic() +
  theme(legend.position = "none")
```

```{r violin-south}
ggplot(data = south_cases,
       aes(x = state, y = cases, color = state, fill = state)) +
  geom_violin(alpha = 0.3) +
  labs(
    title = "COVID-19 Cases Distribution — Southern States",
    x     = "State",
    y     = "Cases"
  ) +
  theme_gray() +
  theme(legend.position = "none")
```

---

# Part 7: Residual Plots

A **residual plot** helps you decide whether a linear model is appropriate for your data. After fitting a linear model, plot fitted values on the x-axis and residuals on the y-axis:

- **Random scatter around y = 0** → linear model is appropriate
- **Pattern (U-shape, funnel)** → linear model is NOT appropriate; consider a non-linear model

```{r residuals-iris}
ggplot(lm(Sepal.Length ~ Sepal.Width, data = iris)) +
  geom_point(aes(x = .fitted, y = .resid),
             alpha = 0.6, color = "steelblue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title    = "Residual Plot — Iris: Sepal.Length ~ Sepal.Width",
    subtitle = "Random scatter around 0 = linear model is reasonable",
    x        = "Fitted values",
    y        = "Residuals"
  ) +
  theme_minimal()
```

```{r residuals-south}
ggplot(lm(cases ~ cases_2021, data = south_cases)) +
  geom_point(aes(x = .fitted, y = .resid),
             alpha = 0.6, color = "firebrick") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  labs(
    title    = "Residual Plot — College COVID Cases: 2020 ~ 2021",
    subtitle = "A pattern here would suggest a non-linear relationship",
    x        = "Fitted values",
    y        = "Residuals"
  ) +
  theme_minimal()
```

---

# Part 8: Correlation Analysis

## Pearson Correlation

The standard correlation coefficient for **linear** relationships between two continuous normally distributed variables:

```{r cor-pearson}
cor(iris$Sepal.Length, iris$Sepal.Width)
```

## Spearman and Kendall Correlations

For **non-normal** data or **ordinal** variables, use rank-based methods:

```{r cor-methods}
# Spearman (rank-based, robust to outliers)
cor(iris$Sepal.Length, iris$Sepal.Width, method = "spearman")

# Kendall's Tau (another rank-based measure)
cor(iris$Sepal.Length, iris$Sepal.Width, method = "kendall")
```

| Method | When to use |
|---|---|
| Pearson | Both variables are normally distributed, relationship is linear |
| Spearman | Non-normal data, or presence of outliers |
| Kendall | Small samples, or ordinal data |

## Significance Test

Test whether the correlation is statistically different from zero:

```{r cor-test}
cor.test(iris$Sepal.Length, iris$Sepal.Width)
```

> A p-value < 0.05 means the correlation is statistically significant. The confidence interval tells you the plausible range of the true population correlation.

---

# Part 9: The Tietjen-Moore Outlier Test

The **Tietjen-Moore test** detects a **known number k** of outliers in a univariate dataset assumed to be normally distributed. It is more powerful than the Grubbs test when multiple outliers are suspected.

## The Test Statistic

$$E_k = \frac{\sum_{i=1}^{n-k}(x_i - \bar{x}_{n-k})^2}{\sum_{i=1}^{n}(x_i - \bar{x})^2}$$

- Values close to **1** → no outliers
- Values close to **0** → strong evidence of outliers

## Implementation

```{r tietjenmoore-function}
TietjenMoore <- function(dataSeries, k) {
  n  <- length(dataSeries)
  
  # Absolute residuals from the mean
  r  <- abs(dataSeries - mean(dataSeries))
  
  # Sort by residual size
  df  <- data.frame(dataSeries, r)
  dfs <- df[order(df$r), ]
  
  # Subset excluding the k largest residuals
  klarge       <- c((n - k + 1):n)
  subdataSeries <- dfs$dataSeries[-klarge]
  
  # Sum of squares: subset vs full
  ksub <- (subdataSeries - mean(subdataSeries))^2
  all  <- (df$dataSeries - mean(df$dataSeries))^2
  
  sum(ksub) / sum(all)
}

findoutliersTietjenMooreTest <- function(dataSeries, k, alpha = 0.05) {
  ek <- TietjenMoore(dataSeries, k)
  
  # Generate reference distribution via simulation (10,000 normal samples)
  test <- numeric(10000)
  for (i in 1:10000) {
    sim     <- rnorm(length(dataSeries))
    test[i] <- TietjenMoore(sim, k)
  }
  
  Talpha <- quantile(test, alpha)
  list(T = ek, Talpha = Talpha,
       conclusion = ifelse(ek < Talpha,
                           "Evidence of outliers (T < Talpha)",
                           "No significant outliers (T >= Talpha)"))
}
```

## Application to the Iris Dataset

```{r tietjenmoore-apply}
# Test Petal.Width for 4 potential outliers
par(mfrow = c(1, 2))
boxplot(iris$Petal.Width,  main = "Petal Width",  ylab = "cm")
boxplot(iris$Petal.Length, main = "Petal Length", ylab = "cm")
par(mfrow = c(1, 1))
```

```{r tietjenmoore-test}
set.seed(123)  # For reproducibility of the simulation

result_pw <- findoutliersTietjenMooreTest(iris$Petal.Width, k = 4)
cat("Petal Width — T:", round(result_pw$T, 4),
    "| Talpha:", round(result_pw$Talpha, 4),
    "| Result:", result_pw$conclusion, "\n")

result_pl <- findoutliersTietjenMooreTest(iris$Petal.Length, k = 2)
cat("Petal Length — T:", round(result_pl$T, 4),
    "| Talpha:", round(result_pl$Talpha, 4),
    "| Result:", result_pl$conclusion, "\n")
```

---

# Part 10: Probability and Normal Distribution Functions

Understanding the normal distribution is fundamental for statistical inference.

## The Four Normal Distribution Functions

| Function | Question it answers |
|---|---|
| `dnorm(x, mean, sd)` | What is the probability **density** at value x? |
| `pnorm(x, mean, sd)` | What is the probability of getting a value **≤ x**? |
| `qnorm(p, mean, sd)` | What value corresponds to the **p-th percentile**? |
| `rnorm(n, mean, sd)` | Generate **n random values** from this distribution |

```{r normal-dist-setup}
sepal.mean <- mean(iris$Sepal.Length)
sepal.sd   <- sd(iris$Sepal.Length)

cat("Sepal Length mean:", round(sepal.mean, 3), "\n")
cat("Sepal Length SD:  ", round(sepal.sd, 3), "\n")
```

## Density Function (`dnorm`)

```{r dnorm-plot}
sepal.density <- dnorm(iris$Sepal.Length,
                       mean = sepal.mean,
                       sd   = sepal.sd)

sepal.df <- data.frame(sepal = iris$Sepal.Length,
                       density = sepal.density)

ggplot(sepal.df, aes(x = sepal, y = density)) +
  geom_point(color = "steelblue", alpha = 0.7) +
  geom_line(stat = "smooth", method = "loess", color = "red", se = FALSE) +
  labs(
    title = "Probability density of Iris Sepal Length",
    x     = "Sepal Length (cm)",
    y     = "Density"
  ) +
  theme_minimal()
```

## Cumulative Distribution Function (`pnorm`)

```{r pnorm-plot}
sepal.cum <- pnorm(iris$Sepal.Length,
                   mean = sepal.mean,
                   sd   = sepal.sd)

sepal.cdf <- data.frame(sepal = iris$Sepal.Length, cumulative = sepal.cum)

ggplot(sepal.cdf, aes(x = sepal, y = cumulative)) +
  geom_point(color = "darkorange", alpha = 0.7) +
  labs(
    title    = "Cumulative distribution of Iris Sepal Length",
    subtitle = "P(X ≤ x) — the probability that sepal length is at most x",
    x        = "Sepal Length (cm)",
    y        = "Cumulative probability"
  ) +
  theme_minimal()
```

## Quantile Function (`qnorm`) and Random Sampling (`rnorm`)

```{r qnorm}
# What sepal length represents the bottom 1% of the distribution?
bottom_1pct <- qnorm(0.01, mean = sepal.mean, sd = sepal.sd)
cat("Bottom 1% of sepal length:", round(bottom_1pct, 3), "cm\n")

# What value cuts off the top 5%?
top_5pct <- qnorm(0.95, mean = sepal.mean, sd = sepal.sd)
cat("Top 5% threshold:", round(top_5pct, 3), "cm\n")
```

```{r rnorm-demo, fig.height=4}
set.seed(42)
small_sample <- rnorm(50,   mean = sepal.mean, sd = sepal.sd)
large_sample <- rnorm(5000, mean = sepal.mean, sd = sepal.sd)

p_small <- ggplot(data.frame(x = small_sample), aes(x)) +
  geom_histogram(bins = 10, fill = "steelblue", color = "white") +
  labs(title = "n = 50 samples", x = "Sepal Length") +
  theme_minimal()

p_large <- ggplot(data.frame(x = large_sample), aes(x)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  labs(title = "n = 5000 samples", x = "Sepal Length") +
  theme_minimal()

gridExtra::grid.arrange(p_small, p_large, ncol = 2)
```

> **Central Limit Theorem in action:** With only 50 samples, the histogram is ragged. With 5000 samples, the bell curve shape becomes clearly visible.

---

# Part 11: Shapiro-Wilk Normality Test

Before using parametric tests (t-test, ANOVA, Pearson correlation), verify that your data is normally distributed:

- $H_0$: The data **comes from** a normal distribution
- $H_1$: The data **does not come from** a normal distribution

```{r shapiro-small}
# Small sample (n = 5) — low power
shapiro.test(iris$Sepal.Length[1:5])
```

```{r shapiro-full}
# Full sample (n = 150)
shapiro.test(iris$Sepal.Length)
```

> **Interpretation:** With a very small sample (n = 5), we lack power to detect non-normality. With the full sample (n = 150), the p-value is significant — meaning the sepal lengths are **not** perfectly normally distributed. However, with large samples, the Shapiro-Wilk test is very sensitive and may reject normality even for minor departures that are practically unimportant. Always pair the test with a Q-Q plot:

```{r qqplot}
qqnorm(iris$Sepal.Length, main = "Q-Q Plot — Iris Sepal Length")
qqline(iris$Sepal.Length, col = "red", lwd = 2)
```

> If points follow the red line closely, the data is approximately normal.

---

# Summary

| Topic | Key Function(s) |
|---|---|
| Data inspection | `str()`, `glimpse()`, `summary()` |
| Manipulation | `arrange()`, `select()`, `filter()`, `group_by()`, `tally()` |
| Scatter plot | `geom_point()` |
| Histogram | `geom_histogram()` |
| Density plot | `geom_density()` |
| Violin plot | `geom_violin()` |
| Residual plot | `ggplot(lm(...)) + geom_point(aes(.fitted, .resid))` |
| Correlation | `cor()`, `cor.test()` |
| Tietjen-Moore test | Custom function (T close to 0 = outliers present) |
| Normal distribution | `dnorm()`, `pnorm()`, `qnorm()`, `rnorm()` |
| Normality test | `shapiro.test()` |

---

# Practice Exercises

1. Load the `mtcars` dataset. Use `str()`, `glimpse()`, and `summary()` to describe its structure. How many numeric and how many factor-type variables are there?

2. Filter `mtcars` to keep only cars with 6 or 8 cylinders (`cyl`). Create a density plot of `mpg` for these two groups with different fill colours.

3. Fit a linear model `mpg ~ wt` on the full `mtcars` dataset. Create a residual plot. Is the linear model appropriate?

4. Use `cor.test()` to test the correlation between `hp` and `mpg` in `mtcars`. Report the Pearson r, p-value, and 95% confidence interval. Is the relationship significant?

5. *Challenge:* Apply the Tietjen-Moore test to `iris$Sepal.Width` for k = 3 potential outliers. Compare with a boxplot. Do the results agree?

---

# References

- Tukey, J. W. (1977). *Exploratory Data Analysis*. Addison-Wesley.
- Wickham, H. & Grolemund, G. (2017). *R for Data Science*. https://r4ds.had.co.nz
- Tietjen, G. L., Moore, R. H. (1972). Some Grubbs-type statistics for the detection of several outliers. *Technometrics*, 14(3), 583–597.
