---
title: "R Data Types and Data Structures"
subtitle: "A Comprehensive Lecture — From Atomic Types to Data Frames and Import"
author: "Ange Maxime TCHOUTANG"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    theme: flatly
    highlight: tango
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  comment = "##",
  message = FALSE,
  warning = FALSE
)
```

---

# Introduction

Every piece of information stored in R has a **type** — a description of what kind of value it is — and a **structure** — a description of how values are organized together. Understanding both is foundational to writing correct, efficient R code.

In this lecture we cover:

- **Atomic data types** — the building blocks of all R objects
- **Vectors** — the simplest multi-value structure
- **Factors** — categorical data with levels
- **Matrices** — two-dimensional rectangular tables of one type
- **Arrays** — matrices generalized to N dimensions
- **Lists** — flexible containers for mixed-type data
- **Data Frames** — the workhorse of data analysis in R
- **Importing and exporting data** — reading real datasets from files

---

# Atomic Data Types

Before building complex structures, you need to understand the five basic data types in R. Every value in R belongs to exactly one of these types.

| Type | Description | Example | Suffix |
|---|---|---|---|
| `numeric` | Decimal numbers (default) | `10.5`, `3.14` | — |
| `integer` | Whole numbers | `5L`, `100L` | `L` |
| `complex` | Complex numbers | `3 + 2i` | `i` |
| `character` | Text / strings | `"Hello"`, `"TP53"` | `" "` |
| `logical` | Boolean TRUE/FALSE | `TRUE`, `FALSE` | — |

Use `class()` to check the type of any object, and `is.*()` / `as.*()` to test and convert types.

```{r data-types}
# Numeric (double — stores decimals)
x <- 10.5
class(x)

# Integer (whole number — must use L suffix)
x <- 1000L
class(x)

# Complex (has a real and imaginary part)
x <- 9 + 3i
class(x)

# Character (text string)
x <- "R is exciting"
class(x)

# Logical (Boolean)
x <- TRUE
class(x)
```

## Checking and Converting Types

```{r type-checking}
age <- 25.7

is.numeric(age)     # Is it numeric?
is.integer(age)     # Is it an integer?
is.character(age)   # Is it a string?

# Convert between types
as.integer(age)     # Truncates decimal → 25
as.character(age)   # → "25.7"
as.numeric("42.3")  # String → number
as.logical(0)       # 0 → FALSE
as.logical(1)       # 1 → TRUE
```

> 💡 `TRUE` and `FALSE` are treated as `1` and `0` in arithmetic operations. This makes logicals extremely useful for counting conditions.

```{r logical-arithmetic}
TRUE + TRUE     # 2
FALSE + TRUE    # 1
sum(c(TRUE, FALSE, TRUE, TRUE))  # counts TRUE values → 3
```

## Useful Math Functions

R has many built-in mathematical functions that work on numeric values:

```{r math-functions}
sqrt(16)         # Square root
abs(-7.5)        # Absolute value
ceiling(4.2)     # Round up to nearest integer
floor(4.8)       # Round down to nearest integer
round(3.14159, 2) # Round to 2 decimal places
log(10)          # Natural logarithm
log10(100)       # Base-10 logarithm
exp(1)           # Euler's number e^1

max(5, 10, 15)   # Maximum value
min(5, 10, 15)   # Minimum value
```

---

# Vectors

A **vector** is R's most fundamental data structure — an ordered collection of values of **the same type**. Even a single value like `x <- 5` is technically a vector of length 1.

## Creating Vectors

```{r vector-create}
# Character vector
fruits <- c("banana", "apple", "orange", "mango")

# Numeric vector
numbers <- c(10, 20, 30, 40, 50)

# Logical vector
flags <- c(TRUE, FALSE, TRUE, TRUE)

# Integer sequence with the colon operator
seq1 <- 1:10
seq1

# Custom sequence with seq()
seq2 <- seq(from = 0, to = 100, by = 20)
seq2

# Repeated values with rep()
rep(c(1, 2, 3), times = 3)   # repeat whole sequence 3 times
rep(c(1, 2, 3), each  = 2)   # repeat each element 2 times
```

> 💡 **The `c()` function** stands for "combine" or "concatenate". It is the primary tool for creating vectors and works for all data types.

## Exploring Vectors

```{r vector-explore}
fruits <- c("banana", "apple", "orange", "mango", "lemon")

length(fruits)    # Number of elements
sort(fruits)      # Sort alphabetically
rev(fruits)       # Reverse order
unique(c(1, 2, 2, 3, 3, 3))  # Unique values only
```

## Accessing Vector Elements

R uses **1-based indexing** — the first element is at position 1 (not 0 like Python or C).

```{r vector-access}
fruits <- c("banana", "apple", "orange", "mango", "lemon")

fruits[1]           # First element
fruits[3]           # Third element
fruits[c(1, 3, 5)]  # Elements 1, 3, and 5
fruits[2:4]         # Elements 2 through 4
fruits[-1]          # All except the first
fruits[-c(1, 3)]    # All except positions 1 and 3
```

## Modifying Vector Elements

```{r vector-modify}
fruits <- c("banana", "apple", "orange")

fruits[1] <- "pear"   # Change first element
fruits
```

## Vector Operations

Arithmetic applied to a vector operates **element-by-element**:

```{r vector-ops}
x <- c(1, 2, 3, 4, 5)

x + 10        # Add 10 to each element
x * 2         # Multiply each by 2
x^2           # Square each element

x1 <- c(1, 2, 3)
x2 <- c(10, 20, 30)
x1 + x2       # Element-wise addition
```

## Named Vectors

```{r named-vector}
scores <- c(Alice = 92, Bob = 87, Carla = 95)
scores["Alice"]           # Access by name
names(scores)             # Get all names
```

## Useful Vector Functions

```{r vector-functions}
v <- c(4, 7, 2, 9, 1, 5, 8, 3, 6)

sum(v)
mean(v)
median(v)
var(v)
sd(v)
range(v)      # min and max
which.max(v)  # index of the maximum
which.min(v)  # index of the minimum
table(c("A", "T", "G", "A", "C", "A"))  # frequency table
```

---

# Factors

A **factor** is R's data structure for **categorical variables** — variables with a fixed set of possible values (called **levels**). Factors are essential for statistical modelling and controlled plotting.

```{r factor-create}
music_genre <- factor(c("Jazz", "Rock", "Classic", "Classic", "Pop", "Jazz", "Rock", "Jazz"))

music_genre
levels(music_genre)   # The unique categories
nlevels(music_genre)  # Number of levels
length(music_genre)   # Total number of items
```

## Setting Custom Level Order

By default levels are ordered alphabetically. You can override this:

```{r factor-levels}
# Set a custom order
sizes <- factor(c("Small", "Large", "Medium", "Small", "Large"),
                levels = c("Small", "Medium", "Large"))
levels(sizes)

# Ordered factor (for ordinal data — allows < > comparisons)
severity <- factor(c("Low", "High", "Medium", "Low"),
                   levels  = c("Low", "Medium", "High"),
                   ordered = TRUE)
severity
severity[1] < severity[2]   # Can compare because ordered = TRUE
```

## Accessing and Changing Factor Values

```{r factor-access}
music_genre <- factor(c("Jazz", "Rock", "Classic", "Pop"))

music_genre[2]           # Access by position

# Change a value (new value must be an existing level)
music_genre[2] <- "Jazz"
music_genre[2]
```

> ⚠️ You cannot assign a value to a factor that is not already one of its levels. If you need to add a new category, you must redefine the factor with the new level included.

## Counting Factor Levels

```{r factor-table}
sex <- factor(c("Male", "Female", "Male", "Male", "Female"))
table(sex)     # Frequency table — very useful in descriptive stats
```

---

# Matrices

A **matrix** is a **two-dimensional** collection of values of **the same type**, organized in rows and columns. It is essentially a vector with a shape attribute.

## Creating a Matrix

```{r matrix-create}
# Fill by column (default, byrow = FALSE)
m1 <- matrix(c(1, 2, 3, 4, 5, 6), nrow = 3, ncol = 2)
m1

# Fill by row (byrow = TRUE)
m2 <- matrix(c(1, 2, 3, 4, 5, 6), nrow = 3, ncol = 2, byrow = TRUE)
m2

# Matrix of character strings
cities <- matrix(c("London", "Paris", "Cairo", "Tokyo"), nrow = 2, ncol = 2)
cities
```

## Naming Rows and Columns

```{r matrix-names}
expr_matrix <- matrix(c(2.1, 3.2, 4.5,
                        1.5, 2.8, 3.9),
                      nrow = 2, ncol = 3)

rownames(expr_matrix) <- c("Gene1", "Gene2")
colnames(expr_matrix) <- c("Sample1", "Sample2", "Sample3")
expr_matrix
```

## Accessing Matrix Elements

```{r matrix-access}
m <- matrix(c("apple","banana","cherry","orange",
              "grape","pineapple","pear","melon","fig"),
            nrow = 3, ncol = 3)

m[1, 2]        # Row 1, Column 2
m[2, ]         # Entire row 2
m[, 3]         # Entire column 3
m[c(1, 2), ]   # Rows 1 and 2
m[, c(1, 3)]   # Columns 1 and 3

# Access by name when names are set
expr_matrix["Gene1", "Sample2"]
expr_matrix["Gene2", ]
```

## Adding and Removing Rows/Columns

```{r matrix-modify}
m <- matrix(c("London","Oxford","Cairo","Austin"), nrow = 2, ncol = 2)

# Add a column
cbind(m, c("Dubai", "Tokyo"))

# Add a row
rbind(m, c("City1", "City2"))

# Remove first row and first column
m[-1, -1]
```

## Matrix Dimensions and Membership

```{r matrix-info}
m <- matrix(c("apple","banana","cherry","orange"), nrow = 2, ncol = 2)

dim(m)            # Rows × Columns
nrow(m)           # Number of rows
ncol(m)           # Number of columns
length(m)         # Total cells = nrow × ncol

"apple" %in% m    # Does "apple" exist in the matrix?
"grape" %in% m    # Does "grape" exist?
```

## Matrix Arithmetic and Summary

```{r matrix-math}
num_mat <- matrix(1:12, nrow = 3, ncol = 4)

rowSums(num_mat)    # Sum of each row
colSums(num_mat)    # Sum of each column
rowMeans(num_mat)   # Mean of each row
colMeans(num_mat)   # Mean of each column

# Combine with the result
cbind(num_mat, row_total = rowSums(num_mat))
```

## Looping over a Matrix

```{r matrix-loop}
m <- matrix(c("a", "b", "c", "d"), nrow = 2, ncol = 2)

for (row in 1:nrow(m)) {
  for (col in 1:ncol(m)) {
    cat("Row", row, "Col", col, "→", m[row, col], "\n")
  }
}
```

## Combining Two Matrices

```{r matrix-combine}
M1 <- matrix(c(1, 2, 3, 4), nrow = 2)
M2 <- matrix(c(5, 6, 7, 8), nrow = 2)

rbind(M1, M2)   # Stack vertically (add rows)
cbind(M1, M2)   # Place side by side (add columns)
```

---

# Arrays

An **array** is a generalization of a matrix to **N dimensions**. While a matrix is 2D (rows × columns), an array can be 3D (rows × columns × layers) or higher.

```{r array-create}
# One-dimensional array (essentially a vector)
one_dim <- c(1:24)

# Three-dimensional array: 4 rows × 3 columns × 2 layers
multi_dim <- array(one_dim, dim = c(4, 3, 2))
multi_dim
```

## Accessing Array Elements

```{r array-access}
# Syntax: array[row, column, layer]
multi_dim[1, , 1]   # All columns of row 1 in layer 1
multi_dim[, 2, 1]   # All rows of column 2 in layer 1
multi_dim[2, 3, 2]  # Row 2, Column 3, Layer 2
```

## Naming Array Dimensions

```{r array-named}
regions <- c("Centre", "Littoral", "Far_North")
months  <- c("Jan", "Feb", "Mar")
years   <- c("2023", "2024")

malaria <- array(
  data = c(450, 390, 820, 520, 440, 900, 610, 500, 950,
           480, 410, 850, 550, 470, 920, 640, 515, 980),
  dim  = c(3, 3, 2),
  dimnames = list(Region = regions, Month = months, Year = years)
)

malaria
malaria["Centre", "Jan", "2023"]          # Single cell
malaria["Far_North", , "2024"]            # All months for Far North in 2024
apply(malaria, 1, mean)                   # Mean per region across months and years
```

## Array Dimensions and Membership

```{r array-info}
dim(multi_dim)      # Dimensions (4 × 3 × 2)
length(multi_dim)   # Total elements = 4 × 3 × 2 = 24

3  %in% multi_dim   # Is 3 in the array?
100 %in% multi_dim  # Is 100 in the array?
```

---

# Lists

A **list** is R's most flexible data structure. Unlike vectors, matrices, and arrays — which require all elements to be the same type — a list can contain **any mix of types and structures**, including other lists.

## Creating a List

```{r list-create}
# Simple list
my_list <- list("apple", 42, TRUE)
my_list

# Named list — recommended for clarity
patient <- list(
  name    = "Alice",
  age     = 34,
  scores  = c(90, 85, 88),
  dataset = head(mtcars, 3)
)
patient
```

## Accessing List Elements

Lists have two kinds of access:

- **`[ ]` (single bracket)** — returns a **sub-list** (type stays as list)
- **`[[ ]]` (double bracket)** — extracts the **actual element**
- **`$` operator** — extracts by name (equivalent to `[[ ]]`)

```{r list-access}
my_list <- list("apple", 42, TRUE, c(1, 2, 3))

my_list[1]      # Returns a list of length 1 — still a list
my_list[[1]]    # Returns "apple" directly
my_list[[4]]    # Returns the vector c(1, 2, 3)
my_list[[4]][2] # Second element of the vector inside the list

# Named access
patient$name         # "Alice"
patient[["scores"]]  # c(90, 85, 88)
patient$scores[1]    # 90
```

## Modifying Lists

```{r list-modify}
my_list <- list("apple", "banana", "cherry")

# Change an element
my_list[[1]] <- "pear"
my_list

# Add an element
my_list <- append(my_list, "orange")
my_list <- append(my_list, "kiwi", after = 2)   # After position 2
my_list

# Remove an element
my_list[-1]   # All except position 1
```

## List Properties

```{r list-info}
my_list <- list("apple", "banana", "cherry")

length(my_list)
"apple" %in% my_list   # Membership test

# Loop through a list
for (item in my_list) {
  print(item)
}

# Join two lists
list1 <- list("a", "b")
list2 <- list(1, 2)
c(list1, list2)
```

## Nested Lists

Lists can contain other lists, making them ideal for storing hierarchical data:

```{r nested-list}
gene_db <- list(
  metadata = data.frame(
    gene     = c("BRCA1", "TP53", "EGFR"),
    function_ = c("DNA repair", "Tumor suppressor", "Cell growth")
  ),
  sequences  = c("ATGCGATC", "GCTAGCTA", "TTGCATGC"),
  expression = matrix(c(2.1, 3.2, 1.5, 2.8, 5.5, 6.7),
                      nrow = 3, ncol = 2,
                      dimnames = list(c("BRCA1","TP53","EGFR"),
                                      c("Sample1","Sample2")))
)

# Access nested elements
gene_db$metadata
gene_db$sequences[2]
gene_db[["expression"]]["TP53", ]
gene_db[[3]]["BRCA1", "Sample2"]
```

---

# Data Frames

The **data frame** is the most commonly used data structure in R for data analysis. It is a **table** where:

- Each **column** is a variable (and can be a different type)
- Each **row** is an observation
- All columns must have the **same number of rows**

Think of it as a spreadsheet or a SQL table.

## Creating a Data Frame

```{r df-create}
# Build manually
athletes <- data.frame(
  player = c("Ronaldo", "Messi", "Salah"),
  number = c(7, 30, 11),
  height = c(187, 169, 175),
  stringsAsFactors = FALSE
)

athletes
```

## Inspecting a Data Frame

```{r df-inspect}
data(airquality)   # Built-in dataset: NYC air quality

dim(airquality)    # Rows × Columns
nrow(airquality)   # Number of rows
ncol(airquality)   # Number of columns
length(airquality) # Same as ncol — number of variables

str(airquality)    # Structure: column names, types, preview
head(airquality)   # First 6 rows
head(airquality, 10) # First 10 rows
tail(airquality)   # Last 6 rows
names(airquality)  # Column names
summary(airquality)  # Statistical summary of each column
```

## Accessing Data Frame Elements

```{r df-access}
df <- data.frame(
  Training = c("Strength", "Stamina", "Other"),
  Pulse    = c(100, 150, 120),
  Duration = c(60, 30, 45)
)

# By column position
df[1]           # Returns a 1-column data frame (keeps class)
df[[1]]         # Returns the vector directly
df[, 1]         # Also returns the vector

# By column name
df["Pulse"]         # 1-column data frame
df[["Pulse"]]       # Vector
df$Pulse            # Vector (most common syntax)

# By row
df[2, ]             # Row 2, all columns
df[1:2, ]           # Rows 1 and 2
df[1, "Duration"]   # Row 1, column "Duration"

# Subset rows with a condition
subset(airquality, Wind > 15)
airquality[airquality$Temp > 90, ]
```

## Adding Rows and Columns

```{r df-add}
athletes <- data.frame(
  player = c("Ronaldo", "Messi", "Salah"),
  number = c(7, 30, 11),
  height = c(187, 169, 175)
)

# Add a column with $
athletes$goals <- c(700, 800, 250)
athletes

# Add a column with cbind()
cbind(athletes, nationality = c("Portugal", "Argentina", "Egypt"))

# Add a row with rbind()
rbind(athletes, data.frame(player = "Neymar",
                           number = 10,
                           height = 175,
                           goals  = 400))
```

## Removing Rows and Columns

```{r df-remove}
df <- data.frame(
  A = c(1, 2, 3),
  B = c(4, 5, 6),
  C = c(7, 8, 9)
)

df[-1, ]       # Remove row 1
df[, -2]       # Remove column 2
df[-c(1, 3), ] # Remove rows 1 and 3
```

## Combining Data Frames

```{r df-combine}
df1 <- data.frame(name = c("Alice", "Bob"), score = c(90, 85))
df2 <- data.frame(name = c("Carla", "Dave"), score = c(78, 92))

# Stack vertically (same columns)
rbind(df1, df2)

# Place side by side (same number of rows)
df3 <- data.frame(grade = c("A", "B"))
cbind(df1, df3)
```

## Handling Missing Values in Data Frames

```{r df-missing}
df <- data.frame(x = c(1, NA, 3), y = c(NA, 5, 6))

is.na(df)                     # Which cells are NA?
colSums(is.na(df))            # Count NAs per column
complete.cases(df)            # Which rows have no NAs?
df[complete.cases(df), ]      # Keep only complete rows
na.omit(df)                   # Same as above

# Replace NAs
df$x[is.na(df$x)] <- 0       # Replace NA in column x with 0
```

## Type Conversion in Data Frames

```{r df-types}
df <- data.frame(age = c("25", "30", "22"), score = c(90, 85, 95))

str(df)                          # age is character — wrong type

df$age <- as.integer(df$age)     # Fix: convert to integer
str(df)

# Available conversions:
# as.numeric(), as.integer(), as.character(), as.factor(), as.logical()
```

## Ordering a Data Frame

```{r df-order}
airquality[order(airquality$Temp), ]            # Ascending by Temp
airquality[order(-airquality$Temp), ]           # Descending
airquality[order(airquality$Month,
                 airquality$Temp), ]            # Sort by 2 columns
```

---

# Comparison: All Data Structures

| Feature | Vector | Factor | Matrix | Array | List | Data Frame |
|---|---|---|---|---|---|---|
| Dimensions | 1D | 1D | 2D | N-D | 1D | 2D |
| Types allowed | One | One (levels) | One | One | Mixed | Mixed (per column) |
| Named elements | Yes | Yes (levels) | Yes (row/col) | Yes | Yes | Yes (col names) |
| Main use | Sequences | Categories | Numeric tables | Multi-dim data | Heterogeneous | Dataset/table |
| Create with | `c()` | `factor()` | `matrix()` | `array()` | `list()` | `data.frame()` |

---

# Importing and Exporting Data

In real-world analysis, data rarely comes from typing it manually. You will read data from CSV files, Excel spreadsheets, tab-separated files, or URLs.

## Reading CSV Files

```{r import-csv, eval=FALSE}
# Classic (returns a data.frame)
data <- read.csv("myfile.csv")

# Modern tidyverse way (returns a tibble)
library(readr)
data <- read_csv("myfile.csv")

# View the working directory
getwd()

# Change it
setwd("path/to/your/data")
```

## Reading Tab-Separated Files (TSV)

```{r import-tsv, eval=FALSE}
annotations <- read.delim("annotations.tsv")       # sep = "\t" by default
annotations <- read.table("annotations.tsv",
                          header = TRUE, sep = "\t")
```

## Reading Excel Files

```{r import-excel, eval=FALSE}
# install.packages("readxl")
library(readxl)

pathway_data <- read_excel("pathway_data.xlsx", sheet = 1)
head(pathway_data)
```

## Importing Data Directly from a URL

```{r import-url, eval=FALSE}
url  <- "https://raw.githubusercontent.com/HackBio-Internship/public_datasets/main/metadata_sample.tsv"
data <- read.csv(url)
dim(data)
```

## Inspecting an Imported Data Frame

```{r inspect-pattern, eval=FALSE}
str(data)                     # Structure and types
glimpse(data)                 # dplyr version (compact)
head(data)                    # First 6 rows
dim(data)                     # Rows × Columns
colSums(is.na(data))          # Missing values per column
sapply(data, class)           # Type of each column
sapply(data, function(x) length(unique(x)))  # Unique values per column
```

## Exporting Data

```{r export, eval=FALSE}
# Write to CSV
write.csv(data, file = "output.csv", row.names = FALSE)

# Save R objects (.Rda format — preserves all R types)
save(data, file = "data.Rda")

# Reload later
load("data.Rda")

# rio package — universal import/export
# install.packages("rio")
library(rio)
export(data, file = "output.xlsx")   # Write to Excel
export(data, file = "output.csv")    # Write to CSV
data2 <- import("myfile.csv")        # Read any format
```

## data.frame vs tibble vs data.table

```{r df-types-compare, eval=FALSE}
library(tibble)
library(data.table)

# Classic data frame
df  <- data.frame(name = c("Alice", "Bob"), age = c(25, 30))

# Tibble — tidyverse format, better printing, no rownames
tb  <- tibble(name = c("Alice", "Bob"), age = c(25, 30))

# data.table — optimized for large datasets (millions of rows)
dt  <- data.table(name = c("Alice", "Bob"), age = c(25, 30))

# Convert between formats
data.frame(tb)      # tibble → data.frame
as_tibble(df)       # data.frame → tibble
as.data.table(df)   # data.frame → data.table
```

---

# A Bioinformatics Application

Here is how all data structures come together in a real genomics workflow:

```{r bioinformatics-example}
# 1. Vector — DNA sequence
dna <- c("A", "T", "G", "C", "A", "T", "G", "A")

# Nucleotide frequencies
freq <- table(dna)
print(freq)

# Most frequent nucleotide
most_freq <- names(which.max(freq))
cat("Most frequent nucleotide:", most_freq, "\n")

# 2. Matrix — gene expression
expr <- matrix(c(2.1, 3.2, 4.5,
                 1.5, 2.8, 3.9,
                 5.5, 6.7, 2.6,
                 7.4, 1.6, 3.5,
                 4.5, 2.5, 6.3),
               nrow = 5, ncol = 3)

rownames(expr) <- c("BRCA1", "TP53", "EGFR", "MYC", "ALK")
colnames(expr) <- c("Control", "Treatment1", "Treatment2")

# Mean expression per gene
rowMeans(expr)

# 3. Data Frame — sample metadata
metadata <- data.frame(
  sample     = c("Control", "Treatment1", "Treatment2"),
  group      = factor(c("CTRL", "TREAT", "TREAT")),
  replicate  = c(1L, 1L, 2L)
)
metadata

# 4. List — bring it all together
experiment <- list(
  sequences  = dna,
  expression = expr,
  metadata   = metadata
)

# Access nested elements
experiment$metadata$group
experiment[["expression"]]["TP53", ]
```

---

# Practice Exercises

**Exercise 1 — Data Types.** Create one variable of each type: `numeric`, `integer`, `complex`, `character`, and `logical`. Use `class()` to verify each. Then convert the numeric to integer and the character `"42.5"` to numeric.

**Exercise 2 — Vectors.** Create a numeric vector with the values `c(5, 12, 3, 8, 1, 9, 4, 7, 2, 6)`. Compute its mean, median, and standard deviation. Extract all values greater than 5. Sort it in descending order.

**Exercise 3 — Factors.** Create a character vector of 10 observations of blood type: `"A"`, `"B"`, `"AB"`, `"O"`. Convert it to a factor. Use `table()` to count frequencies. Then create an ordered factor for disease severity: `"Mild"`, `"Moderate"`, `"Severe"`. Verify that `"Mild" < "Severe"` returns `TRUE`.

**Exercise 4 — Matrix.** Create a 4×3 matrix filled with values 1 to 12 (by row). Name the rows `"R1"`–`"R4"` and columns `"C1"`–`"C3"`. Extract column 2. Compute row sums and column means. Add a new column that is the row mean of each row.

**Exercise 5 — List.** Create a list representing a patient with fields: `name` (character), `age` (integer), `measurements` (a numeric vector of 5 blood pressure readings), and `history` (a data frame with 2 columns: `date` and `diagnosis`). Access each element and print the second blood pressure reading.

**Exercise 6 — Data Frame.** Using the built-in `airquality` dataset: (1) Check its dimensions and structure. (2) Find how many NAs are in each column. (3) Keep only rows where `Temp > 85` and `Wind < 10`. (4) Add a new column `Temp_C = (Temp - 32) * 5/9`. (5) Sort by `Ozone` in descending order.

---

```{r cleanup, include=FALSE}
rm(list = ls())
```

---

*End of Lecture — R Data Types and Data Structures*
