---
title: "Descriptive Statistics in R"
subtitle: "A Comprehensive Lecture — From Central Tendency to Probability Distributions"
author: "Ange Maxime TCHOUTANG"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    theme: flatly
    highlight: tango
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  comment = "##",
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 4.5
)
```

```{r packages}
library(tidyverse)
```

---

# Introduction

**Descriptive statistics** summarizes and describes the main features of a dataset. Rather than drawing conclusions about a broader population (that is inferential statistics), descriptive statistics simply tells us *what is in our data*: how values are distributed, where they cluster, and how spread out they are.

In this lecture we cover:

- **Types of variables** — quantitative vs categorical, discrete vs continuous
- **Exploring a dataset** — `str()`, `summary()`, `dim()`, `head()`
- **Measures of central tendency** — mean, median, mode
- **Measures of dispersion** — range, variance, standard deviation, IQR
- **Measures of position** — quartiles, percentiles, confidence intervals
- **Frequency tables** — `table()`, `prop.table()`, `addmargins()`
- **Cross-tabulation** — two-way and three-way tables
- **Grouped statistics** — `tapply()`, `aggregate()`, `dplyr`
- **Handling missing values** — `na.rm`, `is.na()`, `na.omit()`
- **Visual summaries** — histograms, boxplots, density plots
- **Advanced packages** — `psych::describe()`, `Hmisc::describe()`
- **Probability distributions** — normal, binomial, Poisson, uniform

---

# Types of Variables

Understanding what type of variable you are working with determines which statistical tools are appropriate.

| Type | Sub-type | Examples | R class |
|---|---|---|---|
| **Quantitative** | Continuous | Age, weight, temperature | `numeric` |
| **Quantitative** | Discrete | Number of children, count of cases | `integer` |
| **Categorical** | Nominal | Sex, blood type, cancer type | `factor` |
| **Categorical** | Ordinal | Education level, pain scale, severity | `ordered factor` |

```{r variable-types}
# Quantitative continuous
age   <- c(45, 50, 55, 40, 60, 70, 45, 50, 65, 55)
class(age)

# Categorical nominal — must convert to factor for statistics
sex   <- c("Male", "Female", "Male", "Female", "Male")
sex_f <- factor(sex)
class(sex_f)
levels(sex_f)

# Categorical ordinal — ordered factor
severity <- factor(c("Low", "High", "Medium", "Low", "High"),
                   levels  = c("Low", "Medium", "High"),
                   ordered = TRUE)
severity
severity[1] < severity[3]   # TRUE — ordering works
```

> ⚠️ Always convert character columns to `factor` before doing statistical summaries. R cannot produce meaningful frequency counts or group comparisons on raw character columns.

---

# Building and Exploring a Dataset

## Creating a Dataset for This Lecture

```{r create-data}
set.seed(7)

s      <- c("Female", "Male")
sexe   <- sample(s, 423, replace = TRUE)
a      <- c(15:20)
age    <- sample(a, 423, replace = TRUE)
cl     <- c("A", "B", "C")
classe <- sample(cl, 423, replace = TRUE)

anglais  <- round(rnorm(423, 7, 0.5) + ifelse(sexe == "Female", 2, 0), 0)
info     <- round(rnorm(423, 5, 3)   + ifelse(age >= 19, 2, 0), 0)
art      <- round(rnorm(423, 4, 3)   + ifelse(classe == "C", 2, -1), 0)
histoire <- round(rnorm(423, 6, 2)   + ifelse(classe == "B" & sexe == "Male", 2, 0), 0)

ecole <- data.frame(age, sexe, classe, anglais, info, art, histoire)
rm(age, classe, sexe, anglais, info, art, histoire, a, cl, s)

# Clip scores to [0, 10]
for (v in c("anglais", "info", "art", "histoire")) {
  ecole[[v]] <- pmin(pmax(ecole[[v]], 0), 10)
}

# Convert categorical variables to factors
ecole$sexe   <- as.factor(ecole$sexe)
ecole$classe <- as.factor(ecole$classe)

head(ecole)
```

## First-Inspection Functions

```{r first-inspection}
dim(ecole)         # Rows × Columns
nrow(ecole)        # Number of observations
ncol(ecole)        # Number of variables
names(ecole)       # Variable names
str(ecole)         # Structure: type + preview of each column
```

## `summary()` — The Swiss-Army Knife

`summary()` gives six key statistics for quantitative variables, and frequency counts for factors:

```{r summary}
summary(ecole)
```

For **quantitative** columns: Min, Q1, Median, Mean, Q3, Max.  
For **factor** columns: count of each level.  
For **columns with NAs**: the number of missing values is also shown.

> 💡 `summary()` is always the first function to call on a new dataset.

---

# Measures of Central Tendency

Central tendency tells us *where the data is centered* — the typical or representative value.

## Mean — `mean()`

The arithmetic mean is the sum of all values divided by the count. It is sensitive to **outliers**.

$$\bar{x} = \frac{\sum x_i}{n}$$

```{r mean}
age_patients <- c(45, 50, 55, 40, 60, 70, 45, 50, 65, 55)
mean(age_patients)

# With mtcars
mean(mtcars$mpg)

# Missing values: without na.rm the result is NA
x <- c(3, 5, NA, 8, NA, 2)
mean(x)                  # NA
mean(x, na.rm = TRUE)    # Correct: ignores NAs
```

## Median — `median()`

The middle value when data is sorted. It is **robust to outliers** — a better measure of center for skewed data.

```{r median}
blood_pressure <- c(130, 140, 120, 135, 150, 125, 145, 155, 140, 130)
median(blood_pressure)

median(mtcars$mpg)

# Compare mean vs median on skewed data
income <- c(30, 32, 35, 38, 40, 42, 50, 200)   # outlier: 200k
mean(income)     # pulled up by the outlier
median(income)   # robust — closer to the typical value
```

> 💡 **When to use which?** For symmetric distributions, mean and median are close. For skewed data (income, property prices, hospital stay lengths), prefer the **median**.

## Mode

R has no built-in `mode()` function for statistical mode. Write a simple custom function:

```{r mode}
get_mode <- function(v) {
  uniq_vals <- unique(v)
  uniq_vals[which.max(tabulate(match(v, uniq_vals)))]
}

cancer_types <- c("Lung", "Breast", "Breast", "Lung", "Lung", "Prostate", "Breast")
get_mode(cancer_types)   # "Lung" — appears most frequently

# Mode is especially useful for categorical variables
get_mode(ecole$classe)
```

## Mean, Median, Mode — Summary

| Measure | Function | Sensitive to outliers? | Best for |
|---|---|---|---|
| Mean | `mean()` | Yes | Symmetric distributions |
| Median | `median()` | No | Skewed data, ordinal data |
| Mode | custom `get_mode()` | No | Categorical data, nominal |

---

# Measures of Dispersion

Dispersion tells us *how spread out* the values are around the center.

## Range — `range()` and `diff()`

The range spans from the minimum to the maximum value:

```{r range}
weights <- c(60, 70, 65, 80, 72, 55, 78, 90)

min(weights)
max(weights)
range(weights)              # Returns both min and max

diff(range(weights))        # Actual range width = max - min = 35

# Find the row (observation) with max/min
which.max(mtcars$hp)
which.min(mtcars$hp)

# Get the car name
rownames(mtcars)[which.max(mtcars$hp)]   # "Maserati Bora"
rownames(mtcars)[which.min(mtcars$hp)]   # "Honda Civic"
```

## Variance — `var()`

The variance measures the average squared deviation from the mean. Larger variance = more spread:

$$s^2 = \frac{\sum (x_i - \bar{x})^2}{n - 1}$$

```{r variance}
cholesterol <- c(210, 230, 220, 240, 260, 250)
var(cholesterol)

var(mtcars$mpg)
var(ecole$anglais)
```

## Standard Deviation — `sd()`

The square root of the variance, in the same units as the original data. Much more interpretable than variance:

$$s = \sqrt{s^2}$$

```{r sd}
sd(cholesterol)   # ~18.7 mg/dL — typical deviation from the mean
sd(mtcars$mpg)
sd(ecole$histoire, na.rm = TRUE)
```

> 💡 In a **normal distribution**, approximately 68% of values fall within ±1 SD of the mean, 95% within ±2 SD, and 99.7% within ±3 SD. This is the **68-95-99.7 rule**.

## Coefficient of Variation

Useful to compare variability between variables with different units or scales:

```{r cv}
cv <- function(x, na.rm = TRUE) sd(x, na.rm = na.rm) / mean(x, na.rm = na.rm) * 100

cv(mtcars$mpg)    # % relative variability in mpg
cv(mtcars$hp)     # % relative variability in hp
```

---

# Measures of Position

## Quartiles and Percentiles — `quantile()`

Percentiles divide the dataset into 100 equal parts. The key percentiles are the **quartiles**:

- **Q1 (25th percentile):** 25% of values fall below
- **Q2 (50th percentile):** the median
- **Q3 (75th percentile):** 75% of values fall below

```{r quantiles}
blood_sugar <- c(100, 120, 110, 125, 130, 140, 135, 150, 145, 155)

quantile(blood_sugar)                        # Default: 0%, 25%, 50%, 75%, 100%
quantile(blood_sugar, probs = 0.25)          # Just Q1
quantile(blood_sugar, probs = c(0.1, 0.9))  # 10th and 90th percentile

# For crime rates
quantile(mtcars$mpg, probs = 0.025)   # 2.5th percentile
quantile(mtcars$mpg, probs = 0.975)   # 97.5th percentile
```

## Interquartile Range (IQR) — `IQR()`

The IQR is the range of the middle 50% of the data: **IQR = Q3 − Q1**. It is robust to outliers and used in boxplot construction:

```{r iqr}
IQR(blood_sugar)
IQR(mtcars$mpg)
IQR(ecole$age)
```

## Confidence Intervals for the Mean

A 95% confidence interval gives a range of plausible values for the true population mean:

$$CI = \bar{x} \pm 1.96 \times \frac{s}{\sqrt{n}}$$

```{r confidence-interval}
mpg     <- mtcars$mpg
n       <- length(mpg)
m       <- mean(mpg)
s       <- sd(mpg)

upper_ci <- m + 1.96 * s / sqrt(n)
lower_ci <- m - 1.96 * s / sqrt(n)

cat("Mean:", round(m, 2), "\n")
cat("95% CI: [", round(lower_ci, 2), ",", round(upper_ci, 2), "]\n")

# Or use t.test() directly:
t.test(mpg)$conf.int
```

---

# The `summary()` Function in Depth

`summary()` is the fastest way to get all central tendency and dispersion measures at once:

```{r summary-depth}
summary(mtcars$mpg)    # On a single variable
summary(ecole)         # On the whole data frame

# On a factor — equivalent to table()
summary(ecole$sexe)
table(ecole$sexe)      # Same result
```

## Applying Statistics to Subsets

You can compute statistics on filtered subsets using logical indexing:

```{r subset-stats}
# Mean score of the first 10 students
mean(ecole$anglais[1:10])

# Mean info score for students who scored above average
mean(ecole$info[ecole$info > mean(ecole$info)])

# Mean histoire score by sex
mean(ecole$histoire[ecole$sexe == "Male"])
mean(ecole$histoire[ecole$sexe == "Female"])

# Mean anglais for class C students aged 17
subset_data <- ecole$anglais[ecole$classe == "C" & ecole$age == 17]
mean(subset_data)
length(subset_data)   # Always check the sample size!
```

---

# Applying Functions to All Columns: `sapply()`

```{r sapply-stats}
# Mean of every numeric column
sapply(ecole[, c("age", "anglais", "info", "art", "histoire")], mean)

# Multiple statistics at once
sapply(ecole[, c("anglais", "info", "art", "histoire")],
       function(x) c(mean   = round(mean(x), 2),
                     sd     = round(sd(x), 2),
                     median = median(x),
                     min    = min(x),
                     max    = max(x)))
```

---

# Frequency Tables

Frequency tables are the primary tool for summarizing **categorical** variables.

## One-Way Table — `table()`

```{r table-one}
table(ecole$sexe)
table(ecole$classe)

# Single categorical column from any vector
cancer_types <- c("Lung", "Breast", "Breast", "Lung", "Lung", "Prostate", "Breast")
table(cancer_types)
```

## `prop.table()` — Relative Frequencies (Proportions)

```{r prop-table}
t1 <- table(ecole$sexe, ecole$classe)

prop.table(t1)       # Cell proportions (out of grand total)
prop.table(t1, 1)    # Row proportions  (each row sums to 1)
prop.table(t1, 2)    # Column proportions (each column sums to 1)

# As percentages — round for readability
round(100 * prop.table(t1, 1), 1)   # Row %
round(100 * prop.table(t1, 2), 1)   # Column %
```

## `addmargins()` — Adding Row and Column Totals

```{r addmargins}
addmargins(t1)                                    # Raw counts with totals
addmargins(round(100 * prop.table(t1), 0))        # % with totals
```

## Handling `NA` in Tables

```{r table-na}
data_with_na <- c("A", "B", "A", NA, "C", "B", "A", NA)

table(data_with_na)                          # NAs excluded (default)
table(data_with_na, useNA = "ifany")         # Include NA if present
table(data_with_na, useNA = "always")        # Always show NA column
```

## Two-Way Cross-Tabulation

```{r crosstab}
table(ecole$sexe, ecole$classe)

# Logical conditions in table()
table(mtcars$mpg > 20, mtcars$cyl > 4)
```

## Three-Way Table

```{r three-way}
# ftable() gives a compact flat display for 3+ variables
ftable(ecole$sexe, ecole$classe, ecole$age)
```

## `addmargins()` + `prop.table()` Combined

```{r combined-table}
t1 <- table(ecole$sexe, ecole$classe)
addmargins(round(100 * prop.table(t1, 1), 1))
```

---

# Grouped Statistics

## `tapply()` — Apply a Function by Groups

```{r tapply}
# Mean of a variable broken down by a grouping factor
tapply(ecole$anglais, ecole$sexe,   mean)
tapply(ecole$anglais, ecole$classe, mean)

# By two grouping factors simultaneously
tapply(ecole$histoire,
       list(ecole$sexe, ecole$classe),
       mean)
```

## `aggregate()` — Group Summaries as a Data Frame

```{r aggregate}
aggregate(mpg ~ cyl, data = mtcars, FUN = mean)

# Multiple statistics
aggregate(cbind(anglais, info, art) ~ sexe, data = ecole, FUN = mean)
aggregate(cbind(anglais, info, art) ~ classe + sexe, data = ecole, FUN = mean)
```

## `dplyr` Approach — `group_by()` + `summarise()`

```{r dplyr-grouped}
ecole %>%
  group_by(sexe, classe) %>%
  summarise(
    n          = n(),
    mean_ang   = round(mean(anglais), 2),
    mean_info  = round(mean(info), 2),
    mean_hist  = round(mean(histoire), 2),
    .groups    = "drop"
  ) %>%
  arrange(sexe, classe)
```

---

# Missing Values — Detection and Handling

## Detecting Missing Values

```{r missing-detect}
# Add some NAs for demonstration
x <- c(3, 5, NA, 8, NA, 2, 9, NA)

is.na(x)               # Logical vector: TRUE where NA
sum(is.na(x))          # How many NAs?
mean(is.na(x)) * 100   # What % are NA?

# For a data frame
colSums(is.na(ecole))  # NAs per column
```

## Ignoring NAs in Calculations

```{r na-rm}
mean(x)                 # NA — cannot compute with missing values
mean(x, na.rm = TRUE)   # Correct
median(x, na.rm = TRUE)
sd(x, na.rm = TRUE)
quantile(x, na.rm = TRUE)
```

## Replacing NAs

```{r na-replace}
# Replace NA with 0
x_filled <- ifelse(is.na(x), 0, x)
x_filled

# Replace NA with the mean
x_imputed <- ifelse(is.na(x), mean(x, na.rm = TRUE), x)
x_imputed
```

## Removing Rows with NAs — `na.omit()`

```{r na-omit}
df_clean <- na.omit(ecole)   # Removes any row containing at least one NA
nrow(ecole)
nrow(df_clean)
```

---

# Visual Summaries

Visualizations are an essential part of descriptive statistics. They reveal patterns, skewness, and outliers that summary numbers alone can miss.

## Histogram — Distribution of One Variable

```{r histogram}
hist(mtcars$mpg,
     col  = "steelblue",
     main = "Distribution of Fuel Economy",
     xlab = "Miles per Gallon",
     ylab = "Frequency",
     breaks = 10)
```

## Boxplot — Five-Number Summary

```{r boxplot}
boxplot(mpg ~ cyl, data = mtcars,
        col  = c("#6baed6", "#fd8d3c", "#74c476"),
        main = "Fuel Economy by Number of Cylinders",
        xlab = "Cylinders",
        ylab = "Miles per Gallon")
```

## Density Plot with `ggplot2`

```{r density}
ggplot(ecole, aes(x = anglais, fill = sexe)) +
  geom_density(alpha = 0.4) +
  labs(title = "English Score Distribution by Sex",
       x = "English Score", fill = "Sex") +
  theme_bw()
```

## Histogram with Normal Curve Overlay

```{r hist-normal}
x <- mtcars$mpg

hist(x, probability = TRUE, col = "lightblue",
     main = "MPG with Normal Curve",
     xlab = "Miles per Gallon")
curve(dnorm(x, mean = mean(x), sd = sd(x)),
      add = TRUE, col = "red", lwd = 2)
```

## QQ-Plot — Checking Normality

```{r qqplot}
qqnorm(mtcars$mpg, main = "QQ-Plot: MPG",
       pch = 19, col = "steelblue")
qqline(mtcars$mpg, col = "red", lwd = 2)
```

Points following the red line indicate a normal distribution.

---

# Advanced Descriptive Packages

## `psych::describe()` — Rich Variable Summary

```{r psych-describe}
library(psych)

psych::describe(mtcars[, c("mpg", "hp", "wt")])
```

Output includes: n, mean, sd, median, trimmed mean, mad (median absolute deviation), min, max, range, **skewness**, **kurtosis**, and standard error.

## `Hmisc::describe()` — Detailed with Frequencies

```{r hmisc-describe}
library(Hmisc)

Hmisc::describe(ecole[, c("anglais", "info", "sexe")])
```

For numeric columns: n, missing, distinct values, mean, and quantiles.  
For categorical columns: frequency and proportion of each level.

## `psych::describeBy()` — Summary by Group

```{r describe-by}
psych::describeBy(ecole[, c("anglais", "info", "art")],
                  group = ecole$sexe)
```

---

# Ordinal Variables

Ordinal variables have ordered categories. Create them with `ordered = TRUE`:

```{r ordinal}
ecole$moyenne <- rowMeans(ecole[, c("anglais", "info", "art", "histoire")])

ecole$niveau <- cut(ecole$moyenne,
                    breaks = c(-Inf, 6, 7, Inf),
                    labels = c("Average", "Good", "Excellent"))

ecole$niveau <- ordered(ecole$niveau,
                        levels = c("Average", "Good", "Excellent"))

table(ecole$niveau)
summary(ecole$niveau)

# Ordering enables meaningful comparisons
ecole$niveau[1] < ecole$niveau[2]   # Meaningful because ordered
```

---

# Probability Distributions

R provides four functions for every distribution, following a consistent naming pattern:

| Prefix | Meaning | Example |
|---|---|---|
| `d*()` | Density / probability mass at a point | `dnorm(x)` |
| `p*()` | Cumulative probability P(X ≤ x) | `pnorm(x)` |
| `q*()` | Quantile: value at given probability | `qnorm(p)` |
| `r*()` | Random sample from the distribution | `rnorm(n)` |

## Normal Distribution

The most important distribution in statistics — bell-shaped, symmetric.

```{r normal-dist}
# Generate random data: daily malaria cases (mean=120, sd=30)
set.seed(123)
daily_cases <- rnorm(10, mean = 120, sd = 30)
daily_cases

# Probability density at x = 150
dnorm(150, mean = 120, sd = 30)

# Cumulative probability: P(X ≤ 150)
pnorm(150, mean = 120, sd = 30)

# Probability between 100 and 150
pnorm(150, 120, 30) - pnorm(100, 120, 30)

# 97.5th percentile
qnorm(0.975, mean = 120, sd = 30)

# Plot the standard normal curve
x <- seq(-3, 3, length.out = 200)
plot(x, dnorm(x), type = "l", lwd = 2, col = "steelblue",
     xlab = "Z-score", ylab = "Density",
     main = "Standard Normal Distribution")
abline(v = c(-1.96, 1.96), col = "red", lty = 2)
```

## Binomial Distribution

Models the number of successes in `n` independent yes/no trials:

```{r binomial-dist}
# Probability of exactly 20 positive RDT tests out of 100 (p = 0.30)
dbinom(20, size = 100, prob = 0.30)

# Cumulative: P(X ≤ 20)
pbinom(20, size = 100, prob = 0.30)

# Generate 10 random counts
rbinom(10, size = 100, prob = 0.30)
```

## Poisson Distribution

Models the number of events in a fixed interval when events are rare and independent:

```{r poisson-dist}
# 20 random daily malaria counts (mean λ = 80)
rpois(20, lambda = 80)

# P(exactly 75 cases in a day)
dpois(75, lambda = 80)

# P(≤ 75 cases)
ppois(75, lambda = 80)

# Value below which 90% of days fall
qpois(0.90, lambda = 80)
```

## Uniform Distribution

All values in a range are equally likely:

```{r uniform-dist}
# Simulate 10 random rainfall values between 0 and 250 mm
runif(10, min = 0, max = 250)

# Density at x = 100 (constant for uniform)
dunif(100, min = 0, max = 250)

# P(rainfall < 150 mm)
punif(150, min = 0, max = 250)

# 90th percentile
qunif(0.90, min = 0, max = 250)
```

## Reproducible Random Numbers — `set.seed()`

```{r set-seed}
set.seed(42)
rnorm(5)   # Always produces the same 5 values

set.seed(42)
rnorm(5)   # Identical to above
```

## `sample()` — Random Sampling

```{r sample-func}
# 10 random patients from IDs 1 to 1000
sample(1:1000, size = 10, replace = FALSE)

# Simulate diagnostic outcomes with probabilities
sample(c("Positive", "Negative"), size = 10,
       replace = TRUE, prob = c(0.35, 0.65))
```

---

# Complete Worked Example

```{r complete-example}
library(tidyverse)

# Full descriptive analysis pipeline on mtcars
mtcars %>%
  # 1. Select variables of interest
  select(mpg, hp, wt, cyl) %>%

  # 2. Overall summary
  summary() %>%
  print()

# 3. Summary statistics by group
mtcars %>%
  mutate(cyl_f = factor(cyl)) %>%
  group_by(cyl_f) %>%
  summarise(
    n         = n(),
    mean_mpg  = round(mean(mpg), 1),
    sd_mpg    = round(sd(mpg), 1),
    median_hp = median(hp),
    IQR_wt    = IQR(wt),
    .groups   = "drop"
  )
```

---

# Statistical Summary Functions — Quick Reference

| Task | Function | Package |
|---|---|---|
| Mean | `mean(x, na.rm=TRUE)` | base |
| Median | `median(x, na.rm=TRUE)` | base |
| Mode | custom `get_mode(x)` | — |
| Variance | `var(x, na.rm=TRUE)` | base |
| Std deviation | `sd(x, na.rm=TRUE)` | base |
| Min / Max | `min(x)` / `max(x)` | base |
| Range | `range(x)` / `diff(range(x))` | base |
| Quantiles | `quantile(x, probs=...)` | base |
| IQR | `IQR(x, na.rm=TRUE)` | base |
| Full 6-number summary | `summary(x)` | base |
| Frequency table | `table(x)` | base |
| Proportions | `prop.table(t)` | base |
| With margins | `addmargins(t)` | base |
| By group | `tapply(x, g, FUN)` | base |
| By group (df) | `aggregate(y ~ g, df, FUN)` | base |
| Rich summary | `psych::describe(df)` | psych |
| Detailed summary | `Hmisc::describe(df)` | Hmisc |
| Summary by group | `psych::describeBy(df, g)` | psych |

---

# Practice Exercises

**Exercise 1.** Using the `airquality` dataset, compute the mean, median, standard deviation, and IQR for `Ozone`, `Solar.R`, `Wind`, and `Temp`. Handle missing values appropriately. Which variable has the highest relative variability (CV)?

**Exercise 2.** Using the `mtcars` dataset, create a frequency table of `cyl`. Then create a cross-tabulation of `cyl` vs `am` (transmission type). Compute row percentages and add margins. Interpret the result.

**Exercise 3.** Using the `ecole` dataset created in this lecture, compute the mean score in `anglais` separately for: (a) each sex, (b) each class, (c) each sex × class combination. Which group has the highest mean?

**Exercise 4.** Generate 1000 random values from a Normal distribution with mean = 70 and sd = 10. Compute the mean, median, sd, and 5th/95th percentiles. What proportion of values fall within ±1 sd of the mean? Plot a histogram with the theoretical normal curve overlaid.

**Exercise 5.** Write a function `full_describe(x, group = NULL)` that prints: mean, median, mode, SD, IQR, min, max, number of NAs, and (if `group` is provided) all of these split by group. Test it on `mtcars$mpg` with `group = mtcars$cyl`.

**Exercise 6.** Using the Poisson distribution with λ = 50, compute: (a) the probability of exactly 45 events, (b) the probability of more than 60 events, (c) the 95th percentile. Then generate 500 Poisson samples and verify that their mean and variance are approximately equal to λ.

---

```{r cleanup, include=FALSE}
rm(list = ls())
```

---

*End of Lecture — Descriptive Statistics in R*
