---
title: "Conditional Statements and Control Flow in R"
subtitle: "if/else, ifelse(), case_when(), while loops, for loops, and functions"
author: "Ange Maxime TCHOUTANG"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    theme: flatly
    highlight: tango
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  comment = "##"
)
```

---

# Introduction

**Control flow** is the order in which a program executes its instructions. By default, R runs code line by line from top to bottom. Conditional statements and loops let you change that order — executing certain blocks only when needed, or repeating them automatically.

In this lecture we cover:

- **Comparison and logical operators** — the building blocks of all conditions
- **`if` / `else if` / `else`** — run code based on a condition
- **Nested `if`** — conditions inside conditions
- **`ifelse()`** — vectorized conditional for data manipulation
- **`case_when()`** — the clean, scalable alternative to nested `ifelse()`
- **`switch()`** — select from multiple named cases
- **`while` loops** — repeat while a condition is true
- **`for` loops** — iterate over a sequence or collection
- **`break` and `next`** — control loop execution
- **Nested loops** — loops inside loops
- **Functions** — reusable blocks of code

---

# Comparison and Logical Operators

Every conditional statement relies on a **Boolean** result: either `TRUE` or `FALSE`. Comparison operators produce these results.

## Comparison Operators

| Operator | Meaning | Example | Result |
|---|---|---|---|
| `==` | Equal to | `5 == 5` | `TRUE` |
| `!=` | Not equal to | `5 != 3` | `TRUE` |
| `>` | Greater than | `10 > 3` | `TRUE` |
| `<` | Less than | `2 < 1` | `FALSE` |
| `>=` | Greater than or equal to | `5 >= 5` | `TRUE` |
| `<=` | Less than or equal to | `4 <= 3` | `FALSE` |

```{r comparison-operators}
x <- 10
y <- 20

x == y    # Is x equal to y?
x != y    # Is x different from y?
x > y     # Is x greater than y?
x < y     # Is x less than y?
x >= 10   # Is x greater than or equal to 10?
y <= 20   # Is y less than or equal to 20?
```

> ⚠️ **Common mistake:** Do not confuse `=` (assignment) with `==` (equality test). Always use `==` when comparing values inside conditions.

## Logical Operators

Logical operators combine multiple conditions.

| Operator | Meaning | Example |
|---|---|---|
| `&` | AND (element-wise) | `x > 5 & x < 20` |
| `\|` | OR (element-wise) | `x < 5 \| x > 8` |
| `!` | NOT (negation) | `!(x == 10)` |
| `&&` | AND (scalar only) | Use inside `if()` |
| `\|\|` | OR (scalar only) | Use inside `if()` |

```{r logical-operators}
a <- 200
b <- 33
c <- 500

# AND: both conditions must be TRUE
a > b & c > a

# OR: at least one condition must be TRUE
a > b | a > c

# NOT: inverts the result
!(a == b)
```

> 💡 Use `&` and `|` when working with **vectors** (they test element by element). Use `&&` and `||` only inside `if()` statements with **single scalar values** — they are short-circuit evaluated (stop as soon as the result is known).

---

# The `if` Statement

An `if` statement executes a block of code **only when its condition is `TRUE`**.

```r
if (condition) {
  # code to run if condition is TRUE
}
```

```{r if-basic}
b <- 200
a <- 33

if (b > a) {
  print("b is greater than a")
}
```

R uses **curly braces `{ }`** to define the block of code belonging to each condition. Everything inside the braces is executed together.

---

# `if` / `else if` / `else`

The full conditional structure handles multiple possible outcomes:

```r
if (condition1) {
  # runs if condition1 is TRUE
} else if (condition2) {
  # runs if condition1 was FALSE and condition2 is TRUE
} else {
  # runs if ALL previous conditions were FALSE
}
```

R evaluates conditions **from top to bottom** and executes the **first** block whose condition is `TRUE`. Once a match is found, all remaining branches are skipped.

```{r if-else-if-else}
a <- 200
b <- 33

if (b > a) {
  print("b is greater than a")
} else if (a == b) {
  print("a and b are equal")
} else {
  print("a is greater than b")
}
```

## Full Example — Age Classification

```{r age-classification}
age <- 25

if (age < 0) {
  print("Not yet born!")
} else if (age < 18) {
  print("Minor")
} else if (age < 65) {
  print("Adult")
} else if (age <= 120) {
  print("Senior")
} else {
  print("Immortal!")
}
```

## Combining Conditions with `&` and `|`

```{r combined-conditions}
x <- 100
y <- 100

# Both conditions in one if
if (x > 0 & y > 0) {
  print("Both x and y are positive")
}

# Using OR
score <- 85
if (score >= 90 | score < 0) {
  print("Score is either excellent or invalid")
} else {
  print("Score is in the normal range")
}
```

---

# Nested `if` Statements

You can place `if` statements **inside** other `if` statements. This is called a **nested if** and is used when the inner decision only makes sense if the outer condition is already true.

```{r nested-if}
num <- 50

if (num > 20) {
  print("Greater than 20")
  if (num > 30) {
    print("Also greater than 30")
  } else {
    print("But NOT greater than 30")
  }
} else {
  print("Not above 20")
}
```

```{r nested-if-2}
x <- 41

if (x > 10) {
  print("Above 10")
  if (x > 20) {
    print("And also above 20!")
  } else {
    print("But not above 20.")
  }
} else {
  print("Below or equal to 10")
}
```

> ⚠️ **Nesting caution:** More than 2–3 levels of nested `if` statements become very hard to read and debug. When you find yourself writing deeply nested conditions, consider using `ifelse()` or `case_when()` instead (see below).

---

# Vectorized Conditions: `ifelse()`

The standard `if / else` structure works on a **single value**. When you need to apply a condition to every element of a vector or data frame column, use `ifelse()`.

```r
ifelse(condition, value_if_TRUE, value_if_FALSE)
```

```{r ifelse-basic}
ages <- c(12, 25, 17, 34, 8, 19)

# Classify each age as "Adult" or "Minor"
ifelse(ages >= 18, "Adult", "Minor")
```

```{r ifelse-vector}
scores <- c(45, 82, 67, 91, 55, 73)

# Pass/Fail
ifelse(scores >= 60, "Pass", "Fail")
```

## `ifelse()` with Data Frames

`ifelse()` integrates naturally with `dplyr::mutate()`:

```{r ifelse-mutate, message=FALSE}
library(tidyverse)

starwars %>%
  select(name, height) %>%
  mutate(size = ifelse(height > 180, "Tall", "Short")) %>%
  drop_na() %>%
  head(8)
```

## Handling `NA` values with `ifelse()`

```{r ifelse-na}
x <- c(5, NA, 15, NA, 25)

# Replace NA with 0, keep others unchanged
ifelse(is.na(x), 0, x)
```

> ⚠️ `ifelse()` is powerful but **nested `ifelse()` calls become unreadable** very quickly. For more than two outcomes, use `case_when()`.

---

# `case_when()` — The Scalable Alternative

`case_when()` from `dplyr` is the cleanest way to handle **multiple conditions**. Each condition is written as:

```r
condition ~ result
```

Conditions are evaluated top to bottom. The first `TRUE` match wins. `TRUE ~ ...` at the end acts as the default (equivalent to the final `else`).

```{r case-when-basic}
ages <- c(-2, 8, 16, 25, 67, 150)

case_when(
  ages < 0   ~ "Not yet born",
  ages < 18  ~ "Minor",
  ages < 65  ~ "Adult",
  ages <= 120 ~ "Senior",
  TRUE        ~ "Immortal"
)
```

## `case_when()` vs Nested `ifelse()`

```{r case-when-comparison}
scores <- c(95, 82, 67, 54, 40)

# Nested ifelse — hard to read
ifelse(scores >= 90, "A",
  ifelse(scores >= 80, "B",
    ifelse(scores >= 70, "C",
      ifelse(scores >= 60, "D", "F"))))

# case_when — clean and readable
case_when(
  scores >= 90 ~ "A",
  scores >= 80 ~ "B",
  scores >= 70 ~ "C",
  scores >= 60 ~ "D",
  TRUE         ~ "F"
)
```

## `case_when()` in a Data Frame

```{r case-when-mutate}
starwars %>%
  select(name, height) %>%
  drop_na() %>%
  mutate(category = case_when(
    height > 200 ~ "Very tall",
    height > 180 ~ "Tall",
    height > 150 ~ "Average",
    TRUE         ~ "Short"
  )) %>%
  head(10)
```

## `case_when()` with Missing Value Logic

`case_when()` is especially powerful when handling missing values that have different meanings depending on context:

```{r case-when-na}
driving <- c(0.02, NA, 0.01, NA, 0.03)
other   <- c(0.01, 0.02, NA, NA, 0.01)

case_when(
  is.na(driving) & is.na(other)  ~ 0,
  is.na(driving)                 ~ other,
  is.na(other)                   ~ driving,
  TRUE                           ~ driving + other
)
```

---

# `switch()` — Selecting from Named Cases

`switch()` is ideal when you have a variable that can take several specific string values and you want to execute a different action for each.

```{r switch-basic}
day_type <- function(day) {
  switch(day,
    "Monday"    = ,
    "Tuesday"   = ,
    "Wednesday" = ,
    "Thursday"  = ,
    "Friday"    = "Weekday",
    "Saturday"  = ,
    "Sunday"    = "Weekend",
    "Unknown day"   # default
  )
}

day_type("Monday")
day_type("Saturday")
day_type("Wednesday")
day_type("Holiday")
```

> 💡 Leaving a case value empty (e.g., `"Monday" = ,`) means it **falls through** to the next non-empty case — useful for grouping multiple values to the same result.

---

# The `while` Loop

A `while` loop repeats a block of code **as long as its condition remains `TRUE`**. It checks the condition **before** each iteration.

```r
while (condition) {
  # code to repeat
  # must include something that will eventually make condition FALSE
}
```

> ⚠️ **Always increment your counter!** If the condition never becomes `FALSE`, the loop runs forever (infinite loop). Always make sure something inside the loop changes the variable being tested.

```{r while-basic}
i <- 1

while (i < 6) {
  print(i)
  i <- i + 1   # increment — essential to avoid infinite loop
}
```

## Practical `while` Example — Score Countdown

```{r while-score}
score <- 100

while (score > 0) {
  print(paste("You are still alive. Score:", score))
  score <- score - 10
}
```

## `while` with `if` inside — Adding Milestones

```{r while-milestone}
score <- 100

while (score > 0) {
  if (score == 50) {
    print(">>> You reached the halfway point!")
  }
  print(paste("Score:", score))
  score <- score - 10
}
```

## `break` — Stop the Loop Early

`break` immediately exits the loop, regardless of the condition.

```{r while-break}
x <- 1

while (x <= 10) {
  print(x)
  x <- x + 1
  if (x == 5) {
    break   # stop as soon as x reaches 5
  }
}
```

## `next` — Skip One Iteration

`next` skips the rest of the current iteration and jumps straight to the next one.

```{r while-next}
i <- 0

while (i < 6) {
  i <- i + 1
  if (i == 3) {
    next   # skip printing when i equals 3
  }
  print(i)
}
```

## `while` + `if/else` — Yahtzee Example

```{r while-yahtzee}
dice <- 1

while (dice <= 6) {
  if (dice < 6) {
    print(paste("Dice:", dice, "— No Yahtzee"))
  } else {
    print(paste("Dice:", dice, "— YAHTZEE!"))
  }
  dice <- dice + 1
}
```

---

# The `for` Loop

A `for` loop iterates over every element of a sequence — a numeric range, a vector, a list, or any iterable object. Unlike `while`, it does not need an explicit counter.

```r
for (variable in sequence) {
  # code executed once per element
}
```

## Iterating over a Numeric Range

```{r for-range}
for (i in 1:10) {
  print(i)
}
```

## Iterating over a Vector

```{r for-vector}
fruits <- c("apple", "banana", "cherry", "mango")

for (fruit in fruits) {
  print(fruit)
}
```

## Iterating over a List

```{r for-list}
info <- list("Alice", 25, TRUE)

for (item in info) {
  print(item)
}
```

## Using `length()` to Loop over a Data Structure

```{r for-length}
numbers <- c(10, 20, 30, 40, 50)

for (i in 1:length(numbers)) {
  print(paste("Position", i, "— Value:", numbers[i]))
}
```

## `for` with `if` inside — Filtering During Iteration

```{r for-if}
vec <- islands   # built-in R dataset: land masses and their areas

for (i in 1:length(vec)) {
  if (vec[i] > 10000) {
    print(paste(names(vec[i]), "is a very large land mass:", vec[i], "thousand sq mi"))
  }
}
```

## `break` in a `for` Loop

```{r for-break}
fruits <- list("apple", "banana", "cherry")

for (x in fruits) {
  if (x == "cherry") {
    break   # stop when we reach cherry
  }
  print(x)
}
```

## `next` in a `for` Loop

```{r for-next}
numbers <- c(10, 20, 30, 40, 50)

for (num in numbers) {
  if (num == 30) {
    next   # skip 30
  }
  print(num)
}
```

## `for` + `if/else` — Yahtzee Revisited

```{r for-yahtzee}
dice <- 1:6

for (x in dice) {
  if (x == 6) {
    print(paste("Dice:", x, "— YAHTZEE!"))
  } else {
    print(paste("Dice:", x, "— Not Yahtzee"))
  }
}
```

## `paste()` — Combining Text and Variables

Inside loops, `paste()` (or `paste0()`) lets you embed variable values into a printed message:

```{r paste-demo}
name  <- "Africa"
area  <- 11506

paste("The region", name, "has an area of", area, "thousand sq mi")
paste0("Region: ", name, " (", area, ")")
```

---

# Nested Loops

A **nested loop** is a loop inside another loop. The inner loop completes all its iterations for **each single iteration** of the outer loop.

```{r nested-loop}
adj    <- list("red", "big", "tasty")
fruits <- list("apple", "banana", "cherry")

for (x in adj) {
  for (y in fruits) {
    print(paste(x, y))
  }
}
```

## Nested Loop with a Condition

```{r nested-loop-condition}
for (i in 1:4) {
  for (j in 1:4) {
    if (i == j) {
      print(paste("Diagonal element:", i, ",", j))
    }
  }
}
```

> ⚠️ Nested loops can become slow with large data. For most data analysis tasks, vectorized operations (`ifelse()`, `case_when()`, `apply()`, or `purrr::map()`) are faster and more readable. Use nested loops when the logic genuinely requires sequential iteration.

---

# `break` and `next` — Summary

| Keyword | Works in | Effect |
|---|---|---|
| `break` | `while`, `for`, `repeat` | Immediately **exits** the loop entirely |
| `next` | `while`, `for` | **Skips** the current iteration and moves to the next |

```{r break-next-summary}
# break: exits at first multiple of 7
for (i in 1:50) {
  if (i %% 7 == 0) {
    print(paste("First multiple of 7:", i))
    break
  }
}

# next: prints only even numbers
for (i in 1:10) {
  if (i %% 2 != 0) next   # skip odd numbers
  print(i)
}
```

---

# Functions

Functions let you **package reusable logic** into a named unit. Once defined, you can call the function as many times as needed without duplicating code.

```r
my_function <- function(argument1, argument2) {
  # body: code that runs each time the function is called
  result <- argument1 + argument2
  return(result)   # return the output
}
```

## Defining and Calling a Function

```{r function-basic}
# Define
greet <- function(name) {
  print(paste("Hello,", name, "!"))
}

# Call
greet("Alice")
greet("Bob")
```

## Function with Multiple Arguments

```{r function-multiple-args}
power_mean <- function(x, power = 2) {
  mean(x, na.rm = TRUE) ^ power
}

power_mean(c(1:20, NA))          # uses default power = 2
power_mean(c(1:20, NA), power = 3)  # override default
```

> 💡 **Default arguments** let you define a sensible fallback value. Callers only need to supply the argument if they want a different value.

## Function with `if/else` inside

```{r function-if}
classify_score <- function(score) {
  if (score < 0 | score > 100) {
    return("Invalid score")
  } else if (score >= 90) {
    return("A")
  } else if (score >= 80) {
    return("B")
  } else if (score >= 70) {
    return("C")
  } else if (score >= 60) {
    return("D")
  } else {
    return("F")
  }
}

classify_score(95)
classify_score(73)
classify_score(42)
classify_score(-5)
```

## Function with a `for` Loop inside

```{r function-for}
print_multiples <- function(n, up_to = 10) {
  for (i in 1:up_to) {
    print(paste(n, "x", i, "=", n * i))
  }
}

print_multiples(3)
print_multiples(7, up_to = 5)
```

---

# Applying Functions over Vectors: `lapply()` and `sapply()`

For loops are one way to apply a function to each column or element. `lapply()` and `sapply()` are more concise alternatives.

```{r apply-functions}
# lapply: returns a list
lapply(airquality, mean, na.rm = TRUE)

# sapply: returns a named vector (simpler output)
sapply(airquality, mean, na.rm = TRUE)

# They are equivalent to this for loop:
for (i in 1:ncol(airquality)) {
  cat(names(airquality)[i], ":", mean(airquality[, i], na.rm = TRUE), "\n")
}
```

---

# Control Flow Decision Guide

Use this guide to choose the right tool:

| Situation | Best tool |
|---|---|
| Execute code only if one condition is met | `if` |
| Choose between 2 outcomes | `if / else` |
| Choose between 3+ outcomes | `if / else if / else` |
| Check a condition only if an outer condition was met | Nested `if` |
| Apply a condition to each element of a vector | `ifelse()` |
| Apply 3+ conditions to a vector or column | `case_when()` |
| Switch between named string cases | `switch()` |
| Repeat while a condition holds (unknown iterations) | `while` |
| Iterate over a fixed sequence or collection | `for` |
| Apply a function to every element of a list/vector | `lapply()` / `sapply()` |
| Exit a loop early | `break` |
| Skip one iteration | `next` |

---

# Complete Examples

## Example 1 — BMI Classifier

```{r example-bmi}
classify_bmi <- function(weight_kg, height_cm) {
  bmi <- weight_kg / (height_cm / 100)^2
  
  category <- case_when(
    bmi < 18.5              ~ "Underweight",
    bmi < 25.0              ~ "Normal weight",
    bmi < 30.0              ~ "Overweight",
    TRUE                    ~ "Obese"
  )
  
  paste0("BMI = ", round(bmi, 1), " — ", category)
}

classify_bmi(70, 175)
classify_bmi(50, 170)
classify_bmi(100, 165)
```

## Example 2 — FizzBuzz with a `for` Loop

```{r example-fizzbuzz}
for (i in 1:20) {
  if (i %% 15 == 0) {
    print("FizzBuzz")
  } else if (i %% 3 == 0) {
    print("Fizz")
  } else if (i %% 5 == 0) {
    print("Buzz")
  } else {
    print(i)
  }
}
```

## Example 3 — Countdown with `while`

```{r example-countdown}
countdown <- function(from) {
  n <- from
  while (n > 0) {
    print(paste(n, "..."))
    n <- n - 1
  }
  print("Lift off!")
}

countdown(5)
```

## Example 4 — Vectorized `case_when()` on a Data Frame

```{r example-starwars}
library(tidyverse)

starwars %>%
  select(name, height, mass) %>%
  drop_na() %>%
  mutate(
    bmi      = round(mass / (height / 100)^2, 1),
    category = case_when(
      bmi < 18.5 ~ "Underweight",
      bmi < 25.0 ~ "Normal",
      bmi < 30.0 ~ "Overweight",
      TRUE       ~ "Obese"
    )
  ) %>%
  count(category, sort = TRUE)
```

---

# Practice Exercises

**Exercise 1.** Write an `if / else if / else` block that takes a temperature (in °C) and prints: "Freezing" (< 0), "Cold" (0–15), "Warm" (16–25), or "Hot" (> 25). Test it with at least four values.

**Exercise 2.** Use `ifelse()` to classify a vector of 10 random integers between 1 and 100 as "Even" or "Odd". Then use `case_when()` to classify them as "Low" (1–33), "Medium" (34–66), or "High" (67–100).

**Exercise 3.** Write a `for` loop that iterates over the built-in `month.name` vector and prints each month name along with whether it has 28, 30, or 31 days. Use an `if / else if / else` inside the loop.

**Exercise 4.** Write a `while` loop that starts at 1 and prints the cumulative sum of integers. Stop the loop when the cumulative sum exceeds 100. Print which number caused it to stop.

**Exercise 5.** Write a function `grade_report(name, scores)` that takes a student's name and a numeric vector of scores, and prints: the student's name, their average score, and their letter grade (using `case_when()`). Call the function for at least three students.

**Exercise 6.** Using the `starwars` dataset and `case_when()`, create a new column `weight_class` based on the `mass` variable: "Light" (< 60 kg), "Medium" (60–100 kg), "Heavy" (> 100 kg), and "Unknown" (NA). Count how many characters fall into each class.

---

```{r cleanup, include=FALSE}
rm(list = ls())
```

---

*End of Lecture — Conditional Statements and Control Flow in R*
