---
title: "Data Visualization and Plotting in R"
subtitle: "A Comprehensive Lecture — From Base R Graphics to ggplot2 and Heatmaps"
author: "Ange Maxime TCHOUTANG"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    theme: flatly
    highlight: tango
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  comment   = "##",
  message   = FALSE,
  warning   = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 5
)
```

```{r packages}
library(tidyverse)
```

---

# Introduction

Data visualization is the art and science of turning numbers into pictures. A well-made plot communicates patterns, relationships, and outliers far more efficiently than any table of numbers. In life sciences, visualization is indispensable for summarizing experimental results, comparing gene expression across conditions, communicating epidemiological trends, and exploring the structure of complex datasets.

R has two major plotting systems:

| System | Package | Style | Best for |
|---|---|---|---|
| **Base R graphics** | built-in | Function calls | Quick, exploratory plots |
| **ggplot2** | tidyverse | Grammar of Graphics | Publication-quality figures |

In this lecture we cover both systems in depth, from the simplest point to complex multi-panel figures.

---

# Datasets Used in This Lecture

We will work with several datasets throughout this lecture:

```{r datasets}
# Built-in datasets
library(datasets)
data(trees)     # Girth, Height, Volume of 31 black cherry trees
data(mtcars)    # Motor Trend car data
data(iris)      # Iris flower measurements

# Simulated school dataset
set.seed(7)
sexe   <- sample(c("Female", "Male"), 423, replace = TRUE)
age    <- sample(15:20, 423, replace = TRUE)
classe <- sample(c("A", "B", "C"), 423, replace = TRUE)
anglais  <- round(rnorm(423, 7, 0.5) + ifelse(sexe == "Female", 2, 0), 0)
info     <- round(rnorm(423, 5, 3)   + ifelse(age >= 19, 2, 0), 0)
art      <- round(rnorm(423, 4, 3)   + ifelse(classe == "C", 2, -1), 0)
histoire <- round(rnorm(423, 6, 2)   + ifelse(classe == "B" & sexe == "Male", 2, 0), 0)
moyenne  <- (anglais + info + art + histoire) / 4
ecole    <- data.frame(age, sexe = factor(sexe),
                       classe = factor(classe),
                       anglais, info, art, histoire, moyenne)
for (v in c("anglais","info","art","histoire"))
  ecole[[v]] <- pmin(pmax(ecole[[v]], 0), 10)
rm(age, sexe, classe, anglais, info, art, histoire, moyenne)
```

---

# Base R Graphics

Base R graphics work through sequential function calls. You first call a **high-level function** (which creates a new plot), then optionally add elements with **low-level functions** (which add to an existing plot).

## Key Parameters Common to All Base Plots

Before diving into specific plot types, learn these universal parameters:

| Parameter | Meaning | Example values |
|---|---|---|
| `main` | Title | `"My Title"` |
| `sub` | Subtitle | `"My subtitle"` |
| `xlab` | X-axis label | `"Height (cm)"` |
| `ylab` | Y-axis label | `"Frequency"` |
| `col` | Color | `"red"`, `"#3498DB"`, `rainbow(5)` |
| `xlim` | X-axis range | `c(0, 100)` |
| `ylim` | Y-axis range | `c(0, 50)` |
| `cex` | Point/text size multiplier | `1.5` |
| `lwd` | Line width | `2` |
| `lty` | Line type | `1`=solid, `2`=dashed, `3`=dotted |
| `pch` | Point character/shape | `19`=filled circle, `17`=triangle |

---

## Scatter Plot — `plot()`

The most versatile base R function. With `type=` you control what is drawn.

### `type` Options

| `type` | Draws |
|---|---|
| `"p"` | Points only (default) |
| `"l"` | Lines only |
| `"b"` | Both points and lines |
| `"o"` | Overlapping points and lines |
| `"h"` | Histogram-like vertical lines |
| `"s"` | Staircase |
| `"n"` | Nothing (sets up axes only) |

```{r scatter-basic}
# Basic scatter plot
plot(trees$Height, trees$Volume,
     col  = "#2980B9",
     pch  = 19,
     cex  = 1.2,
     main = "Tree Height vs Volume",
     xlab = "Height (ft)",
     ylab = "Volume (cubic ft)")
```

```{r scatter-points-lines}
# Points + connecting line
heights <- c(10, 15, 20, 25, 30)
time    <- c(1, 2, 3, 4, 5)

plot(time, heights,
     type = "b",
     col  = "blue",
     pch  = 19,
     lwd  = 2,
     xlab = "Time (weeks)",
     ylab = "Plant Height (cm)",
     main = "Plant Growth Over Time")
```

### Adding Elements to an Existing Plot

Low-level functions add elements to the current plot — they do **not** open a new one:

```{r scatter-add-elements}
# Base scatter plot
plot(ecole$info, ecole$histoire,
     pch  = 19,
     col  = "grey70",
     cex  = 0.6,
     main = "Info vs Histoire Scores",
     xlab = "Info Score",
     ylab = "Histoire Score")

# Add a linear regression line
abline(lm(histoire ~ info, data = ecole), col = "blue", lwd = 2)

# Add a LOWESS smoothing curve (no linearity assumption)
lines(lowess(ecole$info, ecole$histoire), col = "red", lwd = 2, lty = 2)

# Add a horizontal reference line at y = 6
abline(h = 6, col = "orange", lty = 3)

# Add a text annotation
text(x = 8, y = 2, "Linear fit (blue)\nLOWESS (red)", col = "black", cex = 0.8)

# Add a legend
legend("topright",
       legend = c("Linear fit", "LOWESS"),
       col    = c("blue", "red"),
       lty    = c(1, 2),
       lwd    = 2)
```

### Multiple Lines with `lines()`

```{r multi-line}
line1 <- c(2, 3, 4, 5, 6, 7)
line2 <- c(3, 5, 7, 9, 11, 13)

plot(line1, type = "l", col = "darkgreen", lwd = 2,
     ylim = c(0, 15),
     main = "Two Lines on One Plot",
     xlab = "Index", ylab = "Value")

lines(line2, type = "l", col = "darkblue", lwd = 2, lty = 2)

legend("topleft",
       legend = c("Line 1", "Line 2"),
       col    = c("darkgreen", "darkblue"),
       lty    = c(1, 2), lwd = 2)
```

### Line Plot for Time Series

```{r line-time-series}
months <- c("Jan","Feb","Mar","Apr","May","Jun",
            "Jul","Aug","Sep","Oct","Nov","Dec")
cases  <- c(50, 60, 55, 70, 80, 90, 85, 95, 100, 110, 120, 130)

plot(cases, type = "o", col = "blue", pch = 19,
     xaxt = "n",
     xlab = "Month", ylab = "Reported Cases",
     main = "Monthly Disease Cases in 2023",
     lwd = 2)
axis(1, at = 1:12, labels = months, las = 2)  # las=2 rotates labels
```

---

## Histogram — `hist()`

Used to display the **distribution of a single numeric variable**.

```{r histogram-basic}
hist(trees$Volume,
     col    = "darkred",
     border = "white",
     main   = "Distribution of Tree Volume",
     xlab   = "Volume (cubic ft)",
     ylab   = "Frequency")
```

```{r histogram-density}
# Show density instead of count on y-axis
hist(ecole$histoire,
     freq   = FALSE,
     col    = "lightblue",
     border = "white",
     breaks = 10,
     main   = "Histoire Score Distribution (Density)",
     xlab   = "Score",
     ylab   = "Density")

# Overlay a normal density curve
curve(dnorm(x, mean = mean(ecole$histoire),
               sd   = sd(ecole$histoire)),
      add = TRUE, col = "red", lwd = 2)
```

### Controlling the Number of Bins — `breaks`

```{r histogram-breaks}
par(mfrow = c(1, 3))  # 3 plots side by side
hist(ecole$histoire, breaks = 3,  main = "breaks = 3",  col = "lightgreen")
hist(ecole$histoire, breaks = 10, main = "breaks = 10", col = "lightblue")
hist(ecole$histoire, breaks = 30, main = "breaks = 30", col = "lightyellow")
par(mfrow = c(1, 1))  # reset
```

> 💡 More breaks = more detail, but too many make the plot noisy. A rule of thumb: use approximately √n bins where n is the sample size.

---

## Density Plot — `density()` + `plot()`

A smoother alternative to the histogram:

```{r density-base}
plot(density(ecole$histoire),
     col  = "steelblue",
     lwd  = 2,
     main = "Density of Histoire Scores",
     xlab = "Score")

# Overlay multiple density curves
plot(density(ecole$histoire[ecole$sexe == "Female"]),
     col  = "red",  lwd = 2,
     main = "Histoire Score Density by Sex",
     xlab = "Score", xlim = c(0, 12))

lines(density(ecole$histoire[ecole$sexe == "Male"]),
      col  = "blue", lwd = 2)

legend("topright",
       legend = c("Female", "Male"),
       col    = c("red", "blue"),
       lwd    = 2)
```

---

## Boxplot — `boxplot()`

Displays the five-number summary (min, Q1, median, Q3, max) and outliers.

```{r boxplot-single}
boxplot(mtcars$mpg,
        col        = "lightblue",
        main       = "Miles per Gallon",
        ylab       = "MPG",
        horizontal = FALSE)
```

```{r boxplot-grouped}
# Grouped boxplot using formula notation: y ~ group
boxplot(mpg ~ cyl, data = mtcars,
        col  = c("#AED6F1", "#F9E79F", "#ABEBC6"),
        main = "Fuel Economy by Number of Cylinders",
        xlab = "Cylinders",
        ylab = "Miles per Gallon")
```

```{r boxplot-two-factors}
# Two grouping factors
boxplot(info ~ sexe * classe, data = ecole,
        col    = c("lightgreen", "lightyellow"),
        notch  = TRUE,
        main   = "Info Score by Sex and Class",
        xlab   = "Sex × Class",
        ylab   = "Info Score",
        las    = 2)
```

```{r boxplot-all-vars}
# All variables of a data frame at once
boxplot(trees,
        col  = c("#F1948A", "#82E0AA", "#85C1E9"),
        main = "Distribution of All Tree Variables")
```

---

## Bar Chart — `barplot()`

Used for **categorical or discrete** data.

```{r barplot-basic}
# Named vector
malaria_2024 <- c(Centre = 52000, Littoral = 47000,
                  Far_North = 125000, North_West = 68000)

barplot(malaria_2024,
        col  = c("skyblue", "orange", "red", "purple"),
        main = "Malaria Cases by Region (2024)",
        ylab = "Cases",
        las  = 2)
```

```{r barplot-from-table}
# From a frequency table
barplot(table(ecole$classe),
        col  = c("#3498DB", "#E74C3C", "#2ECC71"),
        main = "Number of Students per Class",
        xlab = "Class",
        ylab = "Count")
```

```{r barplot-stacked}
# Stacked bars: two categorical variables
barplot(table(ecole$sexe, ecole$classe),
        col         = c("#E74C3C", "#3498DB"),
        main        = "Sex Distribution by Class",
        xlab        = "Class",
        ylab        = "Count",
        legend.text = TRUE,
        args.legend = list(x = "topright"))
```

```{r barplot-sidebyside}
# Side-by-side bars (beside = TRUE)
malaria_compare <- rbind(
  `2023` = c(48000, 43000, 117000, 62000),
  `2024` = c(52000, 47000, 125000, 68000)
)
colnames(malaria_compare) <- c("Centre", "Littoral", "Far_North", "North_West")

barplot(malaria_compare,
        beside = TRUE,
        col    = c("darkgreen", "darkred"),
        main   = "Malaria Cases 2023 vs 2024",
        ylab   = "Cases",
        legend.text = TRUE,
        las    = 2)
```

---

## Pie Chart — `pie()`

Useful for showing the **proportional composition** of a whole.

```{r piechart}
land_cover  <- c(79.814, 0.023, 16.21, 1.636, 2.318)
categories  <- c("Barren Land", "Water", "Built-up Area", "Tree", "Grass")
pct         <- round(land_cover / sum(land_cover) * 100, 1)
labels      <- paste(categories, pct, "%", sep = " ")

pie(land_cover,
    labels = labels,
    col    = c("tan", "blue", "grey50", "darkgreen", "lightgreen"),
    main   = "Land Cover Distribution 2013")
```

> ⚠️ Pie charts are difficult to read precisely. For accurate comparisons, prefer bar charts. Reserve pie charts for highlighting a dominant proportion.

---

## Dot Chart — `dotchart()`

A cleaner alternative to bar charts for ranking:

```{r dotchart}
dotchart(sort(malaria_2024),
         main = "Regional Malaria Cases (2024)",
         xlab = "Cases",
         col  = "steelblue",
         pch  = 19)
```

---

## Scatterplot Matrix — `pairs()`

Displays all pairwise scatter plots in a single grid:

```{r pairs}
pairs(~ age + anglais + info + art + histoire,
      data = ecole,
      main = "Scatterplot Matrix — School Data",
      col  = "steelblue",
      pch  = 16,
      cex  = 0.3)
```

---

## QQ-Plot — `qqnorm()` and `qqline()`

Tests whether a variable follows a normal distribution. Points on the line = normal:

```{r qqplot}
qqnorm(ecole$histoire,
       main = "QQ-Plot: Histoire Scores",
       pch  = 19,
       col  = "steelblue",
       cex  = 0.7)
qqline(ecole$histoire, col = "red", lwd = 2)
```

---

## Multi-Panel Layouts — `par(mfrow = ...)`

`par(mfrow = c(rows, cols))` divides the plot window into a grid:

```{r par-mfrow}
par(mfrow = c(2, 2))   # 2 rows × 2 columns

hist(ecole$histoire[ecole$sexe == "Female"],
     main = "Histoire — Female", col = "#F1948A", xlab = "Score")
hist(ecole$histoire[ecole$sexe == "Male"],
     main = "Histoire — Male",   col = "#85C1E9", xlab = "Score")
boxplot(histoire ~ classe, data = ecole,
        main = "Histoire by Class", col = c("lightblue","lightyellow","lightgreen"))
plot(density(ecole$moyenne),
     main = "Overall Average Density", col = "purple", lwd = 2)

par(mfrow = c(1, 1))   # always reset after multi-panel
```

---

## Adding Annotations

```{r annotations}
plot(1:10, 1:10, type = "b", pch = 19, col = "blue",
     main = "Annotated Plot", xlab = "X", ylab = "Y", lwd = 2)

# Horizontal reference line
abline(h = 5, col = "green",  lty = 2, lwd = 1.5)

# Vertical reference line
abline(v = 5, col = "orange", lty = 3, lwd = 1.5)

# Diagonal regression line
abline(lm(1:10 ~ 1:10), col = "red", lwd = 2)

# Text label
text(x = 5, y = 8, labels = "Midpoint", col = "black", cex = 1.1)

# Legend
legend("topleft",
       legend = c("Data", "Horizontal", "Vertical", "Regression"),
       col    = c("blue", "green", "orange", "red"),
       lty    = c(1, 2, 3, 1),
       lwd    = 2,
       pch    = c(19, NA, NA, NA))
```

---

# Point Shapes (`pch`) and Color Reference

## Point Shapes

```{r pch-reference, fig.height=3}
plot(1:20, rep(1, 20), pch = 1:20, cex = 2,
     xlab = "pch value", ylab = "",
     main = "Point Shapes (pch = 1 to 20)",
     yaxt = "n", ylim = c(0.5, 1.5))
text(1:20, rep(0.7, 20), labels = 1:20, cex = 0.7)
```

**Most used:** `pch = 19` (filled circle), `pch = 16` (filled circle), `pch = 17` (triangle), `pch = 15` (filled square).

## Colors

R accepts colors as names, hex codes, or through palettes:

```{r color-demo, fig.height=2.5}
barplot(rep(1, 8),
        col  = c("red","blue","green","orange","purple","steelblue","#E74C3C","#2ECC71"),
        names.arg = c("red","blue","green","orange","purple","steelblue","#E74C3C","#2ECC71"),
        las = 2, main = "Color Examples", border = NA)
```

Useful palette functions: `rainbow(n)`, `heat.colors(n)`, `terrain.colors(n)`, `RColorBrewer::brewer.pal(n, "Set2")`.

---

# ggplot2 — Grammar of Graphics

`ggplot2` builds plots by combining layers. The minimal template is:

```r
ggplot(data, aes(x = ..., y = ..., ...)) +
  geom_*() +
  labs() +
  theme_*()
```

## Scatter Plot

```{r gg-scatter}
ggplot(mtcars, aes(x = wt, y = mpg, color = factor(cyl))) +
  geom_point(size = 3, alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title  = "Weight vs Fuel Economy",
       x      = "Weight (1000 lbs)",
       y      = "Miles per Gallon",
       color  = "Cylinders") +
  theme_bw()
```

## Histogram

```{r gg-histogram}
ggplot(ecole, aes(x = histoire, fill = sexe)) +
  geom_histogram(binwidth = 1, alpha = 0.6, position = "identity") +
  labs(title = "Histoire Score Distribution by Sex",
       x = "Score", y = "Count", fill = "Sex") +
  theme_minimal()
```

## Density Plot

```{r gg-density}
ggplot(ecole, aes(x = moyenne, fill = classe)) +
  geom_density(alpha = 0.4) +
  labs(title = "Average Score Density by Class",
       x = "Average Score", fill = "Class") +
  theme_bw()
```

## Boxplot

```{r gg-boxplot}
ggplot(ecole, aes(x = classe, y = anglais, fill = classe)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21) +
  geom_jitter(width = 0.15, alpha = 0.2, size = 1) +
  labs(title = "English Score by Class",
       x = "Class", y = "Score") +
  theme_bw() +
  theme(legend.position = "none")
```

## Bar Chart

```{r gg-bar}
ggplot(ecole, aes(x = classe, fill = sexe)) +
  geom_bar(position = "dodge", alpha = 0.8) +
  labs(title = "Student Count by Class and Sex",
       x = "Class", y = "Count", fill = "Sex") +
  theme_bw()
```

## Line Plot — Time Series

```{r gg-line}
monthly <- data.frame(
  month = factor(month.abb, levels = month.abb),
  cases = c(50, 60, 55, 70, 80, 90, 85, 95, 100, 110, 120, 130),
  group = "Region A"
)

ggplot(monthly, aes(x = month, y = cases, group = group)) +
  geom_line(color = "steelblue", linewidth = 1.2) +
  geom_point(color = "steelblue", size = 3) +
  labs(title = "Monthly Disease Cases 2023",
       x = "Month", y = "Reported Cases") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Facets — Small Multiples

```{r gg-facets}
# One panel per class
ggplot(ecole, aes(x = histoire, fill = sexe)) +
  geom_histogram(binwidth = 1, alpha = 0.6, position = "identity") +
  facet_wrap(~ classe) +
  labs(title = "Histoire Score by Class and Sex",
       x = "Score", fill = "Sex") +
  theme_bw()
```

## Saving Plots with `ggsave()`

```{r ggsave-demo, eval=FALSE}
p <- ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  theme_bw()

ggsave("my_plot.png",  plot = p, width = 8, height = 6, dpi = 300)
ggsave("my_plot.pdf",  plot = p, width = 8, height = 6)        # vector format
```

---

# Heatmaps

Heatmaps visualize a **matrix of values** through color intensity. They are especially popular in genomics (gene expression), but apply to any grid of numeric values.

## Base R Heatmap — `heatmap()`

```{r heatmap-base}
# Correlation heatmap using base R
cor_mat <- cor(mtcars[, 1:7])
heatmap(cor_mat,
        col  = colorRampPalette(c("blue", "white", "red"))(50),
        main = "Correlation Heatmap — mtcars")
```

## ggplot2 Heatmap — `geom_tile()`

```{r heatmap-ggplot}
# Reshape correlation matrix to long format
library(tidyr)

cor_df <- cor(mtcars[, 1:7]) %>%
  as.data.frame() %>%
  rownames_to_column("Var1") %>%
  pivot_longer(-Var1, names_to = "Var2", values_to = "r")

ggplot(cor_df, aes(x = Var1, y = Var2, fill = r)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(r, 2)), size = 3) +
  scale_fill_gradient2(low = "steelblue", mid = "white",
                       high = "tomato", midpoint = 0,
                       limits = c(-1, 1)) +
  labs(title = "Correlation Heatmap — mtcars",
       x = NULL, y = NULL, fill = "r") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## `pheatmap` — Publication-Quality Heatmaps

The `pheatmap` package provides fine-grained control ideal for gene expression heatmaps:

```{r pheatmap-demo, eval=FALSE}
# install.packages("pheatmap")
library(pheatmap)

# Gene expression example
set.seed(42)
expr_mat <- matrix(rnorm(200), nrow = 20,
                   dimnames = list(paste0("Gene", 1:20),
                                   paste0("Sample", 1:10)))

pheatmap(expr_mat,
         scale          = "row",       # Z-score normalize each gene
         cluster_rows   = TRUE,
         cluster_cols   = TRUE,
         color          = colorRampPalette(c("#145afc","white","#ee4445"))(50),
         main           = "Gene Expression Heatmap",
         show_rownames  = TRUE,
         show_colnames  = TRUE,
         display_numbers = FALSE,
         filename       = "gene_heatmap.pdf")
```

## `tidyheatmaps` — Tidy Interface to pheatmap

```{r tidyheatmap-demo, eval=FALSE}
# install.packages("tidyheatmaps")
library(tidyheatmaps)

# Requires long-format data with rows, columns, and values variables
tidyheatmap(data_exprs,
            rows    = external_gene_name,
            columns = sample,
            values  = expression,
            scale   = "row",
            annotation_col = c(sample_type, condition, group),
            annotation_row = c(is_immune_gene, direction),
            annotation_colors = ann_colors,
            gaps_row = direction,
            gaps_col = group,
            cellwidth  = 7,
            cellheight = 7,
            show_selected_row_labels = c("Apol6", "Bsn", "Vgf", "Fam96b"),
            filename  = "tidy_heatmap.pdf")
```

---

# Choosing the Right Plot

| Data type | Goal | Recommended plot |
|---|---|---|
| 1 continuous variable | Distribution | Histogram, density plot |
| 1 continuous + 1 categorical | Compare groups | Boxplot, violin plot |
| 2 continuous variables | Relationship | Scatter plot |
| Many continuous variables | Pairwise relationships | `pairs()`, correlation heatmap |
| 1 categorical variable | Counts | Bar chart, dot chart |
| 2 categorical variables | Association | Grouped bar chart, mosaic plot |
| Proportions of a whole | Composition | Pie chart (use sparingly), stacked bar |
| Continuous over time | Trend | Line plot |
| Matrix of values | Intensity | Heatmap |
| Normality check | Distributional shape | QQ-plot |

---

# Base R vs ggplot2 — Side-by-Side

```{r comparison, fig.width=10, fig.height=4}
par(mfrow = c(1, 2))

# Base R
boxplot(mpg ~ cyl, data = mtcars,
        col  = c("#AED6F1","#F9E79F","#ABEBC6"),
        main = "Base R Boxplot",
        xlab = "Cylinders", ylab = "MPG")

# ggplot2 equivalent
print(
  ggplot(mtcars, aes(x = factor(cyl), y = mpg, fill = factor(cyl))) +
    geom_boxplot(alpha = 0.7) +
    labs(title = "ggplot2 Boxplot",
         x = "Cylinders", y = "MPG") +
    theme_bw() +
    theme(legend.position = "none")
)

par(mfrow = c(1, 1))
```

---

# Saving Plots

## Base R — `png()`, `pdf()`, `jpeg()`

```{r save-base, eval=FALSE}
# PNG file
png("my_plot.png", width = 800, height = 600, res = 150)
hist(mtcars$mpg, col = "steelblue", main = "MPG Distribution")
dev.off()    # MUST close the device

# PDF (vector — infinitely scalable)
pdf("my_plot.pdf", width = 8, height = 6)
boxplot(mpg ~ cyl, data = mtcars, col = "lightblue")
dev.off()
```

## ggplot2 — `ggsave()`

```{r save-ggplot, eval=FALSE}
p <- ggplot(mtcars, aes(wt, mpg, color = factor(cyl))) +
  geom_point(size = 3) +
  theme_bw()

ggsave("scatter.png", plot = p, width = 8, height = 6, dpi = 300)
ggsave("scatter.pdf", plot = p, width = 8, height = 6)
```

---

# Practice Exercises

**Exercise 1.** Using the `trees` dataset, create a 2×2 panel with: (a) histogram of `Girth`, (b) boxplot of `Height`, (c) scatter plot of `Girth` vs `Volume` with a regression line, and (d) density plot of `Volume`. Add appropriate titles and axis labels to each.

**Exercise 2.** Using the `mtcars` dataset, create a grouped boxplot of `mpg` by `cyl` with one color per group. Add a horizontal dashed line at the overall mean mpg. Include a legend.

**Exercise 3.** Create a bar chart comparing malaria cases across four Cameroon regions for two years (2023 and 2024) using side-by-side bars. Add a legend, color each year differently, and rotate the x-axis labels.

**Exercise 4.** Using the `ecole` dataset, create with `ggplot2`: (a) a density plot of `moyenne` split by `sexe`, (b) a faceted histogram of `anglais` with one panel per `classe`, (c) a boxplot of `info` grouped by `classe` with jittered points overlaid.

**Exercise 5.** Create a correlation heatmap of all numeric columns in `mtcars` using `ggplot2::geom_tile()`. Color positively correlated pairs red, negatively correlated pairs blue, and zero correlation white. Display the correlation coefficient inside each cell.

**Exercise 6.** Using the `msleep` dataset from `ggplot2`, create a publication-quality scatter plot of `sleep_total` vs `brainwt`, with point size proportional to `bodywt`, colored by `vore`, a log scale on the x-axis, and a `theme_bw()` style. Save it as a PNG file at 300 dpi.

---

```{r cleanup, include=FALSE}
rm(list = ls())
```

---

*End of Lecture — Data Visualization and Plotting in R*
