---
title: "Functions in R"
subtitle: "A Comprehensive Lecture — From Built-in Functions to Advanced Patterns"
author: "Ange Maxime TCHOUTANG"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    theme: flatly
    highlight: tango
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  comment = "##")
```

---

# What is a Function?

A **function** is a named, reusable block of code that performs a specific task. Think of it as a recipe: you define the steps once, then call the recipe by name whenever you need to repeat those steps — possibly with different ingredients each time.

Functions are the single most important tool for writing clean, efficient, maintainable R code. They allow you to:

- **Avoid repetition** — write the logic once, use it many times
- **Organize code** — break complex problems into small, testable units
- **Communicate intent** — a well-named function makes code self-documenting

In R, *everything that does something is a function* — `+`, `<-`, `[`, `print()`, even `if()` are all functions under the hood.

---

# Built-in Functions

R ships with hundreds of ready-to-use functions. You have already used many of them.

## Getting Help on Any Function

```{r help, eval=FALSE}
?mean          # Opens the help page for mean()
help("mean")   # Same result
help(sum)      # Works without quotes too
```

A full reference card of built-in functions is available at:  
<https://cran.r-project.org/doc/contrib/Short-refcard.pdf>

## Common Built-in Functions

### Math functions

```{r builtin-math}
max(20, 77, 85, 12)       # Maximum value
min(20, 77, 85, 12)       # Minimum value
abs(-7.5)                 # Absolute value
sqrt(16)                  # Square root
ceiling(4.2)              # Round up
floor(4.8)                # Round down
round(3.14159, digits = 2) # Round to 2 decimal places
log(10)                   # Natural logarithm (base e)
log10(100)                # Base-10 logarithm
exp(1)                    # e^1 ≈ 2.718
```

### Vector summary functions

```{r builtin-vector}
x <- c(4, 7, 2, 9, 1, 5, 8, 3, 6, 10)

mean(x)       # Arithmetic mean
median(x)     # Median
sd(x)         # Standard deviation
var(x)        # Variance
sum(x)        # Sum of all elements
range(x)      # Min and max as a 2-element vector
IQR(x)        # Interquartile range
length(x)     # Number of elements
sort(x)       # Sort in ascending order
rev(x)        # Reverse order
```

### String functions

```{r builtin-string}
nchar("Tchoutang")              # Number of characters
toupper("hello")                # Uppercase
tolower("WORLD")                # Lowercase
paste("R", "is", "great")       # Concatenate with space
paste0("gene", "_", "BRCA1")    # Concatenate without separator
grepl("^A", c("Apple","Banana","Avocado"))  # Pattern detection
```

### Sequence generation functions

```{r builtin-sequence}
# seq(): flexible numeric sequences
seq(from = 1, to = 10, by = 2)             # Step by 2
seq(from = 4, to = 200, by = 5)            # Larger steps
seq(from = 0, to = 1, length.out = 5)      # 5 evenly spaced values

# rep(): repeat values
rep(c(1, 2, 3), times = 3)   # Repeat the whole vector 3 times
rep(c(1, 2, 3), each  = 2)   # Repeat each element 2 times

# sample(): random sampling
sample(1:100, size = 7)                    # 7 random values, no replacement
sample(c("A","T","G","C"), 10, replace = TRUE)  # Random DNA bases
```

### Data structure inspection functions

```{r builtin-inspect}
str(airquality)       # Structure: types and first values
head(airquality)      # First 6 rows
tail(airquality, 3)   # Last 3 rows
dim(airquality)       # Rows × Columns
nrow(airquality)      # Number of rows
ncol(airquality)      # Number of columns
summary(airquality)   # Statistical summary per column
names(airquality)     # Column names
```

### Handling missing values

```{r builtin-na}
x <- c(3, 5, NA, 8, NA, 2)

is.na(x)                  # Which elements are NA?
sum(is.na(x))             # How many NAs?
mean(x)                   # NA — because of missing values
mean(x, na.rm = TRUE)     # Correct: ignore NAs
```

> 💡 Many R functions have an `na.rm` argument. Setting `na.rm = TRUE` removes missing values before the calculation.

---

# Creating Your Own Functions

## The Basic Template

```r
function_name <- function(parameter1, parameter2, ...) {
  # body: code that runs each time the function is called
  result <- ...
  return(result)
}
```

- **`function_name`** — a descriptive name (use `snake_case` or `camelCase` consistently)
- **`parameter1, parameter2`** — the inputs (also called *arguments* when passed by the caller)
- **`return(result)`** — the output; what the function hands back

## Your First Function — No Parameters

```{r first-function}
greet <- function() {
  print("Hello, everyone!")
}

# Call the function as many times as needed
greet()
greet()
greet()
```

## Function with Parameters

```{r function-params}
# Two parameters
full_name <- function(first, last) {
  paste(first, last)
}

full_name("Alice", "Smith")
full_name("Bob", "Jones")
```

## The `return()` Statement

`return()` explicitly specifies what value the function sends back. R also returns the value of the last evaluated expression even without `return()`, but using it explicitly makes code clearer and stops execution immediately.

```{r return-basic}
# With explicit return
multiply_by_10 <- function(n) {
  return(10 * n)
}

print(multiply_by_10(5))
print(multiply_by_10(15))
print(multiply_by_10(20))
```

```{r return-implicit}
# R also returns the last expression automatically
add_two <- function(x) {
  x + 2    # implicitly returned
}
add_two(6)
```

## A Function with Multiple Operations

```{r function-multi}
# Unit conversion: miles to kilometres
mile_to_km <- function(miles) {
  km <- miles * 1.60934
  return(km)
}

mile_to_km(10)
mile_to_km(26.2)   # a marathon!

# Custom mean function (reimplementing what mean() does)
my_mean <- function(v) {
  total   <- sum(v)
  n       <- length(v)
  return(total / n)
}

my_mean(c(2, 4, 6, 8, 10))
my_mean(c(10, 50, 20))
```

---

# Function Arguments in Depth

## Required vs Optional Arguments

By default, all parameters are required. Calling a function with the wrong number of arguments produces an error:

```{r required-args}
describe <- function(fname, lname) {
  paste(fname, lname)
}

describe("Peter", "Griffin")
# describe("Peter")  # Error: argument "lname" is missing
```

## Default Parameter Values

Provide a fallback value using `=` in the parameter list. Callers can override it or leave it as-is:

```{r default-args}
greet_country <- function(country = "Cameroon") {
  paste("I am from", country)
}

greet_country("France")     # Override the default
greet_country("Nigeria")
greet_country()             # Uses the default: "Cameroon"
```

```{r default-bmi}
# BMI calculator with a default height
calculate_bmi <- function(weight, height = 1.75) {
  bmi <- weight / (height^2)
  return(bmi)
}

calculate_bmi(70)            # Uses default height 1.75
calculate_bmi(70, 1.80)     # Override height
```

## Positional vs Named Arguments

Arguments can be passed by **position** (order matters) or by **name** (order does not matter):

```{r named-args}
power_message <- function(region, value, p) {
  paste(region, "raised to power", p, "is", value^p)
}

# Positional — must be in the right order
power_message("Centre", 50, 2)

# Named — order doesn't matter
power_message(value = 50, p = 2, region = "Centre")
power_message(p = 3, region = "Far North", value = 120)
```

> 💡 **Best practice:** For functions with 3+ arguments, using named arguments makes your code more readable and less error-prone.

## The `...` (Dots) Argument — Passing Extra Arguments

The `...` (pronounced "dot-dot-dot" or "ellipsis") allows a function to accept any number of additional arguments and pass them on to another function:

```{r dots-arg}
# Wraps mean() but passes any extra arguments through
my_mean_safe <- function(x, ...) {
  mean(x, ...)
}

cases <- c(850, 900, 920, 950, 975, NA)

my_mean_safe(cases)                    # Returns NA (because of NA)
my_mean_safe(cases, na.rm = TRUE)      # Passes na.rm to mean()
```

This is the pattern used in most wrapper functions throughout the tidyverse and base R.

---

# Variable Scope

**Scope** determines where a variable is visible — which parts of your code can read or write it.

## Local Variables (Inside a Function)

Variables created inside a function are **local** — they exist only during that function call and disappear afterward:

```{r local-scope}
compute <- function(x) {
  y <- x + 100   # y is LOCAL — only lives inside this function
  return(y)
}

compute(5)
exists("y")   # FALSE — y does not exist in the global environment
```

## Global Variables (Outside a Function)

Variables created outside any function are **global** and can be read from inside any function:

```{r global-read}
species_name <- "Homo sapiens"

display_species <- function() {
  paste("Organism:", species_name)   # Reads the global variable
}

display_species()
```

## Local vs Global — Name Shadowing

If you create a variable *inside* a function with the same name as a global variable, the local copy **shadows** the global one — but does not change it:

```{r name-shadow}
txt <- "global variable"

my_function <- function() {
  txt <- "local variable"    # Creates a local copy
  paste("Inside function: R is", txt)
}

my_function()
txt    # Global is unchanged: still "global variable"
```

## The Global Assignment Operator `<<-`

`<<-` forces a variable to be written into the **global** (or parent) environment from inside a function:

```{r global-assign}
my_function <- function() {
  new_str <<- "powerful!"    # Writes to the global environment
  paste("R is", new_str)
}

my_function()
print(new_str)   # Now visible globally
```

> ⚠️ Use `<<-` sparingly. It creates hidden side effects that make functions harder to test and debug. Prefer returning values explicitly with `return()`.

## Scope Summary

| Variable created… | Visible… | Modified with… |
|---|---|---|
| Outside any function | Everywhere (global) | `<-` or `=` |
| Inside a function | Only inside that function (local) | `<-` or `=` |
| Inside a function, using `<<-` | Globally from inside the function | `<<-` |

---

# Nested Functions

## Calling a Function Inside Another Function

```{r nested-call}
add <- function(x, y) {
  return(x + y)
}

# Use add() inside another call to add()
add(add(2, 2), add(3, 3))   # (2+2) + (3+3) = 10
```

## Defining a Function Inside Another Function

An **inner function** (closure) captures variables from its enclosing function:

```{r nested-define}
outer_func <- function(x) {
  inner_func <- function(y) {
    return(x + y)   # x is captured from outer_func's scope
  }
  return(inner_func)   # Return the inner function itself
}

# Step 1: call outer_func to create the inner function with x = 3
add_three <- outer_func(3)

# Step 2: call the resulting function with y = 5
add_three(5)    # 3 + 5 = 8
add_three(10)   # 3 + 10 = 13
```

This pattern — returning a function that "remembers" some of the outer state — is called a **closure** and is used extensively in functional programming.

---

# Recursive Functions

A **recursive function** calls itself. It is useful for problems that can naturally be broken into smaller identical sub-problems (e.g. factorials, Fibonacci numbers, tree traversal).

Every recursive function needs:
1. A **base case** — a condition that stops the recursion
2. A **recursive call** — where the function calls itself with a simpler input

```{r recursion-triangular}
# Triangular number: sum of 1 + 2 + 3 + ... + k
tri_recursion <- function(k) {
  if (k > 0) {
    result <- k + tri_recursion(k - 1)   # recursive call
    return(result)
  } else {
    return(0)   # base case: stop at 0
  }
}

tri_recursion(6)   # 6 + 5 + 4 + 3 + 2 + 1 = 21
```

```{r recursion-factorial}
# Factorial: n! = n × (n-1)!
factorial_r <- function(n) {
  if (n <= 1) return(1)           # base case
  return(n * factorial_r(n - 1))  # recursive call
}

factorial_r(5)   # 5! = 120
factorial_r(10)
```

> ⚠️ Recursion can consume a lot of memory and time if not designed carefully. R also has a default recursion limit. For large inputs, prefer iterative loops.

---

# Applying Functions over Data: `lapply()`, `sapply()`, `apply()`

Instead of writing a `for` loop to apply the same function to every column or element, R provides a family of **apply functions** that are more concise and often faster.

## `lapply()` — Apply to a List, Return a List

```{r lapply}
# Apply mean() to every column of airquality
result_list <- lapply(airquality, mean, na.rm = TRUE)
result_list
```

## `sapply()` — Apply to a List, Return a Simplified Vector

```{r sapply}
# Same as lapply but returns a named numeric vector
sapply(airquality, mean, na.rm = TRUE)

# sapply with a custom function
sapply(airquality, function(col) sum(is.na(col)))  # Count NAs per column
```

## `apply()` — Apply over Rows or Columns of a Matrix

```{r apply-matrix}
m <- matrix(1:12, nrow = 3, ncol = 4)

apply(m, MARGIN = 1, FUN = sum)   # MARGIN = 1 → apply over rows
apply(m, MARGIN = 2, FUN = mean)  # MARGIN = 2 → apply over columns
```

## Equivalence with a `for` Loop

```{r apply-vs-for}
# These two are equivalent:

# for loop version
for (i in 1:ncol(airquality)) {
  cat(names(airquality)[i], ":", mean(airquality[, i], na.rm = TRUE), "\n")
}

# sapply version — much shorter
sapply(airquality, mean, na.rm = TRUE)
```

---

# `do.call()` — Call a Function Programmatically

`do.call()` calls a function and passes a **list** as its arguments. This is powerful when the number of arguments is determined at runtime:

```{r do-call}
# Build a repeated vector dynamically
make_rep <- function(a) rep(a, 3)

do.call(make_rep, list(a = 5))   # Equivalent to make_rep(a = 5)

# Combine multiple data frames into one
df1 <- data.frame(x = 1:3, y = c("a", "b", "c"))
df2 <- data.frame(x = 4:6, y = c("d", "e", "f"))
df3 <- data.frame(x = 7:9, y = c("g", "h", "i"))

do.call(rbind, list(df1, df2, df3))   # Stack all three at once
```

---

# Functions with `if/else` for Validation

Functions often need to validate their inputs before doing any work:

```{r function-validation}
classify_score <- function(score) {
  if (!is.numeric(score)) return("Error: score must be numeric")
  if (score < 0 | score > 100) return("Error: score must be between 0 and 100")

  if (score >= 90) return("A")
  else if (score >= 80) return("B")
  else if (score >= 70) return("C")
  else if (score >= 60) return("D")
  else return("F")
}

classify_score(95)
classify_score(72)
classify_score(42)
classify_score(-5)
classify_score("high")
```

---

# Functions with Loops Inside

Functions can contain any R code, including loops:

```{r function-loop}
# Print the squares of 1 to n
print_squares <- function(n) {
  for (i in 1:n) {
    print(paste(i, "squared is", i^2))
  }
}

print_squares(5)
```

```{r function-loop-2}
# Calculate GC content of a DNA sequence
gc_content <- function(sequence) {
  bases      <- strsplit(sequence, "")[[1]]
  seq_length <- length(bases)
  g_count    <- sum(bases == "G")
  c_count    <- sum(bases == "C")
  gc_pct     <- (g_count + c_count) / seq_length * 100
  return(round(gc_pct, 2))
}

gc_content("ATGCGCGGCCAATTTTGGCCGCGG")
gc_content("AAAATTTTGGGG")
```

---

# Packages — Extending R with More Functions

R packages are collections of functions, data, and documentation bundled together. Think of them as *toolboxes* — you install them once and load them whenever needed.

## Installing and Loading Packages

```{r packages, eval=FALSE}
# Install once (only need to do this once per machine)
install.packages("dplyr")
install.packages("ggplot2")

# Load for the current session (must do every time you restart R)
library(dplyr)
library(ggplot2)
```

## Using a Package Without Loading It

You can call a function from a package without loading the whole package using `::`:

```{r double-colon, eval=FALSE}
dplyr::filter(mtcars, cyl == 6)   # Use filter() from dplyr without library(dplyr)
```

## Discovering Functions in a Package

```{r package-explore, eval=FALSE}
ls("package:dplyr")    # List all exported functions in dplyr
help(package = "dplyr") # Open the package help page
```

## Getting Help on Functions

```{r help-functions, eval=FALSE}
?mean           # Help page for mean()
?dplyr::filter  # Help page for dplyr's filter()
args(mean)      # Show function signature (parameters)
```

---

# Complete Worked Examples

## Example 1 — Fraction to Percentage Formatter

```{r example-percent}
add_percent <- function(x, digits = 1) {
  pct    <- round(x * 100, digits = digits)
  result <- paste0(pct, "%")
  return(result)
}

positivity_rates <- c(0.245, 0.318, 0.402, 0.178)
add_percent(positivity_rates)
add_percent(positivity_rates, digits = 0)
```

## Example 2 — BMI Classifier

```{r example-bmi}
bmi_category <- function(weight_kg, height_m = 1.70) {
  bmi <- weight_kg / height_m^2
  category <- if (bmi < 18.5) "Underweight"
              else if (bmi < 25.0) "Normal weight"
              else if (bmi < 30.0) "Overweight"
              else "Obese"
  return(paste0("BMI = ", round(bmi, 1), " (", category, ")"))
}

bmi_category(70)
bmi_category(50, 1.65)
bmi_category(100, 1.60)
```

## Example 3 — Bacterial Growth Rate

```{r example-growth}
bacterial_growth_rate <- function(initial, final, time) {
  rate <- (log(final) - log(initial)) / time
  return(round(rate, 4))
}

# Initial = 100 cells, Final = 1000 cells, over 5 hours
r <- bacterial_growth_rate(100, 1000, 5)
print(paste("Bacterial growth rate:", r, "per hour"))
```

## Example 4 — DNA to RNA Conversion

```{r example-dna-rna}
dna_to_rna <- function(dna_base) {
  dna_base <- toupper(dna_base)
  switch(dna_base,
    "A" = "U",
    "T" = "A",
    "G" = "C",
    "C" = "G",
    "Invalid nucleotide"   # default
  )
}

dna_to_rna("A")
dna_to_rna("T")
dna_to_rna("G")
dna_to_rna("X")

# Apply to a whole sequence using sapply
dna_seq <- c("A", "T", "G", "C", "A", "T")
sapply(dna_seq, dna_to_rna)
```

## Example 5 — Custom Statistical Summary

```{r example-stats}
# Custom mode function (R has no built-in mode for statistics)
get_mode <- function(v) {
  uniq_vals <- unique(v)
  uniq_vals[which.max(tabulate(match(v, uniq_vals)))]
}

my_summary <- function(x) {
  cat("N:       ", length(x), "\n")
  cat("Mean:    ", round(mean(x, na.rm = TRUE), 3), "\n")
  cat("Median:  ", median(x, na.rm = TRUE), "\n")
  cat("Mode:    ", get_mode(x[!is.na(x)]), "\n")
  cat("SD:      ", round(sd(x, na.rm = TRUE), 3), "\n")
  cat("Min:     ", min(x, na.rm = TRUE), "\n")
  cat("Max:     ", max(x, na.rm = TRUE), "\n")
  cat("NAs:     ", sum(is.na(x)), "\n")
}

centre_daily <- c(102, 115, 110, 120, 118, 125, 130, 118, 122, 115)
my_summary(centre_daily)
```

## Example 6 — Function with `dplyr` Integration

```{r example-dplyr, message=FALSE}
library(tidyverse)

get_molecular_weight_kDa <- function(protein_name, dataset) {
  result <- dataset %>%
    filter(Protein == protein_name) %>%
    mutate(Weight_kDa = Weight_Da / 1000) %>%
    select(Protein, Weight_kDa)

  if (nrow(result) == 0) {
    return(paste("Protein", protein_name, "not found."))
  }
  return(result)
}

# Sample dataset
proteins <- data.frame(
  Protein    = c("Hemoglobin", "Myosin", "Collagen", "Insulin", "Albumin"),
  Weight_Da  = c(64500, 223000, 300000, 5800, 66400)
)

get_molecular_weight_kDa("Hemoglobin", proteins)
get_molecular_weight_kDa("Insulin",    proteins)
get_molecular_weight_kDa("Keratin",    proteins)
```

---

# Function Design Best Practices

| Principle | Recommendation |
|---|---|
| **Single responsibility** | Each function should do exactly one thing |
| **Descriptive names** | `calculate_gc_content()` not `f1()` |
| **Default arguments** | Provide sensible defaults for optional parameters |
| **Validate inputs** | Check types and ranges at the start; fail early with clear messages |
| **Use `return()` explicitly** | Makes the output obvious, even if implicit return works |
| **Avoid `<<-`** | Prefer passing data in and returning results out |
| **Document** | Add a comment above the function explaining what it does, its inputs, and its output |
| **Keep it short** | If a function is longer than ~20 lines, consider splitting it |

---

# Quick Reference

## Function Anatomy

```r
my_function <- function(x, y = 10, ...) {
  # local variable — invisible outside
  result <- x + y
  return(result)
}
```

## Calling Styles

```r
my_function(5)           # positional: x = 5, y uses default 10
my_function(5, 20)       # positional: x = 5, y = 20
my_function(x = 5, y = 20)  # named — order doesn't matter
my_function(y = 20, x = 5)  # same result
```

## Apply Family

| Function | Input | Output | Typical use |
|---|---|---|---|
| `lapply(X, FUN)` | list/vector | list | Apply to each element |
| `sapply(X, FUN)` | list/vector | vector/matrix | Like `lapply` but simplified |
| `apply(X, MARGIN, FUN)` | matrix | vector | Row (1) or column (2) operation |
| `do.call(FUN, args)` | function + list | anything | Call function with a list of args |

---

# Practice Exercises

**Exercise 1.** Write a function `celsius_to_fahrenheit(temp_c)` that converts Celsius to Fahrenheit (formula: `F = C * 9/5 + 32`). Add a second function `fahrenheit_to_celsius(temp_f)`. Test both with at least three values.

**Exercise 2.** Write a function `describe_vector(v)` that prints the mean, median, standard deviation, minimum, maximum, and number of NAs of any numeric vector. Handle the case where the vector is entirely NA.

**Exercise 3.** Write a function `count_nucleotide(sequence, nucleotide)` that takes a DNA string (e.g. `"ATGCGATC"`) and a nucleotide character (e.g. `"G"`) and returns the count of that nucleotide. Make the nucleotide argument default to `"G"`. Use it with `sapply()` to count all four nucleotides at once.

**Exercise 4.** Write a recursive function `fibonacci(n)` that returns the nth Fibonacci number (F(1) = 1, F(2) = 1, F(n) = F(n-1) + F(n-2)). Compute the first 10 Fibonacci numbers using `sapply()`.

**Exercise 5.** Write a function `summarise_group(df, group_col, value_col)` that takes a data frame and two column names (as strings), groups by `group_col`, and returns the mean and count of `value_col` per group. Use it on the built-in `InsectSprays` dataset with `group_col = "spray"` and `value_col = "count"`.

**Exercise 6.** Write a function `is_palindrome(sequence)` that checks whether a DNA sequence reads the same forwards and backwards (e.g. `"ATGGTA"` is NOT a palindrome, but `"GCATCATG"` is). Return `TRUE` or `FALSE`. Test it on at least four sequences.

---

```{r cleanup, include=FALSE}
rm(list = ls())
```

---

*End of Lecture — Functions in R*
