---
title: "Lecture 00_Introduction to R and RStudio"
author: "Ange Maxime TCHOUTANG"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    number_sections: true
  pdf_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,         # Show R code in the output
  message = FALSE,     # Suppress package messages
  warning = FALSE,     # Suppress warnings
  comment = "#>"       # Prefix for output lines
)
```

---

# Introduction

Welcome to the first lecture of this data science course using R. Before diving into data manipulation, visualization, or modeling, we need to set up a solid foundation by understanding the tools we will be working with every day.

By the end of this lecture, you will be able to:

- Understand what R is and why it is widely used in data science
- Understand the role of RStudio as a development environment
- Navigate the RStudio interface confidently
- Know the main application domains where R is used
- Write and run your first R commands

---

# What is R?

## Definition

**R** is a free, open-source **programming language and software environment** designed specifically for **statistical computing, data analysis, and graphical visualization**.

It was first developed in the early 1990s by **Ross Ihaka** and **Robert Gentleman** at the University of Auckland, New Zealand — which is where the name "R" comes from (the first letter of both authors' first names). It was also inspired by an earlier language called **S**, developed at Bell Laboratories.

Today, R is maintained by the **R Core Team** and is distributed freely through the **Comprehensive R Archive Network (CRAN)** at [https://cran.r-project.org](https://cran.r-project.org).

## Key characteristics of R

| Feature | Description |
|---|---|
| **Free & open-source** | No license cost; source code is publicly available |
| **Cross-platform** | Runs on Windows, macOS, and Linux |
| **Interpreted language** | Commands are executed line by line, interactively |
| **Vectorized** | Operations apply to entire datasets, not just individual values |
| **Extensible** | Thousands of packages available via CRAN, Bioconductor, and GitHub |
| **Active community** | Large global community of users and developers |

## R vs other languages

R is not the only language used in data science. Here is how it compares to the most common alternatives:

| | **R** | **Python** | **Excel** |
|---|---|---|---|
| Primary use | Statistics & data analysis | General purpose + data science | Spreadsheet & business reporting |
| Visualization | Excellent (ggplot2) | Good (matplotlib, seaborn) | Basic |
| Statistical modeling | Excellent (built-in) | Good (scikit-learn, statsmodels) | Limited |
| Learning curve | Moderate | Moderate | Low |
| Cost | Free | Free | Paid (Microsoft 365) |
| Reproducibility | High (R Markdown) | High (Jupyter) | Low |

> **Note:** R and Python are often used together in data science workflows. Knowing R is a strong asset, particularly in academia, biostatistics, econometrics, and social sciences.

## A quick first look at R code

Let's verify that R is working and explore a few basic operations:

```{r first-commands}
# Basic arithmetic
2 + 3
10 / 4
2^8   # 2 to the power of 8

# Creating a variable
my_name <- "Data Scientist"
print(my_name)

# Creating a numeric vector
ages <- c(23, 31, 27, 45, 19, 38)
ages

# Summary statistics in one line
summary(ages)
```

Even from these few lines, you can already see that R is clean, readable, and powerful. The `<-` symbol is the standard **assignment operator** in R (it assigns a value to a variable name).

---

# What is RStudio?

## Definition

**RStudio** is a free, open-source **Integrated Development Environment (IDE)** for R. Think of it as the cockpit from which you pilot R.

While R itself can be used from a basic terminal window, RStudio provides a much more comfortable and productive environment with:

- A code editor with syntax highlighting and auto-completion
- A console to run R commands interactively
- A file browser and project manager
- Built-in tools for viewing plots, data, and help documentation
- Support for R Markdown, version control (Git), and much more

> **Important distinction:** R and RStudio are two separate pieces of software. R is the engine; RStudio is the dashboard. You need to install R first, then RStudio on top of it.

## Installing R and RStudio

1. **Install R** from CRAN: [https://cran.r-project.org](https://cran.r-project.org)
   - Choose your operating system and follow the installer instructions.

2. **Install RStudio Desktop** (free version): [https://posit.co/download/rstudio-desktop](https://posit.co/download/rstudio-desktop)
   - RStudio will automatically detect your R installation.

---

# The RStudio Interface

When you first open RStudio, you are greeted by an interface divided into **four main panes**. Understanding each one is key to working efficiently.

## The Four Panes

```
+---------------------------+---------------------------+
|                           |                           |
|   1. SOURCE EDITOR        |   3. ENVIRONMENT /        |
|   (Scripts & R Markdown)  |      HISTORY              |
|                           |                           |
+---------------------------+---------------------------+
|                           |                           |
|   2. CONSOLE              |   4. FILES / PLOTS /      |
|   (Interactive R)         |      PACKAGES / HELP      |
|                           |                           |
+---------------------------+---------------------------+
```

### Pane 1 — The Source Editor (top-left)

This is where you **write and save your R scripts or R Markdown files**. Code written here is not executed until you explicitly run it (using `Ctrl+Enter` on a selected line, or `Ctrl+Shift+Enter` to run the whole file).

Key features:
- Syntax highlighting (R, Python, SQL, Markdown, and more)
- Auto-completion with `Tab`
- Line numbers
- Code folding (collapse/expand sections)
- Direct knitting of R Markdown files via the **Knit** button

### Pane 2 — The Console (bottom-left)

The console is the **direct interface with R**. Any command typed here is executed immediately when you press `Enter`. It shows the output of your code, messages, warnings, and errors.

```{r console-demo}
# Try typing these directly in the console:
cat("Hello, R World!\n")
Sys.time()   # Current date and time
R.version$version.string   # Your R version
```

You can use the **up arrow** key in the console to recall previously typed commands — very handy!

### Pane 3 — Environment / History (top-right)

This pane has two key tabs:

- **Environment tab** — shows all objects currently loaded in your R session (variables, data frames, functions, etc.). You can click on any object to inspect it.
- **History tab** — a log of every command you have run in the current session.

```{r environment-demo}
# Create some objects and watch them appear in the Environment pane
x <- 42
my_vector <- c(1, 2, 3, 4, 5)
my_dataframe <- data.frame(
  name  = c("Alice", "Bob", "Charlie"),
  score = c(88, 92, 75)
)

# View the data frame
my_dataframe
```

### Pane 4 — Files / Plots / Packages / Help (bottom-right)

This pane is a multi-purpose hub with several tabs:

| Tab | Purpose |
|---|---|
| **Files** | Navigate your computer's file system, set working directory |
| **Plots** | View graphics and plots generated by your code |
| **Packages** | Browse, install, and load R packages |
| **Help** | Read documentation for any R function |
| **Viewer** | Preview HTML outputs, R Markdown documents, Shiny apps |

```{r first-plot}
# A simple plot — watch it appear in the Plots tab!
barplot(
  height = c(88, 92, 75),
  names.arg = c("Alice", "Bob", "Charlie"),
  col = c("#3498db", "#2ecc71", "#e74c3c"),
  main = "Student Scores",
  ylab = "Score",
  ylim = c(0, 100)
)
```

## Useful Keyboard Shortcuts

Mastering these shortcuts will significantly speed up your workflow:

| Action | Windows / Linux | macOS |
|---|---|---|
| Run current line / selection | `Ctrl + Enter` | `Cmd + Enter` |
| Run entire script | `Ctrl + Shift + Enter` | `Cmd + Shift + Enter` |
| Insert `<-` (assignment) | `Alt + -` | `Option + -` |
| Insert pipe `%>%` | `Ctrl + Shift + M` | `Cmd + Shift + M` |
| New R script | `Ctrl + Shift + N` | `Cmd + Shift + N` |
| Save file | `Ctrl + S` | `Cmd + S` |
| Comment / uncomment lines | `Ctrl + Shift + C` | `Cmd + Shift + C` |
| Knit R Markdown document | `Ctrl + Shift + K` | `Cmd + Shift + K` |
| Clear console | `Ctrl + L` | `Ctrl + L` |

## RStudio Projects

A very important feature in RStudio is the concept of a **Project** (`.Rproj` file). A project:

- Keeps all your scripts, data, and outputs for one analysis in the same folder
- Automatically sets the working directory to that folder
- Integrates with Git for version control
- Makes your work **fully reproducible and portable**

> **Best practice:** Always work inside an RStudio Project. Go to **File → New Project** to create one.

---

# Applications of R

R is used across a wide range of fields. Here is an overview of its most important application domains, with code examples.

## Statistics and Data Analysis

This is R's home territory. R was born as a statistical language, and it remains the gold standard for statistical analysis.

```{r statistics}
# Generate sample data
set.seed(42)
group_A <- rnorm(n = 30, mean = 70, sd = 10)
group_B <- rnorm(n = 30, mean = 75, sd = 12)

# Descriptive statistics
cat("--- Group A ---\n")
cat("Mean:", round(mean(group_A), 2), "\n")
cat("SD:  ", round(sd(group_A), 2), "\n")

cat("\n--- Group B ---\n")
cat("Mean:", round(mean(group_B), 2), "\n")
cat("SD:  ", round(sd(group_B), 2), "\n")

# Inferential statistics: two-sample t-test
t.test(group_A, group_B)
```

## Data Visualization

R's `ggplot2` package is one of the most powerful and elegant data visualization tools available in any language.

```{r visualization}
# Install ggplot2 if needed: install.packages("ggplot2")
library(ggplot2)

# Using the built-in 'mtcars' dataset
ggplot(data = mtcars, aes(x = wt, y = mpg, color = factor(cyl))) +
  geom_point(size = 3, alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_manual(
    values = c("4" = "#3498db", "6" = "#2ecc71", "8" = "#e74c3c"),
    name = "Cylinders"
  ) +
  labs(
    title = "Car Weight vs. Fuel Efficiency",
    subtitle = "Data from the mtcars dataset",
    x = "Weight (1000 lbs)",
    y = "Miles per Gallon (mpg)"
  ) +
  theme_minimal()
```

## Data Manipulation and Wrangling

The `tidyverse` ecosystem (especially `dplyr` and `tidyr`) makes cleaning and transforming data fast and readable.

```{r data-wrangling}
# Install if needed: install.packages("dplyr")
library(dplyr)

# Using the built-in 'iris' dataset
iris %>%
  group_by(Species) %>%
  summarise(
    Count        = n(),
    Mean_Length  = round(mean(Sepal.Length), 2),
    Mean_Width   = round(mean(Sepal.Width), 2),
    SD_Length    = round(sd(Sepal.Length), 2)
  ) %>%
  arrange(desc(Mean_Length))
```

## Machine Learning and Predictive Modeling

R has robust tools for machine learning, from classical models to modern frameworks like `tidymodels` and `caret`.

```{r machine-learning}
# Simple linear regression on the built-in 'mtcars' dataset
model <- lm(mpg ~ wt + hp + cyl, data = mtcars)

# Model summary
summary(model)

# Predictions vs actual values
predictions <- predict(model, mtcars)
comparison <- data.frame(
  Actual    = mtcars$mpg,
  Predicted = round(predictions, 2),
  Residual  = round(mtcars$mpg - predictions, 2)
)
head(comparison, 8)
```

## Reproducible Reports with R Markdown

The file you are reading right now **is** an R Markdown document — a living example of this application. R Markdown allows you to:

- Mix narrative text with executable R code
- Generate outputs in HTML, PDF, Word, and more
- Build dashboards (with `flexdashboard`) and interactive apps (with `Shiny`)
- Automate the full pipeline from raw data to final report

This makes your analyses fully **transparent, reproducible, and shareable**.

## Other Notable Application Domains

| Domain | Example use cases |
|---|---|
| **Bioinformatics** | Genomics, RNA-seq analysis (Bioconductor packages) |
| **Econometrics** | Time series, panel data, financial modeling |
| **Social sciences** | Survey analysis, text mining, network analysis |
| **Public health & epidemiology** | Disease surveillance, survival analysis |
| **Geospatial analysis** | Maps and spatial data with `sf`, `leaflet`, `tmap` |
| **Web scraping** | Collecting web data with `rvest` |
| **Interactive dashboards** | Building web apps with `Shiny` |

---

# Summary

Let's recap what was covered in this lecture:

- ✅ **R** is a free, open-source language built for statistical computing and data analysis.
- ✅ **RStudio** is an IDE that makes working with R much more productive through its four-pane layout, project management, and built-in tools.
- ✅ The **four panes** of RStudio serve distinct purposes: Source Editor, Console, Environment/History, and Files/Plots/Packages/Help.
- ✅ R is used across many domains including **statistics, visualization, machine learning, bioinformatics, economics**, and **reproducible reporting**.

---

# Exercises

Try the following on your own to consolidate what you have learned:

1. Open RStudio and confirm your R version using `R.version$version.string` in the console.
2. Create a variable `my_age` and assign your age to it. Then multiply it by 2.
3. Create a vector of 5 country names and display it.
4. Use the `summary()` function on the built-in `iris` dataset (`summary(iris)`).
5. Create a new RStudio Project for this course in a folder on your computer.

---

# Further Reading & Resources

- 📘 [R for Data Science (free online book)](https://r4ds.hadley.nz/) — Hadley Wickham & Garrett Grolemund
- 📘 [The R Markdown Cookbook](https://bookdown.org/yihui/rmarkdown-cookbook/) — Yihui Xie
- 🌐 [CRAN — The R Project](https://cran.r-project.org)
- 🌐 [RStudio Cheat Sheets](https://posit.co/resources/cheatsheets/) — Quick reference cards for key packages
- 🌐 [Stack Overflow — R tag](https://stackoverflow.com/questions/tagged/r) — Community Q&A

---

```{r session-info, echo=FALSE}
cat("Session Info\n")
cat("============\n")
sessionInfo()
```
