---
title: "Operators in R"
subtitle: "A Comprehensive Lecture for Beginners"
author: "Ange Maxime TCHOUTANG"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "##")
```

---

# Introduction

In R, **operators** are special symbols or keywords used to perform operations on values and variables. They are the fundamental building blocks of any R program — you use them constantly, whether you are doing a simple calculation, comparing values, or filtering data.

In this lecture, we will cover the four main families of operators in R:

1. **Assignment operators** — to store values in variables
2. **Arithmetic operators** — to do mathematical calculations
3. **Relational (comparison) operators** — to compare values
4. **Logical operators** — to combine conditions and work with Boolean values

We will also explore how operators interact with **vectors**, and introduce two very useful tools: **Boolean selection** and the **`ifelse()` function**.

> 💡 **Good practice:** Before starting a new session or tutorial, clean your Global Environment to avoid clutter:
> ```r
> rm(list = ls())
> ```

---

# Assignment Operators

Before we can use operators in any meaningful way, we need to be able to store values. In R, this is done with **assignment operators**.

R offers three equivalent ways to assign a value to a variable:

| Operator | Direction | Example |
|----------|-----------|---------|
| `<-` | Left arrow (recommended) | `x <- 10` |
| `=` | Equals sign | `x = 10` |
| `->` | Right arrow | `10 -> x` |

When using the arrow, make sure it points **from the value toward the variable**.

```{r assignment_operators}
# Using the left arrow (most common and recommended)
a <- 12

# Using the equals sign
b = "magician"

# Using the right arrow (value on the left, variable on the right)
32 + 19 -> c

# Display the variables
a
b
c
```

> ⚠️ **Important:** When you assign a new value to an existing variable, the old value is **permanently overwritten**.

```{r overwrite_variable}
a         # current value: 12
a <- 52344
a         # new value: 52344
```

---

# Arithmetic Operators

Arithmetic operators perform mathematical computations. R can be used just like a calculator.

| Operator | Operation | Example | Result |
|----------|-----------|---------|--------|
| `+` | Addition | `5 + 3` | `8` |
| `-` | Subtraction | `5 - 3` | `2` |
| `*` | Multiplication | `5 * 3` | `15` |
| `/` | Division | `5 / 2` | `2.5` |
| `^` | Exponentiation (power) | `2 ^ 3` | `8` |
| `%%` | Modulo (remainder) | `10 %% 6` | `4` |
| `%/%` | Integer division | `10 %/% 6` | `1` |

> 💡 **The modulo `%%`** returns the **remainder** of a division. For example, `10 %% 6 = 4` because 6 goes into 10 once, leaving a remainder of 4.

```{r arithmetic_operators}
# Basic arithmetic
2 + 2       # Addition
10 - 4      # Subtraction
3 * 7       # Multiplication
10 / 4      # Division
2 ^ 8       # Exponentiation (2 to the power of 8)

# Integer division and modulo (Euclidean division)
10 %/% 6    # Quotient: how many times 6 fits in 10
10 %% 6     # Remainder: what is left over
```

## Arithmetic with Variables

You can perform arithmetic directly on variables:

```{r arithmetic_with_variables}
a <- 2
b <- 4

a + b   # sum
a - b   # difference
a * b   # product
a / b   # division
a ^ b   # a raised to the power of b
```

## Arithmetic with Vectors

Operators in R are **vectorized**, meaning they apply element-by-element when used with vectors.

```{r arithmetic_vectors}
v1 <- c(3, 54, 1, 8, 10)

v1 * 3    # Multiply every element by 3
v1 - 5    # Subtract 5 from every element
```

When performing operations between two vectors, the lengths must be **compatible** (one must be a divisor of the other):

```{r vector_arithmetic}
v0 <- c(1, 2, 3, 4, 5)   # length 5
v4 <- c(1, 0)              # length 2 — a divisor of 10, but not of 5
v5 <- c(1, 1, 1, 1)       # length 4

# This works: v0 (length 5) and v4 (length 2) — 
# R will recycle v4, but with a warning since 5 is not a multiple of 2
v4 + v0

# This does NOT work: length 4 and length 5 have no common divisor
# v5 + v0  # would produce a warning
```

---

# Relational (Comparison) Operators

Relational operators compare two values and return a **Boolean** result: either `TRUE` or `FALSE`.

| Operator | Meaning | Example | Result |
|----------|---------|---------|--------|
| `<` | Less than | `2 < 3` | `TRUE` |
| `>` | Greater than | `2 > 3` | `FALSE` |
| `<=` | Less than or equal to | `3 <= 3` | `TRUE` |
| `>=` | Greater than or equal to | `5 >= 6` | `FALSE` |
| `==` | Equal to | `4 == 4` | `TRUE` |
| `!=` | Not equal to | `4 != 5` | `TRUE` |

```{r relational_operators}
2 > 3     # Is 2 greater than 3? FALSE
1 < 100   # Is 1 less than 100? TRUE
5 == 5    # Is 5 equal to 5? TRUE
5 != 3    # Is 5 different from 3? TRUE
4 >= 4    # Is 4 greater than or equal to 4? TRUE
3 <= 2    # Is 3 less than or equal to 2? FALSE
```

> ⚠️ **Common mistake:** Do not confuse `=` (assignment) with `==` (equality test). Use `==` to compare two values.

## Relational Operators on Vectors

When applied to a vector, relational operators return a **vector of Booleans**, one for each element:

```{r relational_on_vectors}
v0 <- 1:10
v0

v0 > 5    # Which elements are strictly greater than 5?
```

Notice that `5` itself returns `FALSE` because `>` is **strict** (not ≥). To include 5, use `>=`.

---

# Logical Operators

Logical operators work with Boolean values (`TRUE` / `FALSE`) to combine or negate conditions.

| Operator | Meaning | Notes |
|----------|---------|-------|
| `!` | NOT (negation) | Inverts TRUE to FALSE and vice versa |
| `&` | AND (element-wise) | Use with vectors |
| `&&` | AND (single value) | Use with scalars only |
| `\|` | OR (element-wise) | Use with vectors |
| `\|\|` | OR (single value) | Use with scalars only |

```{r logical_operators}
# Is 10 AND 5 both greater than 3?
10 & 5 > 3    # TRUE — both conditions are met

# Is 10 AND 1 both greater than 3?
10 & 1 > 3    # FALSE — 1 is not greater than 3

# Is 10 OR 1 greater than 3?
10 | 1 > 3    # TRUE — at least one condition is met
```

> 💡 **Tip:** Use `&` and `|` when working with **vectors** (element-by-element). Use `&&` and `||` only for single scalar comparisons (e.g., inside `if` statements).

---

# Booleans in R

A **Boolean** (or logical) value is a special data type in R that can only be `TRUE` or `FALSE`. These are written in **uppercase**.

```{r booleans_intro}
TRUE
FALSE
```

## Booleans as Numbers

Booleans can be treated as numbers: `TRUE` equals `1` and `FALSE` equals `0`. This makes them very useful for counting and arithmetic.

```{r booleans_as_numbers}
as.numeric(TRUE)   # returns 1
as.numeric(FALSE)  # returns 0

TRUE + TRUE        # 1 + 1 = 2
FALSE - 10         # 0 - 10 = -10
```

## Boolean Selection in Vectors

One of the most powerful uses of Booleans is **indexing**: you can use a logical vector to select elements from another vector. Elements corresponding to `TRUE` are kept; those corresponding to `FALSE` are dropped.

```{r boolean_selection}
v2 <- c(10, 20)
v2

v2[c(TRUE, FALSE)]   # Keep only the first element
v2[c(FALSE, TRUE)]   # Keep only the second element
v2[c(FALSE, FALSE)]  # Keep nothing (empty result)
v2[c(TRUE, TRUE)]    # Keep everything
```

---

# Logical Selection in Vectors

Combining **relational operators** with **Boolean indexing** is the core of data filtering in R. The idea is simple:

1. Apply a condition to a vector → get a Boolean vector
2. Use that Boolean vector to select elements

```{r logical_selection}
v3 <- 1:10
v4 <- c("a", "b", "c", "d", "e", "f", "g", "h", "i", "j")

# Step 1: Generate the Boolean vector
v3 >= 5

# Step 2: Use it to filter the numeric vector
v3[v3 >= 5]

# Step 3: Use the same Boolean to filter a character vector of the same length
v4[v3 >= 5]

# Step 4: Combine conditions with OR
v4[v3 >= 5 | v4 == "b"]   # Keep elements where v3 >= 5 OR the letter is "b"
```

## The `%in%` Operator

The `%in%` operator checks whether each element of a vector **belongs to** a specified set of values. It returns a Boolean vector.

```{r in_operator}
v3 <- 1:10
v4 <- 1:4

# Is 5 in v3?
5 %in% v3    # TRUE — 5 is in 1:10

# Is 5 in v4?
5 %in% v4    # FALSE — v4 only goes up to 4

# Which elements of v3 are also in v4?
v3 %in% v4
```

---

# The `ifelse()` Function

The `ifelse()` function applies a conditional transformation to a vector. It works element-by-element and takes three arguments:

```
ifelse(condition, value_if_TRUE, value_if_FALSE)
```

```{r ifelse_function}
v3 <- 1:10
v3

# Replace the value 10 with "equal to 10", otherwise "not equal to 10"
ifelse(v3 == 10, "equal to 10", "not equal to 10")

# Numeric transformation: multiply by 10 if less than 10, otherwise divide by 2
ifelse(v3 < 10, v3 * 10, v3 / 2)
```

This is extremely useful for creating derived variables or recoding data — and it will be even more powerful once you are comfortable with data frames.

---

# The Assignment Operator in Context

To complete the picture of operators, it is worth noting the **colon operator `:`**, which generates integer sequences. Although it is not usually listed alongside arithmetic operators, it behaves like one.

```{r colon_operator}
1:10        # Sequence from 1 to 10
5:-2        # Descending sequence from 5 to -2
```

For more flexible sequences, use `seq()`:

```{r seq_function}
seq(3, 11, by = 4)       # From 3 to 11, step 4
seq(23, 9, by = -5)      # Descending, step -5
seq(from = 1, to = 10, by = 1)  # Full explicit version
```

---

# Summary Table

Here is a complete summary of all operators covered in this lecture:

| Family | Operator | Meaning |
|--------|----------|---------|
| **Assignment** | `<-`, `=`, `->` | Store a value in a variable |
| **Arithmetic** | `+`, `-`, `*`, `/` | Basic math |
| **Arithmetic** | `^` | Exponentiation |
| **Arithmetic** | `%%` | Modulo (remainder) |
| **Arithmetic** | `%/%` | Integer division |
| **Relational** | `<`, `>` | Strictly less/greater than |
| **Relational** | `<=`, `>=` | Less/greater than or equal |
| **Relational** | `==` | Equal to |
| **Relational** | `!=` | Not equal to |
| **Logical** | `!` | NOT (negation) |
| **Logical** | `&`, `&&` | AND |
| **Logical** | `\|`, `\|\|` | OR |
| **Membership** | `%in%` | Element belongs to a set |
| **Sequence** | `:` | Integer sequence |

---

# Practice Exercises

Test your understanding with these exercises. Try them in your R console before looking at the solutions.

**Exercise 1.** Create two variables `x <- 7` and `y <- 3`. Compute their sum, difference, product, and the remainder when `x` is divided by `y`.

**Exercise 2.** Create a vector `scores <- c(45, 78, 92, 33, 67, 88, 55)`. Use a relational operator to extract only the scores that are strictly greater than 60.

**Exercise 3.** Using the same `scores` vector, use `ifelse()` to create a new vector where scores above 60 are labeled `"pass"` and scores 60 or below are labeled `"fail"`.

**Exercise 4.** Create a vector `fruits <- c("apple", "banana", "mango", "papaya", "kiwi")`. Use `%in%` to check which fruits are in the list `c("banana", "kiwi", "grape")`.

**Exercise 5.** What is the difference between `&` and `&&`? In what situation would you use each?

---

```{r cleanup, include=FALSE}
rm(list = ls())
```

---

*End of Lecture — Operators in R*
