---
title: "Lecture 4: Identifying and Handling Outliers in Exploratory Data Analysis"
author: "Ange Maxime TCHOUTANG"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 5
)
```

---

# Learning Objectives

By the end of this lecture, you will be able to:

- Define what an **outlier** is and why identifying outliers matters
- Use **visual methods** (boxplots, histograms, scatter plots) to spot outliers
- Apply the **Grubbs test** to formally identify outliers in a univariate dataset
- Write a **custom R function** that iteratively flags multiple outliers
- Visualise outlier positions within a histogram using `ggplot2`
- Make informed decisions about what to **do** with identified outliers

---

# What is an Outlier?

An **outlier** is an observation that lies an abnormal distance from the other observations in a dataset. Outliers can arise from:

- **Data entry errors** (e.g., a weight recorded as 580 kg instead of 58 kg)
- **Measurement errors** (malfunctioning sensor, instrument drift)
- **Genuine extreme biological or physical events** (rare but real)

> **Why does it matter?** Outliers can heavily distort the mean, inflate standard deviation, bias regression coefficients, and violate model assumptions.

## A Conceptual Example

```{r toy-example}
x <- c(12, 14, 13, 15, 14, 13, 12, 14, 99)

cat("Mean without outlier:", mean(x[-9]), "\n")
cat("Mean with outlier:   ", mean(x),     "\n")
cat("SD without outlier:", round(sd(x[-9]), 2), "\n")
cat("SD with outlier:   ", round(sd(x),    2), "\n")
```

> A single extreme value (99) more than **doubles** the standard deviation and shifts the mean by nearly 10 units.

---

# Required Packages

```{r load-packages}
# Install the outliers package if needed:
# install.packages("outliers")

library(outliers)   # Statistical outlier tests (Grubbs, Dixon, etc.)
library(ggplot2)    # Data visualisation
```

---

# Visual Detection of Outliers

Before applying any statistical test, always **look at your data**. Visualisation is the first and most intuitive step.

## Working with the `airquality` Dataset

R's built-in `airquality` dataset records daily air quality measurements in New York (May–September 1973). It contains variables: `Ozone`, `Solar.R`, `Wind`, `Temp`, `Month`, `Day`.

```{r airquality-preview}
head(airquality)
summary(airquality)
```

## Boxplot

Boxplots show the IQR and flag points beyond 1.5×IQR as potential outliers:

```{r boxplot-ozone}
ggplot(airquality, aes(y = Ozone)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7, outlier.color = "red", outlier.size = 3) +
  labs(
    title   = "Distribution of Ozone levels — New York 1973",
    subtitle = "Red points are flagged as potential outliers by the IQR rule",
    y       = "Ozone (ppb)"
  ) +
  theme_minimal()
```

## Histogram

A histogram reveals the shape of the distribution and shows isolated bars at extreme values:

```{r histogram-ozone}
ggplot(airquality, aes(x = Ozone)) +
  geom_histogram(bins = 20, fill = "steelblue", color = "white", alpha = 0.8) +
  labs(
    title = "Histogram of Ozone — New York 1973",
    x     = "Ozone (ppb)",
    y     = "Count"
  ) +
  theme_minimal()
```

## Scatter Plot (Two Variables)

For bivariate outliers, a scatter plot is the most effective tool:

```{r scatter-wind-temp}
ggplot(airquality, aes(x = Wind, y = Temp)) +
  geom_point(color = "steelblue", alpha = 0.7, size = 2.5) +
  geom_smooth(method = "lm", se = TRUE, color = "red", linetype = "dashed") +
  labs(
    title   = "Wind speed vs Temperature — New York 1973",
    subtitle = "Points far from the regression line may be bivariate outliers",
    x       = "Wind speed (mph)",
    y       = "Temperature (°F)"
  ) +
  theme_minimal()
```

---

# The IQR Rule for Outlier Detection

The **Interquartile Range (IQR) rule** is the most common classical method. A value is flagged as an outlier if it falls outside:

$$\text{Lower fence} = Q_1 - 1.5 \times IQR$$
$$\text{Upper fence} = Q_3 + 1.5 \times IQR$$

```{r iqr-rule}
ozone_clean <- airquality$Ozone[!is.na(airquality$Ozone)]

Q1  <- quantile(ozone_clean, 0.25)
Q3  <- quantile(ozone_clean, 0.75)
IQR_val <- IQR(ozone_clean)

lower_fence <- Q1 - 1.5 * IQR_val
upper_fence <- Q3 + 1.5 * IQR_val

cat("Q1 =", Q1, "\n")
cat("Q3 =", Q3, "\n")
cat("IQR =", IQR_val, "\n")
cat("Lower fence:", lower_fence, "\n")
cat("Upper fence:", upper_fence, "\n")

# Identify outlier values
outlier_values <- ozone_clean[ozone_clean < lower_fence | ozone_clean > upper_fence]
cat("Outlier values (IQR rule):", outlier_values, "\n")
```

---

# The Grubbs Test for Formal Outlier Detection

## What is the Grubbs Test?

The **Grubbs test** (also called the Extreme Studentized Deviate test) is a formal statistical test for identifying a **single outlier** in a univariate dataset that is assumed to follow a **normal distribution**.

**Hypotheses:**

- $H_0$: There are **no outliers** in the dataset
- $H_1$: There is **exactly one outlier** (the most extreme value)

**Test statistic:**

$$G = \frac{\max|x_i - \bar{x}|}{s}$$

where $\bar{x}$ is the sample mean and $s$ is the sample standard deviation.

If the p-value < 0.05, we reject $H_0$ and conclude that the most extreme value is an outlier.

## Applying a Single Grubbs Test

```{r single-grubbs}
ozone_clean <- airquality$Ozone[!is.na(airquality$Ozone)]

grubbs_result <- grubbs.test(ozone_clean)
print(grubbs_result)
```

**Interpretation:** If `p < 0.05`, the most extreme value (identified in the `alternative` field) is a statistically significant outlier.

---

# Writing a Custom Iterative Outlier Detection Function

The standard `grubbs.test()` identifies only **one outlier at a time**. To find **all outliers** in a dataset, we apply the test iteratively — removing each identified outlier and repeating until no more outliers are found.

```{r grubbs-flag-function}
grubbs.flag <- function(x) {
  
  # Initialise an empty vector to collect all identified outliers
  outliers <- NULL
  
  # Work on a copy of the data
  test <- x
  
  # Run the first Grubbs test
  grubbs_result <- grubbs.test(test)
  pv <- grubbs_result$p.value
  
  # Iteratively identify and remove outliers while p < 0.05
  while (pv < 0.05) {
    
    # Extract the outlier value from the alternative hypothesis text
    outlier_val <- as.numeric(
      strsplit(grubbs_result$alternative, " ")[[1]][3]
    )
    
    # Add this outlier to our collection
    outliers <- c(outliers, outlier_val)
    
    # Remove the identified outlier from the test vector
    test <- x[!x %in% outliers]
    
    # Re-run the Grubbs test on the cleaned vector
    grubbs_result <- grubbs.test(test)
    pv <- grubbs_result$p.value
  }
  
  # Return a data frame with the original values and a logical outlier flag
  return(data.frame(
    X        = x,
    outlier  = (x %in% outliers)
  ))
}
```

**Let's walk through this function step by step:**

| Line | What it does |
|---|---|
| `outliers <- NULL` | Creates an empty container for identified outliers |
| `grubbs.test(test)` | Runs the Grubbs test on the current data |
| `while (pv < 0.05)` | Keeps looping as long as outliers are found |
| `strsplit(...)` | Parses the outlier value from the test result text |
| `test <- x[!x %in% outliers]` | Removes all identified outliers before re-testing |
| `return(data.frame(...))` | Returns the original data with a TRUE/FALSE outlier column |

---

# Applying the Function to Real Data

## Using the UCI Air Quality Dataset

```{r load-airdata, eval=FALSE}
# If you have the UCI Air Quality Excel file:
# library(readxl)
# Air_data <- read_xlsx("AirqualityUCI.xlsx")
```

For demonstration, we use the built-in `airquality` dataset (a standard R dataset with similar variables):

```{r apply-grubbs}
# Use Ozone variable from built-in airquality (remove NAs first)
ozone_vals <- airquality$Ozone[!is.na(airquality$Ozone)]

# Apply the custom function
result_df <- grubbs.flag(ozone_vals)

# View the identified outliers
cat("Total observations:", nrow(result_df), "\n")
cat("Outliers detected:", sum(result_df$outlier), "\n")
cat("Outlier values:", result_df$X[result_df$outlier], "\n")
```

---

# Visualising Outliers in a Histogram

The key advantage of the `grubbs.flag()` function is that it returns a data frame with a logical `outlier` column, which we can map directly to colour in `ggplot2`:

```{r histogram-outliers}
ggplot(result_df,
       aes(x     = X,
           color = outlier,
           fill  = outlier)) +
  geom_histogram(
    binwidth = diff(range(result_df$X)) / 30,
    alpha    = 0.7
  ) +
  scale_fill_manual(
    values = c("FALSE" = "steelblue", "TRUE" = "firebrick"),
    labels = c("Normal", "Outlier")
  ) +
  scale_color_manual(
    values = c("FALSE" = "steelblue", "TRUE" = "firebrick"),
    labels = c("Normal", "Outlier")
  ) +
  labs(
    title    = "Ozone levels with Grubbs-identified outliers highlighted",
    subtitle = "Red bars mark values statistically identified as outliers (p < 0.05)",
    x        = "Ozone (ppb)",
    y        = "Count",
    fill     = "Status",
    color    = "Status"
  ) +
  theme_minimal() +
  theme(legend.position = "top")
```

---

# What to Do With Outliers?

Identifying outliers is only the first step. What you do next depends on the **source** and **context**:

## Decision Framework

```
Outlier detected
      |
      ├─── Is it a data entry / recording error?
      │         YES → Correct or remove it
      │
      ├─── Is it a genuine extreme value?
      │         YES → Keep it; report it; consider robust methods
      │
      ├─── Is it due to a different sub-population?
      │         YES → Stratify your analysis
      │
      └─── Are you unsure?
                → Run analysis both with and without; report both results
```

## Practical Options

| Approach | When to use |
|---|---|
| **Remove** | Confirmed data entry errors |
| **Keep but report** | Genuine rare events |
| **Winsorize** | Replace with the nearest non-outlier value |
| **Robust statistics** | Use median, IQR, or robust regression instead of mean/OLS |
| **Log-transform** | When data are right-skewed with extreme values |

## Winsorization Example

```{r winsorize}
# Winsorize at the 5th and 95th percentile
lower_p <- quantile(ozone_vals, 0.05)
upper_p <- quantile(ozone_vals, 0.95)

ozone_winsorized <- pmin(pmax(ozone_vals, lower_p), upper_p)

cat("Original max:", max(ozone_vals), "\n")
cat("Winsorized max:", max(ozone_winsorized), "\n")
```

---

# Comparing Multiple Outlier Detection Methods

```{r compare-methods}
# Z-score method (flag values with |z| > 3)
z_scores <- scale(ozone_vals)
zscore_outliers <- abs(z_scores) > 3

# IQR method
iqr_outliers <- ozone_vals < lower_fence | ozone_vals > upper_fence

# Grubbs method (from earlier)
grubbs_outliers <- result_df$outlier

comparison <- data.frame(
  Value   = ozone_vals,
  Z_score = as.vector(zscore_outliers),
  IQR     = iqr_outliers,
  Grubbs  = grubbs_outliers
)

# Show values flagged by at least one method
comparison[comparison$Z_score | comparison$IQR | comparison$Grubbs, ]
```

> **Key insight:** Different methods may flag different observations. When methods agree, you have stronger evidence. When they disagree, investigate further.

---

# Summary

| Concept | Key Takeaway |
|---|---|
| Outlier definition | A value abnormally distant from others — may be error or genuine extreme |
| Visual methods | Boxplot, histogram, scatter plot — always look first |
| IQR rule | Simple, robust; flags values beyond Q1 − 1.5×IQR or Q3 + 1.5×IQR |
| Grubbs test | Formal statistical test; assumes normality; tests one outlier at a time |
| Iterative Grubbs | Custom function to detect all outliers; returns a flagging data frame |
| ggplot2 visualisation | Map the outlier flag to fill/color to produce a labelled histogram |
| What to do | Depends on source: correct errors, keep genuine extremes, use robust methods |

---

# Practice Exercises

1. Apply `grubbs.flag()` to the `Wind` column of the `airquality` dataset. How many outliers are found? Visualise the result with a histogram.

2. Compare outlier detection results for `Temp` in `airquality` using three methods: the IQR rule, the Z-score method (|z| > 3), and the Grubbs iterative function. Do the methods agree?

3. Create a scatter plot of `Solar.R` vs `Ozone` from `airquality`. Colour points identified as outliers by `grubbs.flag()` on the `Ozone` axis in red. What do you observe?

4. *Challenge:* Modify `grubbs.flag()` so that it tests for outliers on **both tails** and **one tail at a time** (use `two.sided = FALSE` in `grubbs.test()`). How does this affect the results?

---

# References

- Grubbs, F. E. (1969). Procedures for detecting outlying observations in samples. *Technometrics*, 11(1), 1–21.
- Wickham, H. & Grolemund, G. (2017). *R for Data Science*. O'Reilly Media. https://r4ds.had.co.nz
- R `outliers` package documentation: https://cran.r-project.org/package=outliers
- Barnett, V. & Lewis, T. (1994). *Outliers in Statistical Data* (3rd ed.). Wiley.
