---
title: "Data Manipulation in R"
subtitle: "A Comprehensive Lecture — From Raw Data to Clean, Tidy, Analysis-Ready Tables"
author: "Ange Maxime TCHOUTANG"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    theme: flatly
    highlight: tango
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  comment = "##",
  message = FALSE,
  warning = FALSE
)
```

```{r packages}
library(tidyverse)
library(janitor)
library(lubridate)
```

---

# Introduction — The Seven Steps of Statistical Analysis

Every rigorous data analysis follows a structured workflow. Understanding the whole pipeline helps you know *why* each manipulation step matters:

| Step | Description |
|---|---|
| **1. Research plan** | Define your research question, hypotheses (null + alternative), and expected results |
| **2. Data preparation** | Identify suitable data sources; import raw data into R |
| **3. Data cleaning** | Select variables and population of interest; recode variables; handle missing values; save clean data |
| **4. Exploratory analysis** | Explore distributions and relationships; detect anomalies; verify cleaning |
| **5. Descriptive analysis** | Describe each variable of interest in depth (univariate + bivariate) |
| **6. Inferential analysis** | Test hypotheses; measure strength and direction of relationships using regression models |
| **7. Presentation** | Export tables and plots to reports or presentations |

> 💡 Data cleaning (Step 3) is the most time-consuming step and directly determines the validity of everything that follows. Be patient, systematic, and leave a traceable code trail.

---

# Creating a Dataset to Work With

We will use two datasets throughout this lecture:

```{r create-datasets}
# --- Dataset 1: Accident / driving permit data (small, illustrative) ---
data_acc <- data.frame(
  Prenom.Nom       = c("Udo Adesina","Flora Feijo","Svend Baardsen","Edda Franck","Yeo Ji-tae"),
  sexe             = c("Femme","Femme","Homme","Femme","Homme"),
  Votre.Age        = c(196, 22, 19, 18, 21),    # 196 is a data entry error
  Permis           = c("oui","Non","Oui","Non","Non"),
  accident.condu   = c(0.01, NA, 0.02, NA, NA),
  accident.on.ondu = c(0.02, 0.01, NA, 0.02, NA)
)

# --- Dataset 2: Regional malaria data (analysis dataset) ---
set.seed(123)
regions_full <- c("Centre","Littoral","Far_North","North_West","South_West")
months_full  <- month.abb[1:5]

malaria_big <- expand.grid(region = regions_full,
                           month  = months_full,
                           year   = 2024)
malaria_big$cases <- sample(100:1500, nrow(malaria_big), replace = TRUE)
malaria_big$rain  <- sample(0:350,   nrow(malaria_big), replace = TRUE)

head(data_acc)
head(malaria_big)
```

---

# Step 1 — Inspecting Raw Data

Before touching anything, **understand what you have**.

```{r inspect}
# Dimensions
dim(data_acc)
dim(malaria_big)

# Variable names and types
str(data_acc)
str(malaria_big)

# Compact tidyverse view
dplyr::glimpse(malaria_big)

# Statistical summary
summary(data_acc)
summary(malaria_big)

# First and last rows
head(malaria_big)
tail(malaria_big, 3)

# Column names
names(data_acc)
names(malaria_big)
```

> 💡 **Always call `summary()` immediately** after loading any dataset. It reveals the range of each variable, the number of NAs, and whether numeric columns were accidentally read as characters.

---

# Step 2 — Cleaning Column Names

Messy column names (spaces, capital letters, special characters) cause errors downstream.

## Base R — `make.names()` and `tolower()`

```{r clean-names-base}
# Current names
names(data_acc)

# Make syntactically valid (replaces spaces and special chars with .)
make.names(names(data_acc))

# Lowercase all names
names(data_acc) <- tolower(names(data_acc))
names(data_acc)
```

## `janitor::clean_names()` — Recommended

```{r clean-names-janitor}
library(janitor)

# Converts to lowercase snake_case automatically
data_clean <- clean_names(data_acc)
names(data_clean)
```

`clean_names()` removes spaces, replaces special characters with `_`, converts everything to lowercase, and handles duplicates — all in one call.

---

# Step 3 — Checking and Converting Variable Types

Each variable must have the correct R type before analysis. Always verify with `class()` or `str()`.

## Type Reference Table

| R type | `class()` output | What it stores | Example |
|---|---|---|---|
| Numeric | `"numeric"` | Decimal numbers | `3.14`, `height` |
| Integer | `"integer"` | Whole numbers | `5L`, counts |
| Character | `"character"` | Text strings | `"Male"` |
| Factor | `"factor"` | Categorical (unordered) | `sex`, `blood_type` |
| Ordered factor | `"ordered" "factor"` | Categorical (ordered) | `severity`, `education` |
| Logical | `"logical"` | TRUE/FALSE | `is_positive` |
| Date | `"Date"` | Calendar dates | `2024-01-15` |

## Checking Types

```{r check-types}
class(data_clean$votre_age)
class(data_clean$sexe)
is.numeric(data_clean$votre_age)
is.character(data_clean$sexe)
```

## Converting Types — the `as.*()` Family

```{r convert-types}
# Character → Factor
data_clean$sexe   <- as.factor(data_clean$sexe)
data_clean$permis <- as.factor(data_clean$permis)

# Character → Numeric
x_chr <- c("3.5", "7.2", "1.1")
as.numeric(x_chr)

# Numeric → Integer
as.integer(c(3.7, 5.9))   # truncates — does NOT round

# Factor → Character
as.character(data_clean$sexe)

# Logical coercion
as.numeric(TRUE)    # 1
as.numeric(FALSE)   # 0
```

> ⚠️ `as.integer()` **truncates** decimals (3.7 → 3), it does not round. Use `round()` first if you need rounding.

---

# Step 4 — Working with Factors

Factors are R's data structure for categorical variables. They are essential for statistical modelling and controlled plotting.

## Creating and Inspecting Factors

```{r factors-create}
sexe_f <- factor(c("Femme","Homme","Femme","Homme","Femme"))
sexe_f
levels(sexe_f)   # alphabetical by default
nlevels(sexe_f)
table(sexe_f)    # frequency count
```

## Renaming Levels

```{r factor-rename}
# Method 1 — via labels in factor()
factor(c("F","M","F"), labels = c("Femme","Homme"))

# Method 2 — reassign levels() directly
levels(sexe_f) <- c("Female","Male")
sexe_f
```

## Reordering Levels

```{r factor-reorder}
# Method 1: specify order explicitly
factor(sexe_f, levels = c("Male","Female"))

# Method 2: relevel() — set the reference category (crucial for regression)
relevel(sexe_f, ref = "Male")

# Method 3: reorder() by a numeric variable (great for plots)
taille <- c(165, 178, 162, 182, 170)
reorder(sexe_f, taille)           # ascending
reorder(sexe_f, taille, decreasing = TRUE)  # descending

# Method 4: forcats::fct_infreq() — order by frequency
library(forcats)
fct_infreq(sexe_f)
```

## Ordered Factors

For ordinal variables (variables with a meaningful order):

```{r ordered-factor}
severity <- factor(c("Low","High","Medium","Low","High"),
                   levels  = c("Low","Medium","High"),
                   ordered = TRUE)
severity
severity[1] < severity[3]   # TRUE — comparisons work
```

---

# Step 5 — Working with Dates

R's `Date` class stores calendar dates. The `lubridate` package makes date parsing effortless.

```{r dates}
library(lubridate)

# Parse dates from different string formats
ymd("2008-11-14")           # Year-Month-Day
dmy("14.11.2008")           # Day-Month-Year
dmy("8th of November 2008") # Natural language

# Today's date
today()
Sys.Date()

# Date arithmetic
birth <- ymd("1990-12-25")
age_days  <- today() - birth
age_years <- as.numeric(age_days) / 365.25
round(age_years, 1)

# Extract components
year(today())
month(today())
day(today())
wday(today(), label = TRUE)   # day of the week
```

---

# Step 6 — Selecting Variables and Filtering Rows

## Base R Selection

```{r base-selection}
# Select specific columns
malaria_big[, c("region","cases")]     # by name
malaria_big[, c(1, 4)]                 # by position
malaria_big[, -3]                      # all except column 3

# Filter rows by condition
malaria_big[malaria_big$cases > 1000, ]
malaria_big[malaria_big$region == "Far_North", ]

# Combined row + column selection
malaria_big[malaria_big$cases > 500, c("region","month","cases")]
```

## `dplyr` Selection

```{r dplyr-select-filter}
library(dplyr)

# select() — choose columns
malaria_big %>% select(region, cases, rain)
malaria_big %>% select(-year)                     # drop year
malaria_big %>% select(starts_with("r"))          # helper functions

# filter() — keep matching rows
malaria_big %>% filter(cases > 1000)
malaria_big %>% filter(region == "Far_North" & cases > 800)
malaria_big %>% filter(region %in% c("Centre","Littoral"))
malaria_big %>% filter(cases > 500 | rain < 50)

# Chained: select + filter in one pipeline
malaria_big %>%
  select(region, month, cases) %>%
  filter(cases > 1000) %>%
  arrange(desc(cases))
```

### `select()` Helper Functions

```{r select-helpers}
# All columns containing "a"
malaria_big %>% select(contains("a"))

# Columns ending with "s"
malaria_big %>% select(ends_with("s"))

# Numeric columns only
malaria_big %>% select(where(is.numeric))
```

---

# Step 7 — Renaming Columns

```{r rename}
# Base R
names(malaria_big)[names(malaria_big) == "cases"] <- "malaria_cases"
names(malaria_big)

# Restore
names(malaria_big)[names(malaria_big) == "malaria_cases"] <- "cases"

# dplyr::rename() — new_name = old_name
malaria_big %>%
  rename(n_cases  = cases,
         rainfall = rain) %>%
  head(3)
```

---

# Step 8 — Creating and Modifying Columns with `mutate()`

`mutate()` adds new columns or overwrites existing ones. All other columns are preserved.

## Basic `mutate()`

```{r mutate-basic}
# Create a new column
malaria_big %>%
  mutate(cases_per_100k = cases / 10) %>%
  head(4)

# Create multiple columns at once
malaria_big %>%
  mutate(
    cases_log   = log1p(cases),           # log(cases + 1)
    high_burden = cases > 800,            # logical flag
    .before     = cases                   # position the new cols
  ) %>%
  head(4)
```

## Conditional Column Creation — `ifelse()` and `case_when()`

```{r mutate-conditional}
# ifelse() — two outcomes
malaria_big %>%
  mutate(burden = ifelse(cases > 800, "High", "Low")) %>%
  head(4)

# case_when() — multiple outcomes (cleaner than nested ifelse)
malaria_big %>%
  mutate(burden = case_when(
    cases > 1200 ~ "Very High",
    cases > 800  ~ "High",
    cases > 400  ~ "Moderate",
    TRUE         ~ "Low"        # default
  )) %>%
  count(burden)
```

## `recode()` — Recode Values

```{r recode}
# Rename specific values
malaria_big %>%
  mutate(month = recode(month,
                        Jan = "January",
                        Feb = "February",
                        Mar = "March")) %>%
  head(3)
```

---

# Step 9 — Handling Missing Values

Missing values (`NA`) require careful thought: is the NA truly *missing*, or is it *meaningfully absent* (e.g., "no license → no driving accident")?

## Detecting NAs

```{r na-detect}
data_clean

is.na(data_clean)               # cell-level logical matrix
colSums(is.na(data_clean))      # count NAs per column
sum(is.na(data_clean))          # total NAs in the whole data frame
mean(is.na(data_clean$accident_condu)) * 100  # % missing
```

## The NA Propagation Rule

```{r na-propagation}
NA + 2
NA / 100
mean(c(1, 2, NA, 4))          # NA — cannot compute
mean(c(1, 2, NA, 4), na.rm = TRUE)  # 2.33 — correct
```

> ⚠️ R gives `NA` **priority** in all calculations. Always use `na.rm = TRUE` for summary functions when NAs are present.

## Replacing NAs Conditionally — `ifelse()` and `case_when()`

```{r na-replace}
# Simple replacement: NA → 0
ifelse(is.na(data_clean$accident_condu), 0, data_clean$accident_condu)

# Conditional: if no permit, risk = 0; otherwise use driving accident rate
ifelse(data_clean$permis == "Non", 0, data_clean$accident_condu)

# Complex case: combine two columns, handling all NA combinations
data_clean %>%
  mutate(total_risk = case_when(
    is.na(accident_condu) & is.na(accident_on_ondu) ~ 0,
    is.na(accident_condu)                           ~ accident_on_ondu,
    is.na(accident_on_ondu)                         ~ accident_condu,
    TRUE ~ accident_condu + accident_on_ondu
  ))
```

> 💡 **`case_when()` is the clean solution** to complex NA logic. It avoids the "code-en-escalier" (staircase code) problem that arises from deeply nested `ifelse()` calls.

## Removing Rows with NAs — `na.omit()` and `drop_na()`

```{r na-remove}
# Base R
na.omit(data_clean)

# dplyr (more flexible)
data_clean %>% drop_na()                           # any NA in any column
data_clean %>% drop_na(accident_condu)             # only if this column is NA
```

---

# Step 10 — Sorting with `arrange()`

```{r arrange}
# Ascending (default)
malaria_big %>%
  arrange(cases) %>%
  head(5)

# Descending
malaria_big %>%
  arrange(desc(cases)) %>%
  head(5)

# Multiple sort keys
malaria_big %>%
  arrange(region, desc(cases)) %>%
  head(8)
```

---

# Step 11 — Grouped Operations

`group_by()` + `summarise()` is the most powerful pattern in dplyr.

## `group_by()` + `summarise()`

```{r group-summarise}
# Mean cases per region
malaria_big %>%
  group_by(region) %>%
  summarise(mean_cases = round(mean(cases), 0),
            total_cases = sum(cases),
            n = n(),
            .groups = "drop")

# Summary by region and month
malaria_big %>%
  group_by(region, month) %>%
  summarise(mean_cases = round(mean(cases), 0),
            .groups = "drop") %>%
  arrange(region, month)
```

## `group_by()` + `mutate()` — Add Group Statistics

```{r group-mutate}
# Add the regional mean as a new column (keeps all rows)
malaria_big %>%
  group_by(region) %>%
  mutate(region_mean = round(mean(cases), 0),
         above_avg   = cases > region_mean) %>%
  select(region, month, cases, region_mean, above_avg) %>%
  head(8)
```

## Base R Equivalent — `aggregate()`

```{r aggregate}
aggregate(cases ~ region, data = malaria_big, FUN = mean)
aggregate(cases ~ region + month, data = malaria_big, FUN = mean)
```

## `tapply()` — Apply a Function by Group

```{r tapply}
tapply(malaria_big$cases, malaria_big$region, mean)
tapply(malaria_big$cases, list(malaria_big$region, malaria_big$month), mean)
```

---

# Step 12 — Counting with `count()` and `tally()`

```{r count}
malaria_big %>% count(region)
malaria_big %>% count(region, sort = TRUE)
malaria_big %>% count(region, month)

# tally() — equivalent when already grouped
malaria_big %>%
  group_by(region) %>%
  tally(sort = TRUE)
```

---

# Step 13 — Reshaping Data (Wide ↔ Long)

Many R functions require **long format** (one row per observation-variable combination). `pivot_longer()` and `pivot_wider()` are the modern tools.

## Wide → Long: `pivot_longer()`

```{r pivot-longer}
# Create a wide dataset: one row per region, one column per month
cases_wide <- malaria_big %>%
  select(region, month, cases) %>%
  pivot_wider(names_from = month, values_from = cases)

cases_wide

# Convert back to long
cases_long <- cases_wide %>%
  pivot_longer(cols      = -region,
               names_to  = "month",
               values_to = "cases")

cases_long
```

## Long → Wide: `pivot_wider()`

```{r pivot-wider}
# Monthly cases in wide format: one region per row, one month per column
malaria_big %>%
  select(region, month, cases) %>%
  pivot_wider(names_from  = month,
              values_from = cases)
```

> 💡 **Long format** is required by `ggplot2`, `dplyr::group_by()`, and most statistical models. **Wide format** is typical of raw exported data from spreadsheets.

---

# Step 14 — Splitting and Combining Columns

## `separate()` — Split One Column into Several

```{r separate}
# Example: split a "region_month" column
df_combined <- data.frame(
  region_month = c("Centre_Jan","Littoral_Feb","Far_North_Mar"),
  cases = c(480, 410, 850)
)

df_combined %>%
  tidyr::separate(col  = region_month,
                  into = c("region", "month"),
                  sep  = "_")
```

## `unite()` — Merge Several Columns into One

```{r unite}
malaria_big %>%
  select(region, month, year) %>%
  tidyr::unite("region_period", region, month, year, sep = "-") %>%
  head(4)
```

---

# Step 15 — Joining Data Frames

When your data is spread across multiple tables, joins combine them by a shared key.

```{r joins}
# Population data for each region
population <- data.frame(
  region     = c("Centre","Littoral","Far_North","North_West","South_West"),
  population = c(4200000, 3800000, 6500000, 2900000, 2200000)
)

# Left join: keep all rows from malaria_big, add population
malaria_joined <- malaria_big %>%
  left_join(population, by = "region")

head(malaria_joined)

# Create a new incidence rate column
malaria_joined %>%
  mutate(incidence_per_1000 = round(cases / population * 1000, 2)) %>%
  arrange(desc(incidence_per_1000)) %>%
  head(5)
```

### Types of Joins

| Function | Keeps |
|---|---|
| `left_join(x, y)` | All rows of x; matching rows of y |
| `right_join(x, y)` | Matching rows of x; all rows of y |
| `inner_join(x, y)` | Only rows matching in both x and y |
| `full_join(x, y)` | All rows from both x and y |
| `anti_join(x, y)` | Rows in x with **no** match in y |

---

# Step 16 — Removing Duplicates

```{r duplicates}
# Create a dataset with duplicate rows
df_dup <- rbind(malaria_big[1:3, ], malaria_big[1:3, ], malaria_big[4:6, ])
nrow(df_dup)

# Remove exact duplicate rows
df_nodup <- df_dup %>% distinct()
nrow(df_nodup)

# Keep distinct by specific columns
df_dup %>% distinct(region, month, .keep_all = TRUE) %>% head(4)
```

---

# Step 17 — Exploratory Data Analysis Packages

The following packages accelerate the EDA phase with automated reports and summaries.

## `skimr::skim()` — Rich Summary

```{r skimr}
# install.packages("skimr")
library(skimr)
skim(malaria_big)
```

Output includes: n_missing, completion_rate, mean, sd, p0 (min), p25, p50, p75, p100 (max), and mini-histograms for numeric columns.

## `DataExplorer::plot_missing()` — Missing Value Overview

```{r dataexplorer, eval=FALSE}
# install.packages("DataExplorer")
library(DataExplorer)

plot_missing(malaria_big)       # bar chart: % missing per column
profile_missing(malaria_big)    # table: count + % missing

# Full HTML report in one command:
create_report(malaria_big)
```

## `dlookr::diagnose()` — Data Quality Diagnostics

```{r dlookr, eval=FALSE}
# install.packages("dlookr")
library(dlookr)

diagnose(malaria_big)            # variable names, types, NAs, unique values
diagnose_web_report(malaria_big) # interactive HTML diagnostic report
eda_web_report(malaria_big)      # full EDA report (distributions + correlations)
plot_na_intersect(malaria_big)   # visualize NA overlap between variables
```

## `GGally::ggpairs()` — Pairwise Relationship Matrix

```{r ggpairs}
library(GGally)

malaria_big %>%
  select(cases, rain) %>%
  ggpairs(title = "Pairwise: Cases vs Rainfall")
```

---

# Step 18 — A Complete Data Manipulation Pipeline

Putting all steps together into one reproducible, readable workflow:

```{r complete-pipeline}
library(tidyverse)
library(janitor)

malaria_report <-
  # 1. Start with raw data
  malaria_big %>%

  # 2. Clean column names
  clean_names() %>%

  # 3. Filter population of interest
  filter(year == 2024) %>%

  # 4. Select relevant columns
  select(region, month, cases, rain) %>%

  # 5. Create new variables
  mutate(
    burden = case_when(
      cases > 1200 ~ "Very High",
      cases > 800  ~ "High",
      cases > 400  ~ "Moderate",
      TRUE         ~ "Low"
    ),
    rain_cat = ifelse(rain > 150, "Rainy", "Dry")
  ) %>%

  # 6. Group and summarise
  group_by(region, burden) %>%
  summarise(
    n          = n(),
    mean_cases = round(mean(cases), 0),
    mean_rain  = round(mean(rain), 0),
    .groups    = "drop"
  ) %>%

  # 7. Sort
  arrange(region, desc(mean_cases))

malaria_report
```

---

# Quick Reference — Key `dplyr` Verbs

| Verb | Purpose | Example |
|---|---|---|
| `select()` | Choose columns | `select(region, cases)` |
| `filter()` | Keep matching rows | `filter(cases > 500)` |
| `mutate()` | Add/modify columns | `mutate(rate = cases/pop)` |
| `rename()` | Rename columns | `rename(n = cases)` |
| `arrange()` | Sort rows | `arrange(desc(cases))` |
| `group_by()` | Group for aggregation | `group_by(region)` |
| `summarise()` | Reduce to summary stats | `summarise(m = mean(cases))` |
| `count()` | Count occurrences | `count(region)` |
| `distinct()` | Remove duplicate rows | `distinct()` |
| `left_join()` | Merge two tables | `left_join(pop, by="region")` |
| `pivot_longer()` | Wide → Long | `pivot_longer(cols=-id)` |
| `pivot_wider()` | Long → Wide | `pivot_wider(names_from=month)` |
| `separate()` | Split one column | `separate(col, into=c(...))` |
| `unite()` | Merge columns | `unite("new", col1, col2)` |
| `drop_na()` | Remove NA rows | `drop_na(cases)` |

---

# Practice Exercises

**Exercise 1.** Using the `data_acc` dataset created at the start of this lecture, clean the column names with `janitor::clean_names()`, convert `sexe` and `permis` to factors, and fix the erroneous age value of 196 (replace with `NA`).

**Exercise 2.** Using `malaria_big`, filter to regions with more than 700 cases in January. Select only `region`, `cases`, and `rain`. Add a column `incidence_flag` that is `"High"` if cases > 1000 and `"Normal"` otherwise. Sort by cases descending.

**Exercise 3.** Using `malaria_big`, compute the mean, maximum, and standard deviation of `cases` for each region. Then add the regional mean back to the original data frame as a new column `region_avg` using `group_by()` + `mutate()` (not `summarise()`).

**Exercise 4.** Create a wide version of `malaria_big` where each row is a region and each column is a month's case count. Then convert it back to long format and verify it matches the original.

**Exercise 5.** Join `malaria_big` with the population data frame (provided in this lecture). Compute a `rate_per_1000` incidence column. Identify the top 3 region-month combinations with the highest incidence rate.

**Exercise 6.** Simulate a dataset with 5% of values missing in the `cases` column of `malaria_big`. Use `colSums(is.na())` to verify. Then fill the missing values with the **regional mean** using `group_by()` + `mutate()` + `case_when()` (or `replace_na()`).

---

```{r cleanup, include=FALSE}
rm(list = ls())
```

---

*End of Lecture — Data Manipulation in R*
