---
title: "Importing and Exporting Data in R"
subtitle: "A Comprehensive Lecture — From Local Files to APIs and Databases"
author: "Ange Maxime TCHOUTANG"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    theme: flatly
    highlight: tango
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  comment = "##",
  message = FALSE,
  warning = FALSE,
  eval    = FALSE   # Most import chunks reference external files; set eval=FALSE globally
)
# Only exploratory chunks that use built-in data will use eval=TRUE
```

---

# Introduction

Every real data analysis begins with the same step: **getting your data into R**. Researchers and analysts work with data stored in many different formats — spreadsheets, comma-separated files, text files, statistical software exports, and online databases. Knowing how to import all of them efficiently is a foundational skill.

In this lecture we cover:

- **The working directory** — where R looks for files
- **Base R import functions** — `read.csv()`, `read.table()`, `read.delim()`
- **Modern tidyverse import** — `readr` package
- **Fast import for large data** — `data.table::fread()`
- **Excel files** — `readxl` package
- **SPSS, Stata, SAS** — `haven` package
- **Online / URL files** — reading directly from the web
- **The `rio` package** — universal import/export
- **Inspecting imported data** — first steps after loading
- **Exporting data** — `write.csv()`, `write_csv()`, `fwrite()`, RData, RDS
- **Saving and reloading R objects** — `.RData` and `.rds`
- **data.frame vs tibble vs data.table** — understanding the three tabular formats
- **A complete real-world workflow**

---

# The Working Directory

The **working directory** is the folder on your computer where R looks for files when you use a filename without a full path. Always set it at the start of any session.

```{r wd}
getwd()                      # Print the current working directory

setwd("path/to/your/data")   # Change to a specific folder
# Example on Windows:
# setwd("C:/Users/Ange/Documents/my_project")
# Example on Mac/Linux:
# setwd("/home/ange/my_project")
```

> 💡 **RStudio shortcut:** In RStudio, go to **Session → Set Working Directory → To Source File Location** to automatically set the working directory to the folder where your current script is saved.

> 💡 **Best practice:** Use **RStudio Projects** (`.Rproj` files). When you open a Project, the working directory is automatically set to the project folder — you never need `setwd()` again.

---

# Understanding File Formats

Before importing, you must know what format your data is in. The most common formats are:

| Format | Extension | Separator | Notes |
|---|---|---|---|
| Comma-Separated Values | `.csv` | `,` | Most universal format |
| CSV (European) | `.csv` | `;` | Used when `,` is the decimal separator |
| Tab-Separated Values | `.tsv` or `.txt` | `\t` | Common in bioinformatics |
| Fixed-width text | `.txt` | spaces | Legacy format |
| Excel | `.xlsx`, `.xls` | — | Microsoft spreadsheet |
| SPSS | `.sav` | — | Statistical software |
| Stata | `.dta` | — | Statistical software |
| SAS | `.sas7bdat` | — | Statistical software |
| R binary | `.RData`, `.rds` | — | Native R formats |

---

# Base R Import Functions

## `read.csv()` — Comma-Separated Values

`read.csv()` is the most frequently used import function. It reads a `.csv` file and returns a `data.frame`.

```{r read-csv}
# Read a local CSV file
data <- read.csv("columbus.csv")

# View as a spreadsheet in RStudio
View(data)

# Display the first few rows in the console
head(data)
```

### Key arguments of `read.csv()`

```{r read-csv-args}
data <- read.csv(
  file      = "myfile.csv",  # path to file
  header    = TRUE,          # first row contains column names (default TRUE)
  sep       = ",",           # column separator (default comma)
  dec       = ".",           # decimal point character (default ".")
  na.strings = "NA",         # what strings to treat as missing values
  stringsAsFactors = FALSE,  # keep character columns as strings (recommended)
  skip      = 0,             # number of lines to skip at the top
  nrows     = -1,            # how many rows to read (-1 = all)
  encoding  = "UTF-8"        # file encoding
)
```

## `read.csv2()` — European CSV (semicolon-separated)

Used in countries where the decimal separator is a comma (`,`), so the field separator is a semicolon (`;`):

```{r read-csv2}
# Semicolon-separated, comma as decimal point
data_eu <- read.csv2("european_data.csv")

# Equivalent using read.table():
data_eu <- read.table("european_data.csv", header = TRUE,
                      sep = ";", dec = ",")
```

## `read.table()` — General Text Files

The most flexible base R import function. All other `read.*` functions are wrappers around it:

```{r read-table}
# Tab-separated file with a header
data_tab <- read.table("data.txt", header = TRUE, sep = "\t")

# Comma-separated (same as read.csv)
data_csv <- read.table("data.csv", header = TRUE, sep = ",")

# No header row
data_noheader <- read.table("data.txt", header = FALSE, sep = "\t")
```

## `read.delim()` — Tab-Separated Files

A convenience wrapper for tab-separated files:

```{r read-delim}
# Tab-separated (default separator is \t)
data_tsv <- read.delim("annotations.tsv")

# Equivalent to:
data_tsv <- read.table("annotations.tsv", header = TRUE, sep = "\t")
```

## `file.choose()` — Interactive File Picker

Opens a dialog box to let you navigate and pick a file manually:

```{r file-choose}
# Opens the file browser — useful for beginners or one-off imports
data <- read.csv(file.choose())
data <- read.delim(file.choose())
```

> ⚠️ `file.choose()` is convenient interactively but **not reproducible** — never use it in scripts intended for others or for automation.

---

# Reading Online / URL Files

You can pass a URL directly to any import function instead of a local path. R will download and read the file automatically:

```{r read-url}
# Read a CSV directly from the web
url_csv <- "https://s3.amazonaws.com/assets.datacamp.com/blog_assets/test.csv"
df <- read.csv(url_csv, header = FALSE)

# Read a text file from the web
url_txt <- "https://s3.amazonaws.com/assets.datacamp.com/blog_assets/test.txt"
df_txt <- read.table(url_txt, header = FALSE)

# Read a tab-separated file from GitHub
url_tsv <- "https://raw.githubusercontent.com/HackBio-Internship/public_datasets/main/metadata_sample.tsv"
bio_data <- read.csv(url_tsv)
```

> 💡 **No download needed.** R streams the file directly from the internet. If you need to work offline later, save the file locally first using `download.file()`.

```{r download-file}
# Download the file first, then read it locally
download.file(url = "https://example.com/data.csv",
              destfile = "local_data.csv")
data <- read.csv("local_data.csv")
```

---

# Modern Import: The `readr` Package

The `readr` package (part of the tidyverse) provides faster, more consistent import functions. Function names use underscores instead of dots (`read_csv` vs `read.csv`) and return a **tibble** instead of a `data.frame`.

```{r readr-install, eval=FALSE}
install.packages("readr")   # or install.packages("tidyverse")
library(readr)
```

## `read_csv()` — Fast CSV Import

```{r read-csv-readr}
library(readr)

data_tb <- read_csv("myfile.csv")

# With options:
data_tb <- read_csv(
  "myfile.csv",
  col_names   = TRUE,            # first row is column names
  na          = c("", "NA", "."), # values treated as missing
  skip        = 0,               # skip header lines
  n_max       = Inf,             # max rows to read
  col_types   = NULL             # auto-detect or specify manually
)
```

### Key differences: `read.csv()` vs `read_csv()`

| Feature | `read.csv()` (base R) | `read_csv()` (readr) |
|---|---|---|
| Output type | `data.frame` | `tibble` |
| Speed | Slower | Faster |
| Printing | Shows all columns | Smart truncation |
| `stringsAsFactors` | Default `FALSE` (R ≥ 4.0) | Never creates factors |
| Column type messages | None | Reports inferred types |
| Whitespace in names | Kept as-is | Preserved, shown with backticks |

## Other `readr` Functions

```{r readr-other}
library(readr)

read_csv2("data.csv")             # Semicolon-separated
read_tsv("data.tsv")              # Tab-separated
read_delim("data.txt", delim="|") # Custom delimiter
read_fwf("data.fwf",              # Fixed-width format
         fwf_widths(c(10, 8, 5)))
```

---

# Fast Import for Large Data: `data.table::fread()`

For datasets with **millions of rows**, `fread()` from the `data.table` package is the fastest option in R — often 5–10× faster than `read.csv()`:

```{r fread}
# install.packages("data.table")
library(data.table)

# Auto-detects separator, header, types — returns a data.table
dt <- fread("large_dataset.csv")

# With options
dt <- fread(
  "large_dataset.csv",
  sep       = ",",
  header    = TRUE,
  nrows     = 10000,     # Read only first 10,000 rows
  select    = c("col1", "col2"),  # Import only specific columns
  na.strings = "NA"
)
```

> 💡 `fread()` prints the 5 first and 5 last rows automatically, making it easy to verify what was loaded.

---

# Reading Excel Files: `readxl`

Base R cannot read `.xlsx` files natively. Use the `readxl` package:

```{r readxl}
# install.packages("readxl")
library(readxl)

# Read the first sheet
data_xl <- read_excel("data.xlsx")

# Specify a sheet by name or number
data_xl <- read_excel("data.xlsx", sheet = "Sheet2")
data_xl <- read_excel("data.xlsx", sheet = 2)

# Read only a specific range
data_xl <- read_excel("data.xlsx", range = "A1:E50")

# Skip rows, set column names
data_xl <- read_excel("data.xlsx", skip = 2, col_names = TRUE)

# List all sheets in a workbook
excel_sheets("data.xlsx")

head(data_xl)
```

---

# Reading SPSS, Stata, and SAS Files: `haven`

The `haven` package reads files from the three major statistical software systems:

```{r haven}
# install.packages("haven")
library(haven)

# SPSS (.sav)
spss_data  <- read_sav("data.sav")

# Stata (.dta)
stata_data <- read_dta("data.dta")

# SAS (.sas7bdat)
sas_data   <- read_sas("data.sas7bdat")

# All three return labelled tibbles that preserve the original variable labels
```

> 💡 `haven` preserves SPSS/Stata **variable labels** and **value labels** as attributes. Use `as_factor()` to convert labelled values to factors.

---

# Universal Import/Export: `rio`

The `rio` package provides a single `import()` function that automatically detects the file format from the extension:

```{r rio}
# install.packages("rio")
library(rio)

# Import — works for CSV, Excel, SPSS, Stata, SAS, JSON, RDS, etc.
data <- import("myfile.csv")
data <- import("myfile.xlsx")
data <- import("myfile.sav")
data <- import("myfile.dta")

# Export — equally universal
export(data, "output.csv")
export(data, "output.xlsx")
export(data, "output.dta")

# Convert between formats in one step
convert("data.sav", "data.csv")
```

`rio` is the simplest option when you need to handle many file formats without memorizing different function names.

---

# The RStudio Import Interface

RStudio provides a point-and-click import tool that generates the corresponding R code for you:

**File → Import Dataset → From Text (readr) / From Excel / From SPSS / From Stata / From SAS**

This opens a dialog where you can preview the data, adjust settings (separator, header, column types), and then copy the auto-generated code into your script.

> 💡 **Always copy the generated code into your script.** Clicking "Import" runs the code interactively but does not save it — you need the code to make your analysis reproducible.

---

# Inspecting Imported Data

After importing, always inspect your data before starting any analysis. This prevents wasted time on analyses built on misread columns or wrong types.

```{r inspect, eval=TRUE}
# Using the built-in airquality dataset as a stand-in
data <- airquality

# Basic dimensions
dim(data)          # Rows × Columns
nrow(data)         # Number of rows (observations)
ncol(data)         # Number of columns (variables)
length(data)       # Same as ncol for data frames

# Column names and types
names(data)        # Variable names
str(data)          # Structure: type + first values of each column

# Preview the data
head(data)         # First 6 rows
head(data, 10)     # First 10 rows
tail(data)         # Last 6 rows

# Statistical summary
summary(data)      # Min, Q1, Median, Mean, Q3, Max per column

# Missing values
is.na(data) |> colSums()      # Count NAs per column
colSums(is.na(data))          # Same result
data[!complete.cases(data), ] # Rows that contain at least one NA

# Column types
sapply(data, class)   # Type of each column
sapply(data, typeof)  # Underlying storage type

# Unique values per column
sapply(data, function(x) length(unique(x)))
```

## Fixing Column Types After Import

Sometimes R guesses column types wrong. Fix them immediately after import:

```{r fix-types}
# Suppose 'age' was read as character
data$age <- as.integer(data$age)

# Suppose 'group' should be a factor
data$group <- as.factor(data$group)

# Suppose a date column was read as character
data$date <- as.Date(data$date, format = "%Y-%m-%d")

# Check the fix
str(data)
```

---

# Exporting Data

## Writing CSV Files

```{r write-csv}
# Base R: creates CSV with row numbers column (row.names = TRUE by default)
write.csv(data, file = "output.csv")

# Cleaner: remove the row names column
write.csv(data, file = "output.csv", row.names = FALSE)

# Semicolon-separated (European format)
write.csv2(data, file = "output_eu.csv", row.names = FALSE)
```

## Writing Text and Tab-Separated Files

```{r write-table}
# Tab-separated text file
write.table(data, file = "output.txt",
            sep       = "\t",
            row.names = FALSE,
            quote     = FALSE)   # Don't wrap strings in quotes
```

## Writing with `readr` (Modern Approach)

```{r write-readr}
library(readr)

write_csv(data, "output.csv")       # No row names by default
write_tsv(data, "output.tsv")       # Tab-separated
write_csv2(data, "output_eu.csv")   # Semicolon-separated
```

## Writing with `data.table::fwrite()` (Fastest)

```{r fwrite}
library(data.table)

fwrite(data, file = "output.csv")             # Default CSV
fwrite(data, file = "output.tsv", sep = "\t") # TSV
```

## Compressing Files

```{r write-compressed}
# Save as compressed bz2 file
write.csv(data, file = bzfile("output.csv.bz2"))

# gzip compression
write.csv(data, file = gzfile("output.csv.gz"))
```

---

# Saving and Loading R Objects

## `.RData` — Save Multiple Objects

`.RData` (or `.Rda`) saves **multiple R objects** (data frames, vectors, models, functions — anything) into a single binary file. When loaded, all saved objects reappear in your environment with their original names.

```{r rdata}
# Save multiple objects
a <- c(5, 10, 15)
b <- data.frame(x = 1:3, y = c("a", "b", "c"))
c <- lm(mpg ~ wt, data = mtcars)

save(a, b, c, file = "my_objects.RData")

# Remove them from environment
rm(a, b, c)

# Reload — a, b, c reappear automatically
load("my_objects.RData")

# Save your entire workspace
save.image(file = "workspace_session.RData")
load("workspace_session.RData")
```

## `.rds` — Save a Single Object

`.rds` saves **one R object**. When you load it, you assign it to any variable name you choose — giving you more control than `.RData`:

```{r rds}
# Save a single data frame
saveRDS(data, file = "clean_data.rds")

# Remove it
rm(data)

# Reload under any name you like
clean_data <- readRDS("clean_data.rds")
my_dataset <- readRDS("clean_data.rds")   # Different name, same content
```

### `.RData` vs `.rds` — When to Use Which

| Feature | `.RData` / `.Rda` | `.rds` |
|---|---|---|
| Number of objects | Multiple | One |
| Load syntax | `load("file.RData")` | `obj <- readRDS("file.rds")` |
| Object names | Fixed (original names) | Freely chosen on load |
| Typical use | Saving a full session | Saving a single clean dataset |

---

# The Three Tabular Formats: data.frame, tibble, data.table

R has three main tabular data structures, each suited for different workflows:

## Creating Each Format

```{r three-formats}
library(tibble)
library(data.table)

# Classic data.frame (base R)
df <- data.frame(
  name = c("Alice", "Bob", "Carla"),
  age  = c(25, 30, 22)
)

# Tibble (tidyverse)
tb <- tibble(
  name = c("Alice", "Bob", "Carla"),
  age  = c(25, 30, 22)
)

# data.table (high performance)
dt <- data.table(
  name = c("Alice", "Bob", "Carla"),
  age  = c(25, 30, 22)
)
```

## Converting Between Formats

```{r convert-formats}
# To data.frame
as.data.frame(tb)    # tibble → data.frame
as.data.frame(dt)    # data.table → data.frame

# To tibble
as_tibble(df)        # data.frame → tibble
as_tibble(dt)        # data.table → tibble

# To data.table
as.data.table(df)    # data.frame → data.table
as.data.table(tb)    # tibble → data.table
```

## Comparison Table

| Feature | `data.frame` | `tibble` | `data.table` |
|---|---|---|---|
| Package | Base R | `tibble` / tidyverse | `data.table` |
| Speed | Moderate | Moderate | Very fast |
| Printing | All columns | Smart truncation, type labels | First + last 5 rows |
| Column types | Infers factors | Never creates factors | Never creates factors |
| Syntax | `df[row, col]` | `df[row, col]` | `dt[i, j, by]` |
| Best for | General use | Tidyverse workflows | Large datasets |

---

# A Complete Real-World Workflow

Here is how all the pieces fit together in a full analysis pipeline:

```{r full-workflow, eval=TRUE}
library(tidyverse)

# --- STEP 1: Set your working directory ---
# setwd("path/to/project")

# --- STEP 2: Import data ---
# From CSV (adjust path as needed)
# data <- read_csv("gene_expression.csv")

# From URL (bioinformatics example — works as-is)
url  <- "https://bioconnector.github.io/workshops/data/brauer2007_messy.csv"
data <- read.csv(url)

# --- STEP 3: First inspection ---
dim(data)
head(data, 5)
str(data)
colSums(is.na(data))

# --- STEP 4: Fix types if needed ---
# data$sample <- as.factor(data$sample)

# --- STEP 5: Quick summary ---
summary(data[, 1:5])

# --- STEP 6: Export cleaned data ---
# write_csv(data, "clean_data.csv")
# saveRDS(data, "clean_data.rds")
```

---

# Advanced: Importing from a Database or API

## Importing via a Loop (Multiple Files)

A common research pattern: import one file per year, bind everything into a single data frame:

```{r loop-import}
library(tidyverse)

# Example: import 10 years of annual weather data
years <- 2014:2023

all_years <- map_dfr(years, function(yr) {
  filename <- paste0("Kirkuk_Weather_", yr, ".csv")
  df       <- read.csv(filename)
  df$year  <- yr    # add a year identifier column
  return(df)
})

dim(all_years)
```

The same pattern using a `for` loop:

```{r loop-for}
results <- list()

for (yr in 2014:2023) {
  filename      <- paste0("Kirkuk_Weather_", yr, ".csv")
  results[[yr]] <- read.csv(filename)
}

all_data <- do.call(rbind, results)
```

## Working with Weather Data APIs (`worldmet`)

```{r worldmet}
# install.packages("worldmet")
library(worldmet)
library(dplyr)
library(lubridate)

# A function that imports NOAA weather data for one year, computes annual averages,
# and writes the result to a CSV file
import_annual_averages <- function(year) {
  # Import from NOAA (station code for Kirkuk, Iraq)
  raw <- importNOAA(code = "406210-99999", year = year)

  # Add year column
  raw$date <- as.Date(raw$date)
  raw$year <- year(raw$date)

  # Compute annual mean for all numeric columns
  annual <- raw %>%
    select_if(is.numeric) %>%
    group_by(year) %>%
    summarise(across(everything(), mean, na.rm = TRUE))

  # Export to CSV
  write.csv(annual,
            paste0("Kirkuk_annual_avg_", year, ".csv"),
            row.names = FALSE)
  return(annual)
}

# Apply to a range of years
for (yr in 2014:2023) {
  import_annual_averages(yr)
}
```

---

# Troubleshooting Common Import Problems

| Problem | Likely cause | Solution |
|---|---|---|
| `cannot open file`: No such file | Wrong path / working directory | Check `getwd()`, use full path, or `file.choose()` |
| Extra column of row numbers | `row.names` present in CSV | `read.csv(file, row.names = 1)` or `write.csv(..., row.names = FALSE)` |
| All columns read as character | Missing `header=TRUE` or wrong separator | Specify `header=TRUE, sep=","` |
| Numeric column contains `NA` | Non-numeric values hidden in data | Check `unique()` and fix source file |
| Date column read as character | No date format specified | `as.Date(col, format="%Y-%m-%d")` |
| `"object not found"` after `load()` | Loaded but name is different | Check `ls()` after `load()` |
| File reads fine but tibble columns look wrong | `col_types` mismatch | Use `spec(data)` to review readr's guess |
| Encoding errors (`<U+FEFF>` etc.) | File encoding is not UTF-8 | Add `fileEncoding = "UTF-8-BOM"` or `"latin1"` |

---

# Quick Reference Card

## Import

```r
# Base R
read.csv("file.csv")                    # CSV
read.csv2("file.csv")                   # Semicolon CSV
read.table("file.txt", sep="\t")        # Tab-separated
read.delim("file.tsv")                  # Tab-separated shortcut
read.csv("https://url/file.csv")        # From URL

# readr (tidyverse — returns tibble)
read_csv("file.csv")
read_csv2("file.csv")
read_tsv("file.tsv")
read_delim("file.txt", delim="|")

# data.table (fastest)
fread("file.csv")

# Excel
readxl::read_excel("file.xlsx", sheet=1)

# SPSS / Stata / SAS
haven::read_sav("file.sav")
haven::read_dta("file.dta")
haven::read_sas("file.sas7bdat")

# Universal
rio::import("file.csv")
rio::import("file.xlsx")
```

## Export

```r
# Base R
write.csv(df, "file.csv", row.names=FALSE)
write.csv2(df, "file.csv", row.names=FALSE)
write.table(df, "file.txt", sep="\t", row.names=FALSE)

# readr
write_csv(df, "file.csv")
write_tsv(df, "file.tsv")

# data.table
fwrite(df, "file.csv")

# R binary
saveRDS(df, "file.rds")          # single object
readRDS("file.rds")              # reload
save(a, b, c, file="objs.RData") # multiple objects
load("objs.RData")               # reload all

# Universal
rio::export(df, "file.xlsx")
```

---

# Practice Exercises

**Exercise 1.** Use `getwd()` to display your current working directory. Create a data frame with 5 rows and 3 columns (name, age, score). Export it as a CSV with `write.csv()` (no row names). Then import it back with `read.csv()` and verify the dimensions match.

**Exercise 2.** Download the following CSV directly from a URL into R:  
`https://raw.githubusercontent.com/datasets/iris/master/data/iris.csv`  
Inspect the result using `str()`, `dim()`, `head()`, and `colSums(is.na(.))`.

**Exercise 3.** Import the same file using `readr::read_csv()`. Compare the output with the `read.csv()` result — what is the class of each result? What differences do you notice in how the data is displayed?

**Exercise 4.** Take any data frame from Exercise 1 or 2 and save it in three different formats: `.csv`, `.rds`, and `.RData`. Then reload each format and verify the content is identical using `identical()`.

**Exercise 5.** Write a function `safe_import(filepath)` that: (1) checks whether the file exists with `file.exists()`; (2) imports it with `read.csv()` if it does; (3) returns a clear error message if it does not. Test it with both a valid and an invalid file path.

**Exercise 6.** Write a loop that generates 3 small data frames (one per year: 2022, 2023, 2024), saves each as a separate CSV (`data_2022.csv`, etc.), then reads all three back and combines them into a single data frame using `do.call(rbind, ...)` or `purrr::map_dfr()`.

---

```{r cleanup, include=FALSE}
rm(list = ls())
```

---

*End of Lecture — Importing and Exporting Data in R*
